---
title: nlmixr2 Neural Network ODEs with pmxNODE
author: Matthew Fidler
date: '2025-04-30'
slug: []
categories: [nlmixr2, pmxNODE]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nlmixr2)
library(rxode2)
library(pmxNODE)

# suppress messages
rxode2:::rxSetSilentErr(1L)
```

## Neural Network ODEs and nlmixr2

I have had some requests to talk about `nlmixr2` using neural network
ODEs, since neural networks are something that more people are
exploring with the explosion of artificial intelligence LLMs.

There is a package, `pmxNODE`, by Dominic Bräm that adds neural network
ODEs to pharmacometric modeling tools like NONMEM, Monolix and
nlmixr2.

In addition to the code that Dominic has added, I extended this
package to allow Neural Networks directly in a `rxode2` or `nlmixr2`
model.  I will go through an annotated example to show how these can
be used directly in the `nlmixr2` model.  Currently the `pmxNODE` does
not have any `nlmixr2` examples in their inst directory, so I will
adapt their NONMEM example to use in `nlmixr2`:


```{r ex1}
library(nlmixr2)
library(pmxNODE)

d <- read.csv(system.file("data_example1_nm.csv", package="pmxNODE"),
              na.strings=".")

ex1 <- function()  {
  ini({
    lV <- 2
    add.sd <- .1
    prop.sd <- .1
  })
  model({
    V <- lV
    d/dt(central) <- NN(c, state=central, min_init=0.5, max_init=5) +
      DOSE * NN(t, state=t, min_init=1, max_init=5, time_nn=TRUE)
    Cc <- central/V
    Cc ~ prop(prop.sd) + add(add.sd)
  })
}
```

This example has 2 neural networks, one related to the `central` state
(labeled `c`) with a minimum activation point of 0.5 and maximum
activation point of 5.  The second is a time-based neural network that
moderates the dose. This has a minimum activation point of 1 and a
maximum activation of 5 (and is called out as a time neural network by
`time_nn=TRUE`, and labeled with `t`).  While these neural networks
may take care of both elimination and absorption independently (with
elimination in the central neural network and dosing in the
time-neural network), they may not be completely independent since
they are neural networks.

In general, the `NN` function is implemented by the `rxode2` language
extension described
[here](https://nlmixr2.github.io/rxode2/articles/Integrating-User-Defined-Functions-into-rxode2.html#functions-to-insert-rxode2-code-into-the-current-model),
but done in the `pmxNODE` package.  This means that the `NN()`
function will only be available if you load the `pmxNODE` package.

The `NN()` function has the form:

- Neural Network identifier, required can be a name or a number;

- `state=` defines the state to be used in the `NN()`. For time, use
  `t`.

- `min_init=` defines the minimal activation point for the `NN()`,
  i.e., minimal expected state.

- `max_init=` defines the maximal activation point for the `NN()`, i.e.,
  maximal expected state.

- `n_hidden=` (optional) defines the number of neurons in the hidden
  layer, default is `5`.

- `act=` (optional) defines activation function in the hidden layer,
  `ReLU` and `Softplus` implemented, default is `ReLU()`.

- `time_nn=` (optional) defines whether the `NN()` should be assumed to be
  a time-dependent `NN()` and consequently all weights from input to
  hidden layer should be strictly negative.

For more information about how to use these functions, I suggest
reading the articles related to `pmxNODE`:

- https://doi.org/10.1002/psp4.13265

- https://doi.org/10.1007/s10928-023-09886-4

Since this extra neural network function is a special user function in
the `pmxNODE`, not only does it mean that you need to load the package
to use it with `nlmixr2`, it also means that the full function is
evaluated when evaluating the UI:

```{r uiEval}
m1 <- ex1()
m2 <- ex1()
# You can see the full code by printing the function:
print(m1)
```

Note that the initial estimates are chosen at random for the neural
network ODE function, you can see that the initial estimates of the
functions `m1` and `m2`:

```{r uiDiff}
m1$theta

m2$theta
```

To make sure your analyses are reproducible with the Neural Network
models, you need to then set the seed.

```{r uiSeedEval}
set.seed(42)
ex1 <- ex1()
```

It may be more helpful to have a population-only model with a neural
network and then add between subject variability to the model.

```{r uiRun}
fit <- suppressMessages(nlmixr(ex1, d, "bobyqa", control=list(print=0)))

p <- plot(fit)

# Here I am subsetting the plots to show only individual plots
p <- p[["All Data"]]

# In this case the list of plots is named starting with "individual"
w <- which(vapply(names(p), function(x) grepl("individual", x), logical(1)))

# This creates a new list of plots, and changes it to the same class
# as output by nlmixr2
p <- lapply(w,function(x) p[[x]])
class(p) <- "nlmixr2PlotList"

p
```

You can see there is no between subject variability on the Neural
Network, you can add it (with the best estimates) with the function
`NNbsv()`.  **Note this function is not yet part of the `pmxNODE`
package, details are below.**

```{r uiRunBsv}
newModel <- fit %>%
  model(V <- lV*exp(eta.V)) %>%
  ini(eta.V ~ .1) %>%
  NNbsv()

fit2 <- suppressMessages(nlmixr2(newModel, d, "focei", control=foceiControl(print=0)))

```

Note the `NNbsv()` is a recent addition to `pmxNODE` and needs to be
reviewed to added to the package. The [pull request](https://github.com/braemd/pmxNODE/pull/7) adds this
function to the package.

You can see the differences here:

```{r uiPlotBsv}
p <- plot(fit2)

# Here I am subsetting the plots to show only individual plots
p <- p[["All Data"]]

# In this case the list of plots is named starting with "individual"
w <- which(vapply(names(p), function(x) grepl("individual", x), logical(1)))

# This creates a new list of plots, and changes it to the same class
# as output by nlmixr2
p <- lapply(w, function(x) p[[x]])
class(p) <- "nlmixr2PlotList"

p
```

## Other activation and NN functions in rxode2/nlmixr2 in the future.

`rxode2` has implemented many neural-network activation functions
built-in.  In the future packages like `pmxNODE` may use these
directly and even extend to use some of the other neural network
functions there.  Since they are built-into rxode2, the models may be
a bit faster if this integration occurs.

## NNbsv() function

Since the pull request hasn't been accepted at this point, I am
providing the code below in case you want to use it yourself:

```{r nnbsv}
NNbsv <- function(ui, val=0.1, str="%s <- l%s*exp(eta.%s)") {
  .ui <- rxode2::assertRxUi(ui)
  .n <- names(.ui$theta)
  .etaNames <- dimnames(.ui$omega)[[1]]
  .nn <- vapply(seq_along(.n), function(i){
    grepl("^[l][Wb].*_[1-9]?[0-9]*", .n[i]) &&
      !any(paste0("eta.", .n[i]) %in% .etaNames)
  }, logical(1))
  .n <- .n[which(.nn)]
  if (length(.n) == 0) return(ui)
  .v <- gsub("^[l]", "", .n)
  .s1 <- paste0(.v, " <- l", .v)
  .s2 <- sprintf(str, .v, .v, .v)
  # Change the model expression first.
  .model <- vapply(.ui$lstChr,
                   function(l) {
                     .w <- which(.s1 == l)
                     if (length(.w) != 1) {
                       return(l)
                     }
                     .s2[.w]
                   }, character(1),
                   USE.NAMES=FALSE)
  rxode2::model(.ui) <- .model
  # Now add eta estimates
  .iniDf <- .ui$iniDf
  .w <- which(!is.na(.iniDf$neta1))
  if (length(.w) == 0L) {
    .maxEta <- 0
  } else {
    .maxEta <- max(.iniDf$neta1[.w])
  }
  .i1 <- .iniDf[1,]
  .i1$ntheta <- NA_integer_
  .i1$lower <- -Inf
  .i1$upper <- Inf
  .i1$est <- val
  .i1$label <- NA_character_
  .i1$backTransform <- NA_character_
  .i1$condition <- "id"
  .i1$err <- NA_character_
  .etas <- do.call(`rbind`,
                   lapply(seq_along(.v), function(i) {
                     .cur <- .i1
                     .cur$neta1 <- .maxEta+i
                     .cur$neta2 <- .maxEta+i
                     .cur$name <- paste0("eta.", .v[i])
                     .cur
                   }))
  .iniDf <- rbind(.iniDf, .etas)

  rxode2::ini(.ui) <- .iniDf
  .ui
}

```
