---
title: monolix2rx
author: Matthew Fidler
date: '2024-10-21'
slug: []
categories: [rxode2, monolix2rx, monolix, babelmixr2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rxode2)
library(nlmixr2)
```
# monolix2rx has been released on CRAN.

I am excited to announce that as part of the `nlmixr2` `3.0` series of
releases, you now have a way to import a `Monolix` model into
`rxode2`/`nlmixr2`.  This is similar to how you can import the output
of `nonmem` into a `rxode2` model with `nonmem2rx`.

## Worked example

In this example, we will:

1. Import a `Monolix` model to `rxode2`

2. Check the validation of that import simply with `plot(rx)`

3. Further convert the `rxode2` model to a `nlmixr2` fit object.


```{r}

library(monolix2rx)
# First load in the model; in this case the theo model
# This is modified from the Monolix demos by saving the model
# file as a text file (hence you can access without model library).
# Additionally some of the file paths were shortened so they could
# be included with monolix2rx

pkgTheo <- system.file("theo", package="monolix2rx")
mlxtranFile <- file.path(pkgTheo, "theophylline_project.mlxtran")

# Note in your own project, you would simply use the file path, you do
# not have to use `system.file()`.  This is only because you are loading
# the monlix model from the `monolix2rx` package.
rx <- monolix2rx(mlxtranFile)

rx

# Look at the validation
plot(rx)

# If you want this to be converted to a nlmixr2 fit you can also
# convert it to a nlmixr2 model using babelmixr2

library(babelmixr2)
fit <- as.nlmixr2(rx)

fit

# This means you can do anything you could in `nlmixr2` with this
# model, even add something like conditional weighted residuals:

fit <- addCwres(fit)
print(fit)
```

## Key notes in importing models to `rxode2` and `nlmixr2`

There are a few things to note in any conversion from `monolix` to
`rxode2` and `nlmixr2`:

- The ODE solvers are different (there is no `lsoda` solver in
  `monolix`). This means there are more expected differences in
  solutions when comparing to importing `ADVAN13` from `NONMEM`.

- `rxode2`/`nlmixr2` event tables and event handling are more closely
  aligned to `NONMEM` event tables and event handling.  The Monolix
  data-set definition is different and uses `ADM`, which means when you
  observe differences between the imported model and the solved model,
  this can be due data conversion or model conversion.  You may want
  to simulate yourself to see if you can reproduce other known items
  in the model to validate the model translation is correct.

- Models in Monolix may not import correctly if they are using the
  Monolix model library without having `Monolix` setup to read them
  ([details
  here](https://nlmixr2.github.io/monolix2rx/#translating-models-from-the-monolix-model-library)).

- When using/sharing/converting Monolix models, you will need to
  include the model ODE model file, the data (for QCing, optional),
  the mlxtran file, and the model fit output directory (reading
  parameter estimates, and output, also optional but can be quite
  helpful).

## Notes about the new matrix specification in `lotri`

As promised in the last release, I am quickly commenting on why we
decided to split the lines to allow the following syntax for matrices:

``` r
a ~ 0.2
b ~ c(0.01, 0.2)
```

As noted in the imported matrix with the covariance of the population
parameters, these matrices can be quite large.  Because of their size,
without splitting them in an easier to read format it is difficult to
tell the covariance between `a` and `b` is `0.01`. When the matrices
get large enough, not only do they use this format but they also call
out what parameter is part of the covariance matrix

In the theopylline example, we have one line of the covariance matrix
listed as:

```r
a ~ c(omega_ka = -0.0001227, omega_V = -6.5914e-05, omega_Cl = -0.00041194,
      a = 0.015333)
```

This covariance matrix now clearly shows that the covariance between
`omega_V` and `a` is very small (`-6.59e-5`). To me, the legibility is
worth the extra method of inputting covariance matrices in `lotri`.


## What this means

With the conversion of a `nlmixr2` fit to `NONMEM` and a `nlmixr2` fit
to `Monolix`, this means that using `nlmixr2` you can convert between
`NONMEM`, `nlmixr2` and `Monolix` making the `nlmixr2` a format and
tool you can use to convert models and streamline processes of
pharmacometric modeling.

This is what we are talking about in our free tutorial at ACoP 2024
"Using Past Models to Bridge to Open Models and Open Science using
nlmixr2": Thursday, November 14, 2024 8:00 AM - 12:00 PM MST Location:
Sonoran Sky Ballroom 1 - 4.  Bring your laptops with `babelmixr2`
setup and you can follow along.

Here we will talk about the conversion process, and some things you
can do with the `nlmixr2` model format including optimal design in
`PopED` (which we have a poster about at ACoP), and individualizing
dosing using `posologyr`.
