---
title: nlmixr2bayes -- Stan, without rewriting your model
author: Matthew Fidler
date: '2026-09-29'
slug: []
categories: [nlmixr2, nlmixr2bayes, stan, bayes]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

Bayesian estimation is the feature we have been asked about the
longest, and the one we have been slowest to give you. Until now the
answer was some version of "write it again in Stan" -- hand-code the
model with [Torsten](https://metrumresearchgroup.github.io/Torsten/),
or move it to NONMEM's `$BAYES`. Either way you end up maintaining two
models: the nlmixr2 one your FOCEi and SAEM runs use, and the Stan one
your posterior comes from. When they disagree, you get to find out
which of the two you typed wrong.

[`nlmixr2bayes`](https://github.com/nlmixr2/nlmixr2bayes) takes the
other road. Instead of translating the model into Stan's language, it
**links Stan to the likelihood you already have**. `rxode2` keeps
solving the ODEs and handling the doses; `nlmixr2est` keeps evaluating
the same conditional likelihood that `focei`, `saem` and `agq` use; and
Stan calls into that as an external function, receiving analytic
gradients rather than autodiffing its way through a solver it wrote
itself.

One model. One likelihood. One source of truth.

```{r install, eval=FALSE}
# install.packages("pak")
pak::pak("nlmixr2/nlmixr2bayes")
```

## Your model, plus priors

Here is the theophylline model you have seen in every one of these
posts. The only thing that is new is the `prior()` lines in `ini({})`:

```{r oneCmt}
library(nlmixr2)
library(nlmixr2bayes)

one.cmt <- function() {
  ini({
    tka <- 0.45
    tcl <- log(c(0, 2.7, 100))
    tv <- 3.45
    add.sd <- c(0, 0.7)
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    prior(tka) ~ dnorm(0.45, 1)
    prior(tcl) ~ dnorm(1, 1)
    prior(tv) ~ dnorm(3.45, 1)
    prior(add.sd) ~ dcauchy(0, 2.5)
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}
```

Priors live next to the parameters they describe, in the same
`lotri`/`rxode2` syntax as everything else in the `ini({})` block. The
distribution catalog is deliberately one-to-one with Stan's -- all 39
real-valued distributions -- so `dnorm()`, `dcauchy()`, `dbeta()`,
`invWishart()` and friends mean exactly what Stan means by them. You
can see the whole list with `lotri::lotriPriorDists()`.

Fitting is one call:

```{r nutsFit}
fit := nlmixr2(one.cmt, theo_sd, est="nuts",
               control=nutsControl(seed=42, chains=2, iter=1000))
```

`est = "nuts"` names the *algorithm* -- Stan's No-U-Turn sampler --
rather than the tool, the same way `"foce"` and `"focei"` name
algorithms in the FOCEi family. (`est = "stan"` with its default
control is the same thing.) The short two-chain run keeps this page
quick to rebuild; the defaults are 4 chains of 2000.

```{r nutsPrint}
print(fit)
```

What comes back is an ordinary `nlmixr2` fit object. The sampler
diagnostics sit in the header, the posterior is summarized in the usual
`$parFixed` table, and IPRED/CWRES/`plot(fit)` all work, because
nothing about the fit object knows or cares that Stan produced the
numbers.

### No priors is an error, not a default

A model without priors is refused, and the error message prints the
exact `prior()` lines you need to paste back into `ini({})`:

```{r noPriors, error=TRUE}
noPriors <- function() {
  ini({
    tka <- 0.45
    tcl <- log(c(0, 2.7, 100))
    tv <- 3.45
    add.sd <- c(0, 0.7)
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}
nlmixr2(noPriors, theo_sd, est="nuts")
```

Those suggested priors come from the model's own `ini()` values and
bounds -- a starting point to edit, not a recommendation.

This is on purpose. Bayesian output built on silently-invented priors
looks every bit as publishable as the real thing, and is wrong. The
one place we do supply a default is the `omega` blocks, where a block
without a `prior()` gets `lkj_corr_cholesky(2)` on the correlation and
a half-Cauchy on each SD -- and even then it is *announced* in a
message rather than assumed quietly.

## What linking buys you

The interesting part is not that this saves typing (though roughly 200
lines of Stan becomes the 20-line model function you already had). It
is that the linked likelihood can do things Stan's own ODE interface
cannot express at all:

- **Dosing events.** Boluses, infusions, `addl`, `ss`, multiple
  compartments, exact event timing -- the rxode2 event table, not a
  hand-rolled reimplementation inside a Stan program.

- **Derivatives with respect to estimated dose handling.** An estimated
  `alag()`, `f()`, `dur()` or `rate()` has a derivative that has to carry
  the jump condition at the dose event. rxode2 already computes it.

- **Delay differential equations.** `d/dt(R) <- kin - kout * delay(R, tau)`
  is an ordinary model here, with `tau` estimated like any other theta.
  Stan has no DDE solver; it never needs one, because it never sees the
  ODE system.

- **The residual-error and censoring machinery.** M2/M3/M4 censoring,
  the full residual catalog, user-written `ll()` endpoints, an estimated
  Box-Cox/Yeo-Johnson lambda -- all of it already validated in
  `nlmixr2est`.

- **`linCmt()`.** Closed-form solutions evaluate 5-10x cheaper than an
  ODE solve, and a native Stan ODE implementation has no way to opt in.

Everything on that list is a consequence of the same decision: Stan is
handed a likelihood, not a differential equation.

## Is it fast?

Reasonably, yes -- and faster than the hand-written Stan version of the
same model, which was not the outcome I expected when I started.

|                                             | native Stan | nlmixr2bayes (linked) |
|---------------------------------------------|-------------|-----------------------|
| wall clock                                  | 142.7 s     | **80.8 s**            |
| worst bulk ESS                              | 236         | **286-378**           |
| worst ESS / s                               | 1.66        | **4.04**              |
| gradient evaluation                         | 1.98 ms     | 1.36 ms (C level)     |
| posterior means                             | agree to < 0.01 on every parameter |

That is **2.4x native Stan's effective samples per second** on a
pure-ODE model, with the same posterior. Three things get it there:
the chains run in forked processes by default on unix (each fork gets a
copy-on-write duplicate of the linked state, and the draws are
bit-identical to a sequential run); the whole tier-2 gradient comes
from *one* solve per subject through a fused sensitivity entry, rather
than two; and `Omega^-1` is rebuilt on the C side instead of by
evaluating an R closure every single evaluation.

Note there is no Stan threading in that story. Layering Stan's
threading runtime on top of rxode2's would just have the two contend
with each other. Parallelism has two axes instead: chains across
forked processes (`chainCores`), or -- with `chainCores = 1` --
subject-parallel OpenMP *inside* each likelihood evaluation (`cores`).
The ODE solve is the rate-limiting step either way, so that is where
the cores go.

You can reproduce the table with `Rscript inst/bench/native-vs-linked.R`
in the package.

## A fast first look: ADVI and Pathfinder

The same generated program and the same linked likelihood also run
Stan's variational approximations and multi-path Pathfinder:

```{r advi}
fitAdvi := nlmixr2(one.cmt, theo_sd, est="advi",
                   control=adviControl(seed=42))
fitPf := nlmixr2(one.cmt, theo_sd, est="pathfinder",
                 control=pathfinderControl(seed=42))
```

Each is sugar for `stanControl(algorithm=)`, and each has its own
control object -- `nutsControl()`, `adviControl()`,
`pathfinderControl()` -- so passing the control *is* the `est=`, the
way `foceControl()` implies `est="foce"`:

```{r sugar, eval=FALSE}
nlmixr2(one.cmt, theo_sd, adviControl(seed=42))  # same fit
```

On the benchmark model above, ADVI finishes in about 30-40 seconds and
Pathfinder in about 90 -- 3-10x faster than the corresponding NUTS run,
and returning the same complete nlmixr2 fit. The population parameters
line up:

```{r compare}
cbind(NUTS=fit$parFixedDf$Estimate,
      ADVI=fitAdvi$parFixedDf$Estimate,
      Pathfinder=fitPf$parFixedDf$Estimate) |>
  `rownames<-`(rownames(fit$parFixedDf)) |>
  round(3)
```

But ADVI is an approximation, and it understates tails and posterior
correlations. Instead of Rhat and ESS, those fits carry the Pareto
$\hat{k}$ diagnostic of the importance ratios:

```{r khat}
fitAdvi$stanDiagnostics$khat
fitPf$stanDiagnostics$khat
```

$\hat{k} \le 0.7$ means the approximation supports importance
correction; above that you get a warning that says so. Hierarchical PK
posteriors routinely exceed it. Treat ADVI and Pathfinder as the fast
first look and NUTS as the answer.

## Reading the fit

Beyond the usual nlmixr2 accessors, a Stan fit carries the posterior
itself. `$parFixed`'s confidence intervals are a Wald approximation
built from the posterior covariance; for anything skewed -- SDs, small
omegas -- use the exact quantiles:

```{r posteriorSummary}
fit$posteriorSummary
```

The `om.<eta>` rows are omega variances on the natural scale and
`cov.<eta1>.<eta2>` rows carry the covariances, so you can find what
you are looking for without counting rows.

`$objDf` grows two Bayesian rows next to the frequentist one:

```{r objDf}
fit$objDf
```

The FOCEi row is a plug-in objective evaluated at the posterior point
estimate -- a bridge to the likelihoodist world for comparability, not
a Bayesian quantity. WAIC and LOO are computed from the per-subject
conditional log-likelihood draws, which makes them
**leave-one-subject-out**: the subject is the exchangeable unit of a
hierarchical model. The full `loo` objects are at `fit$loo` and
`fit$waic`.

And the whole `rstan::stanfit` is right there, so the rest of the
Bayesian ecosystem works unchanged:

```{r stanfit, fig.height=3, fig.width=7}
sf <- fit$stanfit
rstan::traceplot(sf, pars=c("tka", "tcl", "tv"))
```

```{r pairs, fig.height=5, fig.width=6}
bayesplot::mcmc_pairs(sf, pars=c("tka", "tcl", "tv"))
```

One detail I am quietly proud of: the generated program's parameters
read like your model. `tka` is `tka`, `add.sd` is `add_sd`, and an
omega block's internal parameters are suffixed by the etas that are in
it (`sd_eta_ka`), not by an anonymous block index. `fit$stanCode` and
`fit$stanData` hold the exact program and data list that produced the
draws, and `stanControl(run = FALSE)` will hand you the program without
compiling anything if you just want to read it.

```{r diagnostics}
fit$stanDiagnostics
```

Divergences, Rhat, bulk/tail ESS and treedepth saturation are checked
right after sampling and again at print time, each with a message that
names the knob to turn. Do not use estimates from a run with
divergences or Rhat above 1.01.

## When you do want to write Stan

Sometimes the model really is outside what the nlmixr2 ui can say --
an exotic hierarchy, a hand-tuned prior structure, derived quantities
you want computed inside the sampler. The same package ships a second,
independent bridge for that: **rxstan**. You write the `.stan` program
yourself and declare rxode2 as the ODE solver backend:

```{r rxstan, eval=FALSE}
h <- rxsRegister(pkModel, events=ev, sens=c("lka", "lcl", "lv"),
                 output="center", atol=1e-10, rtol=1e-10)
sm <- rxsStanModel("my_model.stan", modelName="my_model")
fit2 <- rstan::sampling(sm, data=list(handle=h, ...), chains=2, iter=1000)
```

Inside the Stan program, `rx_solve(handle, p)` returns the ODE states
*and* rxode2's analytic derivatives of them; Stan differentiates
whatever you build on top of that itself. It is Torsten-shaped
hand-written Stan, with rxode2 underneath instead of Stan's `ode_*`
functions -- so the dosing events, the jump-condition sensitivities and
the DDEs are still there. `rxsCheckGradient()` will confirm the
analytic sensitivities against finite differences, which is the check
to re-run any time you loosen the solver tolerances. See
`vignette("rxstan-handcoded")` and `vignette("rxstan-dde")`.

## Is it right?

This is the part that took the longest, and the part I would want to
see before trusting somebody else's Bayesian package. Because the
bridge operates at the likelihood level, the oracles are
likelihood-level too. The independent ones -- checks against something
the bridge shares no code with:

| check | agreement |
|---|---|
| Stan's assembled gradient vs. Richardson finite differences of its own log density | ~3e-8 |
| `d/d(eta)` and `d/d(theta)` vs. central differences of the linked conditional | ~3e-8 |
| quadrature marginal over the linked conditional vs. `est="agq"` (nAGQ=101) | 3e-7 |
| censored (M2/M3/M4) and `ll()` conditionals vs. hand-computed textbook densities | exact up to a parameter-free constant |
| prior-only sampling (likelihood stripped) vs. every declared prior | one-sample KS per parameter |
| full posterior vs. a hand-written **native-Stan** implementation of the same model | within 3x combined MCSE on every parameter |
| the linked likelihood engine itself | NONMEM-validated upstream in `nlmixr2est` |

Plus exact internal identities (the mu-referenced theta gradient column
must equal the eta gradient; the assembled gradient must be unchanged
by `Omega^-1` swaps spanning four orders of magnitude; 500+ interleaved
evaluations must be bitwise identical, since NUTS's reversibility
assumes the target is a pure function of state), simulate-and-recover
fits including a correlated omega block, and simulation-based
calibration.

And the wrong-by-construction variants are asserted to *fail*: a
sign-flipped gradient assembly fails the finite-difference test, and
`likelihood="focep"` -- whose value and gradient are gradients of
different functions -- is refused outright by capability flags.

## Scope, and what is refused

Supported today: mixed *and* population-only models (a single-subject
Bayesian fit runs through its own path with one external scalar
carrying the whole data log-likelihood), ODE, DDE and `linCmt()`
models, the normal residual family, censored data, user `ll()`
endpoints, mu-referenced and non-mu etas, mu-referenced covariates
(subject-constant and time-varying), estimated dose-handling
parameters, estimated transform-both-sides lambda, IOV
(`iov.x ~ v | OCC`), 2-component finite mixtures with per-subject
membership posteriors, and all 39 real-valued priors in the lotri
catalog including LKJ and inverse-Wishart on omega blocks.

Refused with an explanatory error rather than silently approximated:
mixtures with more than two components (for now), and the eight
discrete distributions, since every `ini({})` parameter is real-valued.

The bridge is also deliberately first-order. `precomputed_gradients()`
builds a reverse-mode `var` and Stan Math has no forward-mode
counterpart for externally supplied partials, so Riemannian HMC and the
embedded-Laplace machinery are out of scope. In practice that costs
nothing: NUTS, ADVI, Pathfinder and optimization are all first-order,
and a Laplace-*marginalized* target is exactly what nlmixr2est's own
FOCEi/Laplace machinery already computes natively.

## Try it

```{r install2, eval=FALSE}
pak::pak("nlmixr2/nlmixr2bayes")
```

If you have an nlmixr2 model you already trust, adding a handful of
`prior()` lines and changing `est="focei"` to `est="nuts"` is the whole
migration. The posterior you get back is sampled against the identical
likelihood code your FOCEi run used -- which means that for the first
time, "what does the Bayesian version say?" is a question about the
data and the priors, and not about whether you translated the model
correctly.

Issues, and the list of what is still open, are at
[github.com/nlmixr2/nlmixr2bayes](https://github.com/nlmixr2/nlmixr2bayes/issues).

## References

Zhang L, Carpenter B, Gelman A, Vehtari A. Pathfinder: Parallel
quasi-Newton variational inference. *J Mach Learn Res*.
2022;23(306):1-49.
