---
title: "State-Dependent Dosing Properties in rxode2"
author: Matthew Fidler
date: '2026-03-31'
slug: []
categories: [rxode2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 5,
  warning  = FALSE,
  message  = FALSE
)
```

## Introduction

A long-standing limitation in `rxode2` was that the dosing-modifier
expressions — `rate()`, `dur()`, `f()`, and `alag()` — could only
reference things that did not depend on state.
If you wanted infusion duration, bioavailability, or lag time to depend
on the *current concentration* or any other ODE state variable, you
were out of luck.

`rxode2` 5.0.2 lifts that restriction.  All four dosing modifiers, as
well as model-event times (`mtime()`), can now reference state
variables.  The expression is evaluated at the moment the dose fires,
using whatever the system state happens to be at that instant.

This opens the door to more realistic pharmacokinetic models where,
for example:

- **Infusion rate or duration** depends on a patient's current drug
  level (e.g., adaptive infusion protocols).
- **Bioavailability** changes with the concentration of a competing
  substrate or an enzyme state.
- **Absorption lag time** varies with gastric emptying driven by
  another compartment.
- **Model event times** are triggered by threshold crossings in a
  state variable.

In this post we walk through each feature with reproducible examples
and compare the simulated trajectories against closed-form analytical
solutions.

```{r load-packages}
library(rxode2)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Analytical Helper

All of the rate and duration examples below involve a one-compartment
depot that drains with first-order rate constant $k_a$ and receives a
zero-order (constant-rate) infusion.
The closed-form solution is:

$$
\text{depot}(t) =
\begin{cases}
\dfrac{R}{k_a}\bigl(1 - e^{-k_a t}\bigr) & 0 \le t \le D \\[6pt]
\dfrac{R}{k_a}\bigl(1 - e^{-k_a D}\bigr)\,e^{-k_a(t-D)} & t > D
\end{cases}
$$

where $R$ is the infusion rate and $D$ is the infusion duration
($D = \text{amt}/R$ or set directly).

```{r analytic-depot}
depot_analytic <- function(t, R, D, ka) {
  vapply(t, function(ti) {
    if (ti <= 0) return(0)
    if (ti <= D) {
      (R / ka) * (1 - exp(-ka * ti))
    } else {
      depot_D <- (R / ka) * (1 - exp(-ka * D))
      depot_D * exp(-ka * (ti - D))
    }
  }, numeric(1))
}
```

---

## State-Dependent Modeled Rate

### The idea

`rate(depot) <- rate0 + state` tells `rxode2` that the zero-order
infusion rate into `depot` is **not** a fixed parameter but an
expression that includes the ODE variable `state`.
The rate is evaluated when the dose fires, and the infusion stop-event
is placed at $t_{\text{dose}} + \text{amt}/R_{\text{eff}}$.

### Baseline: state = 0

When the state is zero the expression collapses to the fixed-rate
case.  This lets us verify that the new code path produces exactly the
same answer as before.

```{r rate-state0}
mod_rate <- rxode2({
  rate(depot) <- rate0 + state
  d/dt(depot) <- -ka * depot
  d/dt(state) <- 0
})

amt   <- 100
rate0 <- 10
ka    <- 0.5
times <- seq(0, 30, by = 0.25)

et_rate <- et() |>
  et(amt = amt, rate = -1, cmt = "depot") |>
  et(times)

s_rate0 <- rxSolve(mod_rate, et_rate,
                    params = c(rate0 = rate0, ka = ka),
                    inits  = c(state = 0, depot = 0))

R0 <- rate0
D0 <- amt / R0
analytic0 <- depot_analytic(times, R0, D0, ka)

df_rate0 <- data.frame(
  time     = times,
  rxode2   = s_rate0$depot[s_rate0$time %in% times],
  analytic = analytic0
)

ggplot(df_rate0, aes(x = time)) +
  geom_line(aes(y = analytic), colour = "grey40", linewidth = 1.2) +
  geom_point(aes(y = rxode2), colour = "#E74C3C", size = 1.4, alpha = 0.7) +
  labs(title = "State-Dependent Rate — state = 0 (baseline)",
       subtitle = paste0("rate(depot) = ", rate0, " + 0  →  R = ", R0,
                         ", D = ", D0, " h"),
       x = "Time (h)", y = "Depot amount") +
  theme_minimal(base_size = 14)
```

The red dots (simulation) sit exactly on the grey curve (analytical).
The new code path reproduces the classical fixed-rate result.

### Non-zero state shifts the rate

Setting `state` to different constant values changes the effective
infusion rate and therefore the duration and peak depot amount.

```{r rate-state-shift}
S0_values <- c(-3, 0, 2, 5)

results_rate <- lapply(S0_values, function(S0) {
  R_eff <- rate0 + S0
  D_eff <- amt / R_eff

  s <- rxSolve(mod_rate, et_rate,
               params = c(rate0 = rate0, ka = ka),
               inits  = c(state = S0, depot = 0))

  data.frame(
    time     = times,
    depot    = s$depot[s$time %in% times],
    analytic = depot_analytic(times, R_eff, D_eff, ka),
    S0       = S0,
    label    = paste0("S0 = ", S0, "  (R = ", R_eff, ", D = ",
                      round(D_eff, 2), " h)")
  )
})

df_rate <- bind_rows(results_rate)

ggplot(df_rate, aes(x = time, colour = label)) +
  geom_line(aes(y = analytic), linewidth = 1.1) +
  geom_point(aes(y = depot), size = 1, alpha = 0.6) +
  labs(title = "State-Dependent Rate — varying initial state",
       subtitle = "rate(depot) <- rate0 + state;   d/dt(state) <- 0",
       x = "Time (h)", y = "Depot amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

Each trajectory matches its own analytical solution perfectly.
Higher `state` -> higher rate -> shorter infusion -> lower peak (the drug
is pushed in faster but drains simultaneously).

### Consecutive doses with decaying state

When `state` decays over time each dose sees a different effective
rate.  Dose 1 gets the high rate (large state); dose 2 gets a
lower rate because state has decayed.

```{r rate-multidose}
mod_rate_decay <- rxode2({
  rate(depot) <- rate0 + state
  d/dt(depot) <- -ka * depot
  d/dt(state) <- -kstate * state
})

et_multi <- et() |>
  et(amt = 50, rate = -1, time = 0) |>
  et(amt = 50, rate = -1, time = 15) |>
  et(seq(0, 30, by = 0.25))

s_decay <- rxSolve(mod_rate_decay, et_multi,
                   params = c(rate0 = 5, ka = 0.4, kstate = 0.2),
                   inits  = c(depot = 0, state = 4))

s_nostate <- rxSolve(mod_rate_decay, et_multi,
                     params = c(rate0 = 5, ka = 0.4, kstate = 0.2),
                     inits  = c(depot = 0, state = 0))

df_multi_rate <- bind_rows(
  data.frame(time = s_decay$time,   depot = s_decay$depot,
             Scenario = "state0 = 4 (decaying)"),
  data.frame(time = s_nostate$time, depot = s_nostate$depot,
             Scenario = "state0 = 0 (no shift)")
)

ggplot(df_multi_rate, aes(time, depot, colour = Scenario)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = c(0, 15), linetype = "dashed", alpha = 0.4) +
  annotate("text", x = 0.5,  y = max(df_multi_rate$depot) * 0.95,
           label = "Dose 1", hjust = 0, size = 3.5) +
  annotate("text", x = 15.5, y = max(df_multi_rate$depot) * 0.95,
           label = "Dose 2", hjust = 0, size = 3.5) +
  labs(title = "Two Doses with Decaying State-Dependent Rate",
       subtitle = "rate(depot) <- rate0 + state;  d/dt(state) <- -kstate * state",
       x = "Time (h)", y = "Depot amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

The orange curve is different because dose 1 fires when $state \approx 4$
(rate = 9) while dose 2 fires when `state` has decayed toward zero
($rate \approx 5$).

### Per-subject initial state changes the infusion profile

Because the rate expression is evaluated per-subject at the dose time,
subjects that start with different compartment values get different
infusion profiles automatically.

```{r rate-per-subject}
mod_rate_subj <- rxode2({
  rate(depot) <- central * 0.1 + 1.0
  d/dt(depot) <- -ka * depot
  d/dt(central) <- ka * depot - kel * central
})

et_subj <- et(amt = 100, rate = -1) |> et(seq(0, 24, by = 0.5))

s_c0 <- rxSolve(mod_rate_subj, et_subj, c(ka = 0.5, kel = 0.2),
                 inits = c(depot = 0, central = 0))
s_c5 <- rxSolve(mod_rate_subj, et_subj, c(ka = 0.5, kel = 0.2),
                 inits = c(depot = 0, central = 5))

df_subj_rate <- bind_rows(
  data.frame(time = s_c0$time, central = s_c0$central,
             Subject = "central₀ = 0  (rate = 1.0)"),
  data.frame(time = s_c5$time, central = s_c5$central,
             Subject = "central₀ = 5  (rate = 1.5)")
)

ggplot(df_subj_rate, aes(time, central, colour = Subject)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Per-Subject State-Dependent Rate",
       subtitle = "rate(depot) <- central * 0.1 + 1.0",
       x = "Time (h)", y = "Central amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

---

## State-Dependent Modeled Duration

### The idea

`dur(depot) <- dur0 + state` sets the infusion **duration** directly
(`rate = -2` in NONMEM convention).
The derived rate is $R = \text{amt} / (\text{dur0} + \text{state})$.

### Baseline and state-shifted duration

```{r dur-demo}
mod_dur <- rxode2({
  dur(depot) <- dur0 + state
  d/dt(depot) <- -ka * depot
  d/dt(state) <- 0
})

amt  <- 100
dur0 <- 10
ka   <- 0.5
times <- seq(0, 40, by = 0.25)

et_dur <- et() |>
  et(amt = amt, rate = -2, cmt = "depot") |>
  et(times)

S0_dur <- c(-3, 0, 2, 5)

results_dur <- lapply(S0_dur, function(S0) {
  D_eff <- dur0 + S0
  R_eff <- amt / D_eff

  s <- rxSolve(mod_dur, et_dur,
               params = c(dur0 = dur0, ka = ka),
               inits  = c(state = S0, depot = 0))

  data.frame(
    time     = times,
    depot    = s$depot[s$time %in% times],
    analytic = depot_analytic(times, R_eff, D_eff, ka),
    S0       = S0,
    label    = sprintf("S0 = %d  (D = %g h, R = %.2f)", S0, D_eff, R_eff)
  )
})

df_dur <- bind_rows(results_dur)

ggplot(df_dur, aes(x = time, colour = label)) +
  geom_line(aes(y = analytic), linewidth = 1.1) +
  geom_point(aes(y = depot), size = 0.9, alpha = 0.5) +
  labs(title = "State-Dependent Duration",
       subtitle = "dur(depot) <- dur0 + state;  D = dur0 + S0;  R = amt / D",
       x = "Time (h)", y = "Depot amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

### Verifying the exact infusion stop time

A critical check: does the infusion actually **stop** at exactly
$D = \text{dur0} + \text{state}$?
We sample densely around the expected stop time and compare the
depot value at $t = D$ to the analytical peak.

```{r dur-stop-time}
dur0_check <- 5
ka_check   <- 0.3
S0_check   <- c(0, 3, -1)

stop_check <- lapply(S0_check, function(S0) {
  D <- dur0_check + S0
  R <- amt / D
  fine_times <- sort(unique(c(seq(0, D + 8, by = 0.05), D)))

  et_fine <- et() |>
    et(amt = amt, rate = -2, cmt = "depot") |>
    et(fine_times)

  s <- rxSolve(mod_dur, et_fine,
               params = c(dur0 = dur0_check, ka = ka_check),
               inits  = c(state = S0, depot = 0))

  obs_D      <- s$depot[which.min(abs(s$time - D))]
  expected_D <- (R / ka_check) * (1 - exp(-ka_check * D))

  data.frame(S0 = S0, D = D, R = round(R, 3),
             depot_sim = round(obs_D, 4),
             depot_analytic = round(expected_D, 4),
             match = abs(obs_D - expected_D) < 1e-3)
})

knitr::kable(bind_rows(stop_check),
             caption = "Depot at infusion stop time D: simulation vs analytical")
```

### Multiple doses with decaying state

```{r dur-multidose}
mod_dur_decay <- rxode2({
  dur(depot) <- dur0 + state
  d/dt(depot) <- -ka * depot
  d/dt(state) <- -kstate * state
})

et_dur_multi <- et() |>
  et(amt = 50, rate = -2, time = 0) |>
  et(amt = 50, rate = -2, time = 20) |>
  et(seq(0, 35, by = 0.25))

s_dur_decay <- rxSolve(mod_dur_decay, et_dur_multi,
                       params = c(dur0 = 5, ka = 0.4, kstate = 0.2),
                       inits  = c(depot = 0, state = 3))

s_dur_flat  <- rxSolve(mod_dur_decay, et_dur_multi,
                       params = c(dur0 = 5, ka = 0.4, kstate = 0.2),
                       inits  = c(depot = 0, state = 0))

df_dur_multi <- bind_rows(
  data.frame(time = s_dur_decay$time, depot = s_dur_decay$depot,
             Scenario = "state0 = 3 (decaying)"),
  data.frame(time = s_dur_flat$time,  depot = s_dur_flat$depot,
             Scenario = "state0 = 0 (fixed dur)")
)

ggplot(df_dur_multi, aes(time, depot, colour = Scenario)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = c(0, 20), linetype = "dashed", alpha = 0.4) +
  labs(title = "Multiple Doses with Decaying State-Dependent Duration",
       subtitle = "dur(depot) <- dur0 + state;  d/dt(state) <- -kstate * state",
       x = "Time (h)", y = "Depot amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

---

## State-Dependent Bioavailability

### The idea

`f(depot) <- f0 + state` means the bioavailable fraction is evaluated
at the moment each dose fires.
For a bolus dose into a mono-exponential depot:

$$
\text{depot}(t) = \text{amt} \cdot (f_0 + S_0)\; e^{-k_a\,t}
$$

### Single dose — varying state

```{r f-single}
mod_f <- rxode2({
  f(depot)    <- f0 + state
  d/dt(depot) <- -ka * depot
  d/dt(state) <- 0
})

amt_f <- 100
f0    <- 0.8
ka_f  <- 0.5
times_f <- seq(0, 20, by = 0.25)

et_f <- et() |>
  et(amt = amt_f, cmt = "depot") |>
  et(times_f)

S0_f <- c(-0.2, 0, 0.1, 0.5)

results_f <- lapply(S0_f, function(S0) {
  F_eff <- f0 + S0
  s <- rxSolve(mod_f, et_f,
               params = c(f0 = f0, ka = ka_f),
               inits  = c(state = S0, depot = 0))

  data.frame(
    time     = times_f,
    depot    = s$depot[s$time %in% times_f],
    analytic = amt_f * F_eff * exp(-ka_f * times_f),
    S0       = S0,
    label    = sprintf("S0 = %.1f  (F = %.2f)", S0, F_eff)
  )
})

df_f <- bind_rows(results_f)

ggplot(df_f, aes(x = time, colour = label)) +
  geom_line(aes(y = analytic), linewidth = 1.1) +
  geom_point(aes(y = depot), size = 0.9, alpha = 0.5) +
  labs(title = "State-Dependent Bioavailability — Single Dose",
       subtitle = "f(depot) <- f0 + state;  depot(t) = amt × (f0 + S0) × exp(−ka t)",
       x = "Time (h)", y = "Depot amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

### Two doses with decaying state

With a decaying state, the second dose sees a lower bioavailability
than the first.

```{r f-two-dose}
mod_f_decay <- rxode2({
  f(depot)    <- f0 + state
  d/dt(depot) <- -ka * depot
  d/dt(state) <- -kstate * state
})

f0_2     <- 0.8
ka_2     <- 0.5
kstate_2 <- 0.3
amt_2    <- 100
dose_times <- c(0, 5)
times_f2   <- seq(0, 15, by = 0.25)

et_f2 <- et() |>
  et(amt = amt_2, time = 0, cmt = "depot") |>
  et(amt = amt_2, time = 5, cmt = "depot") |>
  et(times_f2)

S0_f2 <- c(0, 0.2, 0.5)

results_f2 <- lapply(S0_f2, function(S0) {
  s <- rxSolve(mod_f_decay, et_f2,
               params = c(f0 = f0_2, ka = ka_2, kstate = kstate_2),
               inits  = c(depot = 0, state = S0))

  # Analytical:  state(t) = S0 * exp(-kstate * t)
  # f at each dose time = f0 + S0 * exp(-kstate * t_d)
  expected <- vapply(times_f2, function(t) {
    contrib <- 0
    for (td in dose_times) {
      if (t >= td) {
        f_td   <- f0_2 + S0 * exp(-kstate_2 * td)
        contrib <- contrib + amt_2 * f_td * exp(-ka_2 * (t - td))
      }
    }
    contrib
  }, numeric(1))

  data.frame(
    time     = times_f2,
    depot    = s$depot[s$time %in% times_f2],
    analytic = expected,
    S0       = S0,
    label    = sprintf("S0 = %.1f", S0)
  )
})

df_f2 <- bind_rows(results_f2)

ggplot(df_f2, aes(x = time, colour = label)) +
  geom_line(aes(y = analytic), linewidth = 1.1) +
  geom_point(aes(y = depot), size = 0.9, alpha = 0.5) +
  geom_vline(xintercept = dose_times, linetype = "dashed", alpha = 0.3) +
  labs(title = "Two Doses with Decaying State-Dependent Bioavailability",
       subtitle = "f(depot) <- f0 + state;  d/dt(state) <- -kstate × state",
       x = "Time (h)", y = "Depot amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

Note how the second dose (dashed line at $t = 5$) adds less drug to
the depot when $S_0 > 0$ because `state` has partially decayed,
reducing $f$.

---

## State-Dependent Lag Time (alag)

### The idea

`alag(depot) <- state_val` delays each dose by the current value of
`state_val` at the scheduled dose time.
A bolus intended for $t = 0$ actually fires at $t = \text{state\_val}$.

### Constant lag via a state variable

We use `d/dt(state_val) <- 0` so the lag is a known constant
and the trajectory is analytically verifiable.

```{r alag-constant}
mod_alag <- rxode2({
  alag(depot) <- state_val
  d/dt(depot)     <- -ka * depot
  d/dt(central)   <- ka * depot - kel * central
  d/dt(state_val) <- 0
})

params_alag <- c(ka = 0.5, kel = 0.2)

# Two doses at t = 0 and t = 3
et_alag <- et(amt = 100, time = 0, cmt = "depot") |>
  et(amt = 100, time = 3, cmt = "depot") |>
  et(seq(0, 12, by = 0.25))

lag_values <- c(0, 1, 2)

results_alag <- lapply(lag_values, function(sv) {
  s <- rxSolve(mod_alag, et_alag, params_alag,
               inits = c(depot = 0, central = 0, state_val = sv))

  t_fire1 <- sv
  t_fire2 <- 3 + sv

  analytic <- vapply(s$time, function(t) {
    d <- 0
    if (t >= t_fire1) d <- d + 100 * exp(-0.5 * (t - t_fire1))
    if (t >= t_fire2) d <- d + 100 * exp(-0.5 * (t - t_fire2))
    d
  }, numeric(1))

  data.frame(time = s$time, depot = s$depot,
             analytic = analytic, lag = sv,
             label = paste0("lag = ", sv, " h"))
})

df_alag <- bind_rows(results_alag)

ggplot(df_alag, aes(x = time, colour = label)) +
  geom_line(aes(y = analytic), linewidth = 1.1) +
  geom_point(aes(y = depot), size = 0.9, alpha = 0.5) +
  labs(title = "State-Dependent Lag Time — Constant State",
       subtitle = "alag(depot) <- state_val;  d/dt(state_val) <- 0",
       x = "Time (h)", y = "Depot amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

Each dose is shifted rightward by exactly the `state_val` amount.

### Lag depending on a dynamic state

When the lag depends on a compartment that changes over time, each
dose gets a different effective lag.

```{r alag-dynamic}
mod_alag_dyn <- rxode2({
  alag(depot) <- central * 0.1
  d/dt(depot)   <- -ka * depot
  d/dt(central) <- ka * depot - kel * central
})

et_alag_dyn <- et() |>
  et(amt = 100, time = 0) |>
  et(amt = 100, time = 6) |>
  et(seq(0, 24, by = 0.25))

s_lag0 <- rxSolve(mod_alag_dyn, et_alag_dyn,
                  c(ka = 0.5, kel = 0.2),
                  inits = c(depot = 0, central = 0))

s_lag5 <- rxSolve(mod_alag_dyn, et_alag_dyn,
                  c(ka = 0.5, kel = 0.2),
                  inits = c(depot = 0, central = 5))

df_alag_dyn <- bind_rows(
  data.frame(time = s_lag0$time, central = s_lag0$central,
             Subject = "central0 = 0  (lag = 0 h)"),
  data.frame(time = s_lag5$time, central = s_lag5$central,
             Subject = "central0 = 5  (lag = 0.5 h)")
)

ggplot(df_alag_dyn, aes(time, central, colour = Subject)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Dynamic State-Dependent Lag Time",
       subtitle = "alag(depot) <- central × 0.1",
       x = "Time (h)", y = "Central amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

The subject starting with `central = 5` has a 0.5 h lag on the first
dose, visibly delaying the rise in central concentrations.

---

## Combining Features: Lag + Duration

Dosing modifiers compose.
Here we combine a fixed lag with a state-dependent duration to show
that the infusion-stop event is placed at the correct absolute time:
$t_{\text{stop}} = t_{\text{dose}} + \text{lag} + \text{dur}$.

```{r alag-plus-dur}
mod_combo <- rxode2({
  alag(depot) <- 0.5
  dur(depot)  <- central * 0.1 + 1.0
  d/dt(depot)   <- -ka * depot
  d/dt(central) <- ka * depot - kel * central
})

et_combo <- et(amt = 100, rate = -2) |> et(seq(0, 24, by = 0.25))

s_combo <- rxSolve(mod_combo, et_combo,
                   c(ka = 0.5, kel = 0.2),
                   inits = c(depot = 0, central = 3))

s_combo0 <- rxSolve(mod_combo, et_combo,
                    c(ka = 0.5, kel = 0.2),
                    inits = c(depot = 0, central = 0))

df_combo <- bind_rows(
  data.frame(time = s_combo$time, depot = s_combo$depot,
             central = s_combo$central,
             Scenario = "central₀ = 3  (lag 0.5 h, dur ≈ 1.3 h)"),
  data.frame(time = s_combo0$time, depot = s_combo0$depot,
             central = s_combo0$central,
             Scenario = "central₀ = 0  (lag 0.5 h, dur = 1.0 h)")
)

p_combo_depot <- ggplot(df_combo, aes(time, depot, colour = Scenario)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Combined Lag + State-Dependent Duration — Depot",
       subtitle = "alag(depot) <- 0.5;  dur(depot) <- central × 0.1 + 1.0",
       x = "Time (h)", y = "Depot amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

p_combo_central <- ggplot(df_combo, aes(time, central, colour = Scenario)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Combined Lag + State-Dependent Duration — Central",
       x = "Time (h)", y = "Central amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

p_combo_depot
p_combo_central
```

The subject with `central₀ = 3` has a slightly longer infusion
duration (\$0.1 \times 3 + 1 = 1.3$ h) compared to the baseline
(\$1.0$ h).
Both subjects share the same 0.5 h absorption lag.
The plots confirm that the stop-event is placed correctly:
the depot rises only after the lag, and the infusion ends at
$$t = 0 + 0.5 + \text{dur}$$.

---

## State-Dependent Model Event Times (mtime)

### The idea

`mtime()` inserts an extra observation/solver-restart time into the
integration grid.
Previously the argument had to be a constant or parameter expression.
Now it can reference state variables, enabling event times that depend
on the evolving system.

A practical use case is triggering a covariate shift (e.g., enzyme
induction onset) at a time that depends on cumulative drug exposure.

```{r mtime-demo}

mod_mtime <- rxode2({
  # Model event time depends on a state variable
  # state decays; mtime fires at t = 2 + state (evaluated at init)
  mtime(mt) <- 2 + state

  # Effect switches on at the model event time
  eff <- ifelse(time >= mt, 1, 0)

  d/dt(depot)   <- -ka * depot
  d/dt(central) <- ka * depot - kel * central * (1 + 0.5 * eff)
  d/dt(state)   <- -kstate * state
})


et_mt <- et(amt = 100) |> et(seq(0, 24, by = 0.1))

s_mt0 <- rxSolve(mod_mtime, et_mt,
                 c(ka = 1.0, kel = 0.2, kstate = 0.1),
                 inits = c(depot = 0, central = 0, state = 0))

s_mt3 <- rxSolve(mod_mtime, et_mt,
                 c(ka = 1.0, kel = 0.2, kstate = 0.1),
                 inits = c(depot = 0, central = 0, state = 3))

s_mt6 <- rxSolve(mod_mtime, et_mt,
                 c(ka = 1.0, kel = 0.2, kstate = 0.1),
                 inits = c(depot = 0, central = 0, state = 6))

df_mtime <- bind_rows(
  data.frame(time = s_mt0$time, central = s_mt0$central,
             Scenario = "state0 = 0  (mtime = 2 h)"),
  data.frame(time = s_mt3$time, central = s_mt3$central,
             Scenario = "state0 = 3  (mtime = 5 h)"),
  data.frame(time = s_mt6$time, central = s_mt6$central,
             Scenario = "state0 = 6  (mtime = 8 h)")
)

ggplot(df_mtime, aes(time, central, colour = Scenario)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = c(2, 5, 8), linetype = "dashed", alpha = 0.3) +
  annotate("text", x = c(2.3, 5.3, 8.3),
           y = rep(max(df_mtime$central) * 0.95, 3),
           label = c("mtime=2", "mtime=5", "mtime=8"),
           size = 3, hjust = 0) +
  labs(title = "State-Dependent Model Event Times",
       subtitle = "mtime(mt) <- 2 + state;  eff switches on at mt",
       x = "Time (h)", y = "Central amount", colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

The dashed vertical lines mark the model event time for each scenario.
Higher initial `state` pushes the effect onset later, resulting in
higher peak concentrations (the enhanced elimination kicks in later).

---

## Summary of New Capabilities

The table below summarizes what is now possible with state-dependent
dosing properties:

```{r summary-table, echo=FALSE}
summary_df <- data.frame(
  Feature = c("rate(cmt)",
              "dur(cmt)",
              "f(cmt)",
              "alag(cmt)",
              "mtime()"),
  `NONMEM Analog` = c("RATE = -1 with state-dep expression",
                       "RATE = -2 with state-dep expression",
                       "F1 depending on state",
                       "ALAG1 depending on state",
                       "MTIME depending on state"),
  `When Evaluated` = c("At dose firing time",
                        "At dose firing time",
                        "At dose firing time",
                        "At dose sorting / firing time",
                        "At model initialization"),
  `Key Behaviour` = c("Infusion stop re-sorted; R = expr at dose time",
                       "Duration = expr at dose time; R = amt/dur",
                       "Effective amount = amt × f(state)",
                       "Dose shifted to t_dose + alag(state)",
                       "Solver restart point depends on state"),
  check.names = FALSE
)

knitr::kable(summary_df, align = "lllll",
             caption = "State-dependent dosing modifiers added in PR #991")
```

### Key implementation details

1. **Evaluation timing.**
   All dosing modifiers are evaluated at the moment the dose event
   fires (or, for `alag`, at the sort phase before firing).
   The solver uses the state vector at that instant.

2. **Runtime re-sort.**
   When a state-dependent `rate()` or `dur()` changes the infusion
   stop time, `rxode2` re-sorts the event queue at runtime so that
   the stop event lands at the correct absolute time.

3. **Per-subject evaluation.**
   In population simulations each subject's initial conditions
   (and evolving states) are used independently, so the same model
   naturally produces different dosing profiles across subjects.

4. **Multiple doses.**
   Each dose is evaluated independently.  With a decaying or
   accumulating state, consecutive doses get different effective rates,
   durations, bioavailabilities, or lags.

5. **Composability.**
   The modifiers compose freely — you can have a state-dependent lag
   *and* a state-dependent duration on the same compartment.

---

## A More Realistic Example: Adaptive Infusion

To close, here is a slightly more pharmacologically motivated example.
Suppose a drug is given by IV infusion, and the clinical protocol
adjusts the infusion rate based on the patient's current plasma
concentration (a form of therapeutic drug monitoring built into the
model):

```{r adaptive-infusion}
mod_adaptive <- rxode2({
  # Infusion rate is adjusted based on current central concentration:
  # higher concentration → slower infusion (safety cap)
  rate(central) <- Rmax / (1 + central / IC50)

  d/dt(central) <- -(CL / V) * central
  cp <- central / V
})

et_adaptive <- et() |>
  et(amt = 500, rate = -1, time = 0) |>
  et(amt = 500, rate = -1, time = 12) |>
  et(seq(0, 36, by = 0.1))

params_adpt <- c(Rmax = 100, IC50 = 5, CL = 5, V = 50)

# Subject starting from scratch
s_naive <- rxSolve(mod_adaptive, et_adaptive, params_adpt,
                   inits = c(central = 0))

# Subject already at steady state (high concentration)
s_loaded <- rxSolve(mod_adaptive, et_adaptive, params_adpt,
                    inits = c(central = 250))

df_adaptive <- bind_rows(
  data.frame(time = s_naive$time,  cp = s_naive$cp,
             Subject = "Drug-naïve (central₀ = 0)"),
  data.frame(time = s_loaded$time, cp = s_loaded$cp,
             Subject = "Pre-loaded (central₀ = 250)")
)

ggplot(df_adaptive, aes(time, cp, colour = Subject)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = c(0, 12), linetype = "dashed", alpha = 0.3) +
  labs(title = "Adaptive Infusion Rate Based on Plasma Concentration",
       subtitle = "rate(central) <- Rmax / (1 + central / IC50)",
       x = "Time (h)",
       y = "Plasma concentration (mg/L)",
       colour = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

The pre-loaded subject receives a much slower infusion (the
denominator \$1 + C/\text{IC50}$ is large), producing a markedly
different PK profile from the same dosing regimen — all handled
automatically by `rxode2` evaluating the rate expression at dose time.

---

## Conclusion

rxode2 5.0.2 adds a powerful feature to `rxode2`: the ability to make
dosing properties — infusion rate, infusion duration, bioavailability,
absorption lag time, and model event times — depend on the current
state of the ODE system.

This required changes deep in the event-sorting and dose-firing
machinery, including runtime re-sorting of infusion stop events and
careful evaluation of state vectors at the correct time points.
The result is a clean, intuitive syntax (`rate(cmt) <- expr`,
`dur(cmt) <- expr`, `f(cmt) <- expr`, `alag(cmt) <- expr`) that
works exactly the same whether the expression involves only parameters
or also state variables.

To try it out, install the latest version of `rxode2` from
CRAN.

As always, please report any issues on
[GitHub](https://github.com/nlmixr2/rxode2/issues).

## Session Information

```{r session-info}
sessionInfo()
```