---
title: "Adaptive dosing in rxode2"
subtitle: "Some patients don't follow a script — neither should your event table"
author: Matthew Fidler
date: '2026-09-09'
slug: []
categories: [rxode2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7.8, fig.height = 4.6,
                      dpi = 110, out.width = "100%")
library(rxode2)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
```

In the past, every `rxode2` simulation started the same way: you coded
down the doses first, and then you solved.  The event table is fixed
before the solver takes its first step.

That is fine for a fixed schedule.  It is not fine clinical adaptive
dosing.  Real protocols have clinical endpoints decide: hold
this cycle, cut the dose one level, start the rescue infusion, draw an
extra sample.

This means that dose that a patient receives on day 42 is a *function
of the trajectory* between day 0 and day 42 -- which means you cannot
write it into an event table, because you do not know it until you
have solved.

`rxode2` can now modify an individual's event history *while the model
is being solved*.  The new [adaptive dosing
vignette](https://nlmixr2.github.io/rxode2/articles/adaptive-dosing.html)
covers the whole feature; this post
uses the famous Friburg myelosuppression model and a real dose-modification
table from a product label.  Everything below needs `rxode2` >= 5.1.7.

## The idea in three lines

Decision logic goes inside the `model({})` block, it runs at the times
the solver visits, and when a condition is met you push a new event:

```{r taste}
titrate <- function() {
  ini({
    ka <- 0.5
    cl <- 1
    v <- 10
  })
  model({
    d/dt(depot)   <- -ka*depot
    d/dt(central) <-  ka*depot - cl/v*central
    cp <- central/v
    # from day 1 on, top up if the daily check finds us below target

    # This is the code to say every 24 hours check to see if cp < 1,
    # then dose if needed
    if (t > 0 && t %% 24 == 0 && cp < 1) {
      bolus(50)
    }
  })
}

titrateSolve <- titrate |>
  rxSolve(et(amt = 100, time = 0) |> et(seq(0, 96, by = 1)))

ggplot(titrateSolve, aes(time, cp)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  geom_line(linewidth = 0.7, colour = "#2166ac") +
  theme_bw() +
  labs(x = "time (h)", y = "concentration",
       title = "Top up whenever the daily check finds us below target")
```

The helpers all push a standard NONMEM-style event, so there is
nothing exotic downstream:

| helper | `evid` | what it pushes |
|---|---|---|
| `bolus()` | 1 | a bolus dose, optionally with `ii`/`addl`/`ss` |
| `infuse()` | 1 | a fixed-**rate** infusion |
| `infuseDur()` | 1 | a fixed-**duration** infusion |
| `reset()` | 3 | a system reset |
| `replace()` | 5 | set a compartment to a value |
| `multiply()` | 6 | scale a compartment |
| `phantom()` | 7 | transit dosing bookkeeping (`tad()`, `podo()`) without mass |
| `obs()` | 0 | extra observation rows |
| `evid_()` | any | the low-level interface |

Note that the condition is only tested at times the solver actually stops
at; either the decision times have to be in the event table, or pinned
with an `mtime()` assignment:

```r
mtime(check48) <- 48        # note the assignment -- mtime(48) will not parse
if (t == check48 && cp < 1) {
  bolus(50)
}
```

A few things to keep in mind:

`cp < 1` stays true almost always (except the first dose) so:

 - alone it will fire almost every observation

 - For it to be useful, it needs to be anchored to a visit
   (`t == check48`) or to a visit schedule, like every 24 hours (`t %% 24 == 0`).

This can be easy to get wrong.  To prevent dosing at more times than
you expect, you can use `maxExtra` to only add a specific number of
doses per problem.

As a note, the `t > 0` above matters too: at time zero `cp` is still
`0`, so without it the model will add a dose at time zero, even though
the baseline dose is in the event table.

One last note, at the very end of the dosing if a new dose would be
added, it won't be shown in the observations since the model doesn't
check there or has no observations after the dose.

## A model that needs this

One of the standard places where you need adaptive dosing is cancer
treatment.  For example, chemotherapy-induced neutropenia.

[Friberg et al.
(2002)](https://ascopubs.org/doi/10.1200/JCO.2002.02.140) described
myelosuppression model under linear drug effect, three maturation
transit compartments, and a rebound feedback term
$(\mathrm{Circ}_0/\mathrm{Circ})^\gamma$.  We took the estimates of
docetaxel from [Ozawa, Minami and Sato
(2007)](https://pubmed.ncbi.nlm.nih.gov/17888031/), who fit a
three-compartment PK model together with the Friberg neutropenia model
in Japanese cancer patients:

```{r model}
doce <- function() {
  ini({
    cl <- 35.7 # L/h
    v1 <- 6.94 # L
    q2 <- 5.58 # L/h
    v2 <- 7.39 # L
    q3 <- 12.5 # L/h
    v3 <- 225 # L
    circ0 <- 5.05 # baseline neurtophil count *10^9/L
    mtt <- 113 # mean transit time (hr)
    gam <- 0.196 # shape
    slope <- 17.9 # mL / ug
  })
  model({
    ktr <- 4/mtt
    cp  <- central/v1

    d/dt(central) <- -(cl/v1)*central - (q2/v1)*central + (q2/v2)*periph1 -
                      (q3/v1)*central + (q3/v3)*periph2
    d/dt(periph1) <-  (q2/v1)*central - (q2/v2)*periph1
    d/dt(periph2) <-  (q3/v1)*central - (q3/v3)*periph2

    circS <- max(circ, 1e-4)
    d/dt(prol) <- ktr*prol*(1 - slope*cp)*(circ0/circS)^gam - ktr*prol
    d/dt(tr1)  <- ktr*(prol - tr1)
    d/dt(tr2)  <- ktr*(tr1  - tr2)
    d/dt(tr3)  <- ktr*(tr2  - tr3)
    d/dt(circ) <- ktr*(tr3  - circ)

    prol(0) <- circ0
    tr1(0) <- circ0
    tr2(0) <- circ0
    tr3(0) <- circ0
    circ(0) <- circ0

    anc <- circ # Absolute Neutrophil Count (ANC)
  })
}
```

Six cycles of docetaxel 100 mg/m^2 as a one-hour infusion every three
weeks, in a 1.8 m^2 typical patient, is an ordinary event table:

```{r typical}
bsa <- 1.8

fixedEv <- et(amt = 100*bsa, dur = 1, cmt = "central", ii = 21*24, addl = 5) |>
  et(seq(0, 126*24, by = 6))

typical <- rxSolve(doce, fixedEv)

ggplot(typical, aes(time/24, anc)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = c(0.5, 1.5), linetype = 2, colour = "grey40") +
  theme_bw() +
  labs(x = "day", y = "ANC (10^9/L)",
       title = "Docetaxel 100 mg/m2 q3w, typical patient")
```

The nadir is `r round(min(typical$anc), 3)` * 10^9/L on day
`r round(typical$time[which.min(typical$anc)]/24, 1)`, sitting just
above the grade-4 line, and the count is back over 1.5 *10^9/L well
before the next cycle: deep grade 3, no dose held, no dose cut.

## The protocol only matters for the patients in the tail

As with all mixed effects models, for some patients it does matter.
If we simulate 50 subjects on the same fixed schedule:

```{r pop-fixed}
set.seed(20260909)
n <- 50
pop <- data.frame(
  id    = 1:n,
  cl    = 35.7*exp(rnorm(n, 0, 0.30)),
  slope = 17.9*exp(rnorm(n, 0, 0.35)),
  circ0 = 5.05*exp(rnorm(n, 0, 0.25)),
  mtt   =  113*exp(rnorm(n, 0, 0.20)))

popEv <- fixedEv |> et(id = 1:n)

popFixed <- rxSolve(doce, popEv, params = pop, returnType="data.frame")

doseDays <- (0:5)*21*24

popFixed |>
  filter(time %in% doseDays) |>
  summarise(`doses at ANC < 1.5` = sum(anc < 1.5), n = n())

popFixed |>
  mutate(cycle = pmin(6, floor(time/504) + 1)) |>
  group_by(id, cycle) |>
  summarise(days = sum(anc < 0.5)*6/24, .groups = "drop") |>
  summarise(`cycles with >7 d grade 4` = sum(days > 7),
            `subjects affected` = n_distinct(id[days > 7]))
```

Nine of the 300 planned doses land on a day when the neutrophil count
is below 1500 cells/mm^3, and 19 cycles across four subjects sit at
grade 4 for more than a week.  The [Taxotere
label](https://products.sanofi.us/taxotere/taxotere.pdf) tells you
what to do: don't dose below 1500 cells/mm^3, and after
febrile neutropenia or neutrophils under 500 cells/mm^3 for more than
one week, step 100 mg/m^2 down to 75 mg/m^2, and then to 55 mg/m^2 if it
happens again.

That is three coupled rules, and none of them can be written into
`et()` because the answer depends on the solution.

## Writing the label into the model

The protocol needs a memory: how long this cycle has been at grade 4,
which dose level we are on, when the next cycle is due.

`rxode2` can track this using sticky variables.  Most left-hand-side
variables in a model block are
[**sticky**](https://nlmixr2.github.io/rxode2/articles/rxode2-sticky.html):
rather than being recomputed from nothing at each time point, they
keep whatever value you last gave them, and they start out `NA` for
each new individual.  That gives the standard idiom -- test for `NA`
to initialise, then update in place:

```r
if (is.na(level)) {
  level <- 1        # NA only at the first time point of each subject
}
```

Sticky variables make the protocol easy to read:
updating one is an ordinary assignment.

Some things you need to keep in mind:

- The block runs once per *record*, not at internal solver steps

    - The weekly assessment fires with both `t %% 168 == 0` and `t >= nextDue`.
      Then the `nextDue` assessment is incremented.

None of this touches the PK or the PD.  To make sure that the model
isn't mis-typed, or copied or even so that if it i it will retain the
dosing schedule, the best solution is to use model piping  to appends the
dosing protocol to `doce` and leave the model alone:

```{r protocol}
## Pipe the protocol onto the model we already have instead of writing
## the PK/PD out a second time.
##
## `auto = FALSE` keeps the auto selection of parameters off, so you
## can set the one new `bsa` parameter

doceProtocol <- doce |>
  model({
    ## --- protocol memory, in sticky variables ---------------------
    ## NA only at the first time point of each subject; after that
    ## these carry over from one time point to the next by themselves.
    if (is.na(level)) {
      level   <- 1        # 1 = 100, 2 = 75, 3 = 55 mg/m^2
      nextDue <- 21*24    # hour the next cycle is due
      dayLow  <- 0        # days at grade 4 since the last dose
      lastT   <- 0
    }
    dayLow <- dayLow + (anc < 0.5)*(time - lastT)/24
    lastT  <- time

    ## dose level for the cycle we are about to start
    newLevel <- level
    if (dayLow > 7) {
      newLevel <- level + 1
    }
    if (newLevel > 3) {
      newLevel <- 3
    }
    mgm2 <- 100
    if (newLevel > 1.5) {
      mgm2 <- 75
    }
    if (newLevel > 2.5) {
      mgm2 <- 55
    }
    doseMg  <- mgm2*bsa
    startMg <- 100*bsa

    ## --- cycle 1, then a weekly assessment ------------------------
    if (t == 0) {
      infuseDur(startMg, 1, central)
    }
    if (t > 0 && t < 126*24 && t %% 168 == 0 && t >= nextDue) {
      if (anc >= 1.5) {
        infuseDur(doseMg, 1, central)
        level   <- newLevel
        dayLow  <- 0
        nextDue <- t + 21*24
      } else {
        nextDue <- t + 7*24   # hold a week and reassess
      }
    }
  }, append = TRUE, auto = FALSE) |>
  ini(bsa <- 1.8)                     # m^2
```

Note that the amounts are assigned to plain variables first --
`infuseDur()` wants a single symbol for `amt`, not an expression --
and that the event table now carries *no doses at all*.  Every dose is
pushed by the model:

```{r pop-adaptive}
popAdapt <- rxSolve(doceProtocol,
                    et(seq(0, 126*24, by = 6)) |>
                      et(id = 1:n),
                    params = pop,
                    maxExtra = 500,
                    returnType="data.frame")
```

The [sticky variable
vignette](https://nlmixr2.github.io/rxode2/articles/rxode2-sticky.html)
has the other standard uses -- running $C_{max}$ and $T_{max}$, time
after dose, anything you would otherwise reach for a state to
remember:

```r
if (is.na(cMax)) {
  cMax <- 0
  tMax <- 0
}
if (cp > cMax) {
  cMax <- cp
  tMax <- time
}
```

One thing to keep in mind: `dayLow` is mildly grid-dependent.

One nice thing is the model's decisions can be read straight back out
of the solution, because `nextDue` advances by 504 h when a dose is
given and 168 h when a cycle is held. It's a sticky variable, so it
changes on the same row where the decision was made — look behind, and
the ANC you report is the one the protocol actually acted on.

Because `nextDue` moves by 504 h when a dose is given and 168 h when a
cycle is held, the decisions the model made can be read straight back
out of the solution.  `nextDue` is a sticky variable, so it changes on
the row the decision was taken -- look behind, and the ANC you report
is the one the protocol actually acted on.

```{r decisions}
decisions <- popAdapt |>
  group_by(id) |>
  arrange(id, time) |>
  mutate(step = c(0, diff(nextDue))) |>   # sticky: changes on the decision row
  ungroup()

given <- decisions |> filter(step == 504)
held  <- decisions |> filter(step == 168)

data.frame(what = c("doses given (cycles 2+)", "lowest ANC at a dose",
                    "cycles held", "subjects ever held"),
           value = c(nrow(given), round(min(given$anc), 2),
                     nrow(held), n_distinct(held$id)))

table(`mg/m2 given` = given$mgm2)
```

This can be handy in summarizing what happened in the simulated
protocol.

In this case, no dose is now given below 1500 cells/mm^3 -- the lowest
count at any administration is `r round(min(given$anc), 2)` × 10^9/L
-- and twenty of the 249 later doses are given at a reduced level.

Counting 21-day windows the same way as before, the number spent at
grade 4 for more than a week falls from 19 to five:

```{r grade4}
decisions |>
  mutate(window = pmin(6, floor(time/504) + 1)) |>
  group_by(id, window) |>
  summarise(days = sum(anc < 0.5)*6/24, .groups = "drop") |>
  summarise(`windows with >7 d grade 4` = sum(days > 7))
```

## The five patients it happened to

Four subjects were dose-reduced and one had cycles held.  Two of them
are worth looking at.

Subject 47 is slow to recover (mean transit time 203 h): under the
fixed schedule the count is still between 1.1 and 1.5 × 10^9/L on every
planned dosing day from cycle 2 onwards.  Subject 24 clears docetaxel slowly
(`CL` 20.3 L/h) and is sensitive to it (slope 31.7 mL/µg), spends over
nine days at grade 4 in cycle 1 and is still over the line two cycles
after the first reduction -- the only subject in the trial to walk the
whole 100 → 75 → 55 ladder.

```{r compare, fig.height=5.2}
ids <- c(24, 47)

lab <- c("24" = "subject 24 -- dose reduced 100 -> 75 -> 55 mg/m2",
         "47" = "subject 47 -- each cycle delayed one week")

bind_rows(
  popFixed |> filter(id %in% ids) |> mutate(schedule = "fixed q3w"),
  popAdapt |> filter(id %in% ids) |> mutate(schedule = "label protocol")) |>
  mutate(who = lab[as.character(id)]) |>
  ggplot(aes(time/24, anc, colour = schedule)) +
  geom_hline(yintercept = c(0.5, 1.5), linetype = 2, colour = "grey50") +
  geom_line(linewidth = 0.7) +
  facet_wrap(~who, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("fixed q3w" = "grey45",
                                 "label protocol" = "#2166ac")) +
  labs(x = "day", y = "ANC (10^9/L)", colour = NULL)
```

The two panels show the two things a protocol can change: when and how-much.

Subject 47 illustrates *when*: the blue curve's cycles are 28 days apart instead of 21.

Subject 24 is about *how much* -- reducing the dose does not lift the nadir, it shortens the time spent there.  Its own cycles go from 9.2 days at grade 4 before the first reduction to 5.8 after the second; s

Either way, the grey and blue curves are the same model, the same
parameters and the same 126 days.  The only difference is that in the
blue one the event history was written by the solver.

## Notes from actually using it

A few things that will save you an some time:

- **Decision times must exist.**  The logic runs at output times.  If
  the assessment is not in `et()`, pin it with `mtime()`.

- **Guard against re-triggering.**  `anc >= 1.5` is true for weeks.
  Anchoring on `t %% 168 == 0` is what turns a standing condition into
  a single decision.

- **Bookkeeping belongs in sticky variables, not states.**  Counters,
  dose levels and "when is the next visit" carry over from one time
  point to the next on their own, you could use states instead but it
  takes more time to solve the system.  `is.na()` is how you
  initialise them.

- **Currently the dose `amt` wants a symbol.**  Compute `doseMg <-
  mgm2*bsa` on its own line; `infuseDur(mgm2*bsa, ...)` will not
  parse. In the future this may change.

- **Set `maxExtra`.**  It is a cheap way to find out that your
  condition fires 4000 times instead of six.

- **At the last output time the event is dropped but the sticky
  assignment is not.**  There is no row left to put a pushed event in,
  so the dose never happens -- but `nextDue` and `dayLow` are ordinary
  assignments and go through anyway.

- **Reading decisions back out of the state vector only works if the
  decision times are on the grid.**  `filter(step == 504)` is exact
  equality on a float; it survives here because 168 and 504 land
  exactly on a 6 h grid.  Solve on `by = 5` and the summary silently
  reports zero doses rather than erroring.

- **`linCmt()` works too**, and `odeToLin()` will convert a *linear*
  ODE model that uses adaptive dosing into the analytic form -- worth
  it when you are running that kind of model over many trial designs.
  It cannot help the example above, whose PD is nonlinear.

## Where to read more

The [adaptive dosing
vignette](https://nlmixr2.github.io/rxode2/articles/adaptive-dosing.html)
is the reference: every helper, the `evid_()` low-level interface,
phantom dosing for transit models, adaptive sampling with `obs()`, and
the `linCmt()` and `odeToLin()` story.

The obvious next step is that these protocol rules are now part of the
model, which means they can be part of a trial simulation -- compare
two dose-modification tables, or ask what a different assessment
schedule would have bought you, without leaving `rxode2`.

## References

- Friberg LE, Henningsson A, Maas H, Nguyen L, Karlsson MO. [Model of
  chemotherapy-induced myelosuppression with parameter consistency
  across drugs](https://ascopubs.org/doi/10.1200/JCO.2002.02.140).
  *J Clin Oncol*. 2002;20(24):4713-4721.
- Ozawa K, Minami H, Sato H. [Population pharmacokinetic and
  pharmacodynamic analysis for time courses of docetaxel-induced
  neutropenia in Japanese cancer
  patients](https://pubmed.ncbi.nlm.nih.gov/17888031/).
  *Cancer Sci*. 2007;98(12):1985-1992.
- [TAXOTERE (docetaxel) prescribing
  information](https://products.sanofi.us/taxotere/taxotere.pdf),
  Sanofi-Aventis.
