---
title: "Adaptive dosing in rxode2"
subtitle: "When the protocol depends on the patient, the event table cannot be written in advance"
author: Matthew Fidler
date: '2026-09-09'
slug: []
categories: [rxode2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, fig.width = 7.8, fig.height = 4.6,
                      dpi = 110, out.width = "100%")
library(rxode2)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
```

Every `rxode2` simulation you have ever run started the same way: you
wrote down the doses first, and then you solved.  The event table is
fixed before the solver takes its first step.

That is fine for a phase I schedule.  It is not fine for most of what
actually happens in the clinic.  Real protocols read a lab value and
then decide: hold this cycle, cut the dose one level, start the
rescue infusion, draw an extra sample.  The dose that a patient
receives on day 42 is a *function of the trajectory* between day 0 and
day 42 -- which means you cannot write it into an event table, because
you do not know it until you have solved.

`rxode2` can now modify an individual's event history *while the model
is being solved*.  The new [adaptive dosing
vignette](https://nlmixr2.github.io/rxode2/articles/adaptive-dosing.html)
covers the whole feature; this post is about why you would want it,
using a published myelosuppression model and a real dose-modification
table from a product label.  Everything below needs `rxode2` >= 5.1.7 --
several fixes in that release are load-bearing for these numbers.

## The idea in three lines

Decision logic goes inside the `model({})` block, it runs at the times
the solver visits, and when a condition is met you push a new event:

```{r taste}
titrate <- function() {
  ini({
    ka <- 0.5; cl <- 1; v <- 10
  })
  model({
    d/dt(depot)   <- -ka*depot
    d/dt(central) <-  ka*depot - cl/v*central
    cp <- central/v
    # from day 1 on, top up if the daily check finds us below target
    if (t > 0 && t %% 24 == 0 && cp < 1) {
      bolus(50)
    }
  })
}

titrateSolve <- titrate |>
  rxSolve(et(amt = 100, time = 0) |> et(seq(0, 96, by = 1)))

ggplot(titrateSolve, aes(time, cp)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  geom_line(linewidth = 0.7, colour = "#2166ac") +
  labs(x = "time (h)", y = "concentration",
       title = "Top up whenever the daily check finds us below target")
```

The helpers all push a standard NONMEM-style event, so there is
nothing exotic downstream:

| helper | `evid` | what it pushes |
|---|---|---|
| `bolus()` | 1 | a bolus dose, optionally with `ii`/`addl`/`ss` |
| `infuse()` | 1 | a fixed-**rate** infusion |
| `infuseDur()` | 1 | a fixed-**duration** infusion |
| `reset()` | 3 | a system reset |
| `replace()` | 5 | set a compartment to a value |
| `multiply()` | 6 | scale a compartment |
| `phantom()` | 7 | transit dosing bookkeeping (`tad()`, `podo()`) without mass |
| `obs()` | 0 | extra observation rows |
| `evid_()` | any | the low-level interface |

(`reset()`, `replace()` and `multiply()` are model-block functions
only -- they are not exported, so `?replace` will show you base R's.)

Two things are worth knowing before you write anything serious.
First, the condition is only tested at times the solver actually stops
at, so the decision times have to be in the event table, or pinned
with an `mtime()` assignment:

```r
mtime(check48) <- 48        # note the assignment -- mtime(48) will not parse
if (t == check48 && cp < 1) {
  bolus(50)
}
```

Second, a condition like `cp < 1` stays true for many steps in a row,
so anchor it to a visit -- `t %% 24 == 0`, `t == check48` -- or you
will push a dose at every output time.  `rxSolve()`'s `maxExtra`
argument is the seatbelt for when you get that wrong.

The `t > 0` above matters too: at time zero `cp` is still `0`, so
without it the model tops up on top of the dose in the event table.
And a decision taken at the *last* output time has nowhere to put its
event, so it is silently dropped -- the top-up the model wants at hour
96 never happens here.

## A model that needs this

For a worked example we want a published model where the protocol is
genuinely closed-loop.  Chemotherapy-induced neutropenia is the
canonical one.

[Friberg et al.
(2002)](https://ascopubs.org/doi/10.1200/JCO.2002.02.140) described
myelosuppression with a proliferating compartment under a linear drug
effect, three transit compartments for maturation, and a feedback term
$(\mathrm{Circ}_0/\mathrm{Circ})^\gamma$ that drives the rebound.  For
docetaxel we can take a self-consistent set of estimates from [Ozawa,
Minami and Sato (2007)](https://pubmed.ncbi.nlm.nih.gov/17888031/),
who fit a three-compartment PK model together with the Friberg
neutropenia model in Japanese cancer patients: `CL` 35.7 L/h, `V1`
6.94 L, `Q2` 5.58 L/h, `V2` 7.39 L, `Q3` 12.5 L/h, `V3` 225 L, mean
transit time 113 h, $\gamma$ 0.196 and a slope of 17.9 mL/µg.  Their
baseline neutrophil count was each patient's own observed value, so
for the typical subject we simply take 5.05 × 10⁹/L.  (`gam` rather
than `gamma`, because `gamma()` is a function `rxode2` supports and
you do not want a parameter shadowing it.)

```{r model}
doce <- function() {
  ini({
    cl <- 35.7; v1 <- 6.94; q2 <- 5.58; v2 <- 7.39; q3 <- 12.5; v3 <- 225
    circ0 <- 5.05; mtt <- 113; gam <- 0.196; slope <- 17.9
  })
  model({
    ktr <- 4/mtt
    cp  <- central/v1

    d/dt(central) <- -(cl/v1)*central - (q2/v1)*central + (q2/v2)*periph1 -
                      (q3/v1)*central + (q3/v3)*periph2
    d/dt(periph1) <-  (q2/v1)*central - (q2/v2)*periph1
    d/dt(periph2) <-  (q3/v1)*central - (q3/v3)*periph2

    circS <- max(circ, 1e-4)
    d/dt(prol) <- ktr*prol*(1 - slope*cp)*(circ0/circS)^gam - ktr*prol
    d/dt(tr1)  <- ktr*(prol - tr1)
    d/dt(tr2)  <- ktr*(tr1  - tr2)
    d/dt(tr3)  <- ktr*(tr2  - tr3)
    d/dt(circ) <- ktr*(tr3  - circ)

    prol(0) <- circ0; tr1(0) <- circ0; tr2(0) <- circ0
    tr3(0) <- circ0; circ(0) <- circ0

    anc <- circ
  })
}
```

Six cycles of docetaxel 100 mg/m² as a one-hour infusion every three
weeks, in a 1.8 m² typical patient, is an ordinary event table:

```{r typical}
bsa <- 1.8

fixedEv <- et(amt = 100*bsa, dur = 1, cmt = "central", ii = 21*24, addl = 5) |>
  et(seq(0, 126*24, by = 6))

typical <- rxSolve(doce, fixedEv)

ggplot(typical, aes(time/24, anc)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = c(0.5, 1.5), linetype = 2, colour = "grey40") +
  labs(x = "day", y = "ANC (10^9/L)",
       title = "Docetaxel 100 mg/m2 q3w, typical patient")
```

The nadir is `r round(min(typical$anc), 3)` × 10⁹/L on day
`r round(typical$time[which.min(typical$anc)]/24, 1)`, sitting just
above the grade-4 line, and the count is back over 1.5 × 10⁹/L well
before the next cycle: deep grade 3, no dose held, no dose cut.  (The
label puts the median nadir nearer day 7; the 113 h mean transit time
of this parameterisation is slower than the ~89 h Friberg reported for
docetaxel, and it shows.)  Nothing here needs adaptive dosing.

## The protocol only matters for the patients in the tail

The reason the label carries a dose-modification table is the other
patients.  Add between-subject variability -- 30% on clearance, 35% on
the slope, 25% on baseline and 20% on mean transit time, illustrative
rather than taken from the paper -- and simulate 50 subjects on the
same fixed schedule:

```{r pop-fixed}
set.seed(20260909)
n <- 50
pop <- data.frame(
  id    = 1:n,
  cl    = 35.7*exp(rnorm(n, 0, 0.30)),
  slope = 17.9*exp(rnorm(n, 0, 0.35)),
  circ0 = 5.05*exp(rnorm(n, 0, 0.25)),
  mtt   =  113*exp(rnorm(n, 0, 0.20)))

popEv <- fixedEv |> et(id = 1:n)

popFixed <- rxSolve(doce, popEv, params = pop) |> as.data.frame()

doseDays <- (0:5)*21*24

popFixed |>
  filter(time %in% doseDays) |>
  summarise(`doses at ANC < 1.5` = sum(anc < 1.5), n = n())

popFixed |>
  mutate(cycle = pmin(6, floor(time/504) + 1)) |>
  group_by(id, cycle) |>
  summarise(days = sum(anc < 0.5)*6/24, .groups = "drop") |>
  summarise(`cycles with >7 d grade 4` = sum(days > 7),
            `subjects affected` = n_distinct(id[days > 7]))
```

Nine of the 300 planned doses land on a day when the neutrophil count
is below 1500 cells/mm³, and 19 cycles across four subjects sit at
grade 4 for more than a week.  The [Taxotere
label](https://products.sanofi.us/taxotere/taxotere.pdf) tells you
what to do about both: do not dose below 1500 cells/mm³, and after
febrile neutropenia or neutrophils under 500 cells/mm³ for more than
one week, step 100 mg/m² down to 75 mg/m², and then to 55 mg/m² if it
happens again.

That is three coupled rules, and none of them can be written into
`et()` because the answer depends on the solution.

## Writing the label into the model

The bookkeeping the protocol needs -- how long this cycle has been at
grade 4, which dose level we are on, when the next cycle is due --
lives in compartments with zero derivatives, which the adaptive
helpers can then update with `replace()`.  Weekly assessments come
from the output grid, and the `t %% 168 == 0` guard makes sure each
one fires exactly once.

```{r protocol}
doceProtocol <- function() {
  ini({
    cl <- 35.7; v1 <- 6.94; q2 <- 5.58; v2 <- 7.39; q3 <- 12.5; v3 <- 225
    circ0 <- 5.05; mtt <- 113; gam <- 0.196; slope <- 17.9
    bsa <- 1.8
  })
  model({
    ktr <- 4/mtt
    cp  <- central/v1

    d/dt(central) <- -(cl/v1)*central - (q2/v1)*central + (q2/v2)*periph1 -
                      (q3/v1)*central + (q3/v3)*periph2
    d/dt(periph1) <-  (q2/v1)*central - (q2/v2)*periph1
    d/dt(periph2) <-  (q3/v1)*central - (q3/v3)*periph2

    circS <- max(circ, 1e-4)
    d/dt(prol) <- ktr*prol*(1 - slope*cp)*(circ0/circS)^gam - ktr*prol
    d/dt(tr1)  <- ktr*(prol - tr1)
    d/dt(tr2)  <- ktr*(tr1  - tr2)
    d/dt(tr3)  <- ktr*(tr2  - tr3)
    d/dt(circ) <- ktr*(tr3  - circ)

    prol(0) <- circ0; tr1(0) <- circ0; tr2(0) <- circ0
    tr3(0) <- circ0; circ(0) <- circ0

    anc <- circ

    ## --- protocol bookkeeping -------------------------------------
    d/dt(dayLow)  <- (anc < 0.5)/24   # days at grade 4 since last dose
    d/dt(level)   <- 0                # 1 = 100, 2 = 75, 3 = 55 mg/m2
    d/dt(nextDue) <- 0                # hour the next cycle is due
    level(0)   <- 1
    nextDue(0) <- 21*24

    ## dose level for the cycle we are about to start
    newLevel <- level
    if (dayLow > 7) {
      newLevel <- level + 1
    }
    if (newLevel > 3) {
      newLevel <- 3
    }
    mgm2 <- 100
    if (newLevel > 1.5) {
      mgm2 <- 75
    }
    if (newLevel > 2.5) {
      mgm2 <- 55
    }

    doseMg  <- mgm2*bsa
    startMg <- 100*bsa
    zeroDay <- 0
    dueNext <- t + 21*24
    dueHold <- t + 7*24

    ## --- cycle 1, then a weekly assessment ------------------------
    if (t == 0) {
      infuseDur(startMg, 1, central)
    }
    if (t > 0 && t %% 168 == 0 && t >= nextDue) {
      if (anc >= 1.5) {
        infuseDur(doseMg, 1, central)
        replace(newLevel, level)
        replace(zeroDay,  dayLow)
        replace(dueNext,  nextDue)
      } else {
        replace(dueHold, nextDue)   # hold a week and reassess
      }
    }

    doseLevel <- level
  })
}
```

Note that the amounts are assigned to plain variables first --
`infuseDur()` wants a single symbol for `amt`, not an expression --
and that the event table now carries *no doses at all*.  Every dose is
pushed by the model:

```{r pop-adaptive}
popAdapt <- rxSolve(doceProtocol,
                    et(seq(0, 126*24, by = 6)) |> et(id = 1:n),
                    params = pop, maxExtra = 500) |>
  as.data.frame()
```

This is a deliberately literal reading of two lines of the label, not
a full protocol.  `dayLow` counts days below 0.5 × 10⁹/L whether or
not they are consecutive, where "for more than one week" implies a
single episode; febrile neutropenia, the label's other trigger, is not
modelled at all; and a patient still in trouble at 55 mg/m² should be
considered for discontinuation, whereas this model caps the level and
keeps dosing.  Each of those is a few more lines in the same block.

Because `nextDue` moves by 504 h when a dose is given and 168 h when a
cycle is held, the decisions the model made can be read straight back
out of the solution.  One subtlety: a pushed event takes effect in the
row *after* the record that pushed it, so we look ahead rather than
behind -- otherwise every ANC we report is one output step stale.

```{r decisions}
decisions <- popAdapt |>
  group_by(id) |>
  arrange(time) |>
  mutate(step = c(diff(nextDue), 0)) |>   # look ahead, not behind
  ungroup()

given <- decisions |> filter(step == 504)
held  <- decisions |> filter(step == 168)

data.frame(what = c("doses given (cycles 2+)", "lowest ANC at a dose",
                    "cycles held", "subjects ever held"),
           value = c(nrow(given), round(min(given$anc), 2),
                     nrow(held), n_distinct(held$id)))

table(`mg/m2 given` = given$mgm2)
```

No dose is now given below 1500 cells/mm³ -- the lowest count at any
administration is `r round(min(given$anc), 2)` × 10⁹/L -- and twenty
of the 249 later doses are given at a reduced level.  Counting 21-day
windows the same way as before, the number spent at grade 4 for more
than a week falls from 19 to four:

```{r grade4}
decisions |>
  mutate(window = pmin(6, floor(time/504) + 1)) |>
  group_by(id, window) |>
  summarise(days = sum(anc < 0.5)*6/24, .groups = "drop") |>
  summarise(`windows with >7 d grade 4` = sum(days > 7))
```

Two caveats, because that number flatters the protocol.  All four
survivors are *cycle 1*, which is identical in both arms by
construction -- the protocol cannot act until it has seen a cycle go
badly.  And once a cycle is delayed a fixed 21-day window stops
lining up with a cycle, so this is a calendar comparison rather than a
cycle-by-cycle one.  The blunter summary is that total time at grade 4
across the trial falls from `r sum(popFixed$anc < 0.5)*6/24` to
`r sum(popAdapt$anc < 0.5)*6/24` days, about 9%: the protocol turns
nine-day episodes into six-day ones, which happens to straddle the
seven-day threshold.

## The five patients it happened to

Four subjects were dose-reduced and one had cycles held.  Two of them
are worth looking at.

Subject 47 is slow to recover (mean transit time 203 h): under the
fixed schedule the count is still between 1.1 and 1.5 × 10⁹/L on every
planned dosing day from cycle 2 onwards, so every one of those doses
is one the label says not to give.  Subject 24 clears docetaxel slowly
(`CL` 20.3 L/h) and is sensitive to it (slope 31.7 mL/µg), spends over
nine days at grade 4 in cycle 1 and again after the first reduction --
the only subject in the trial to walk the whole 100 → 75 → 55 ladder.

```{r compare, fig.height=5.2}
ids <- c(24, 47)

lab <- c("24" = "subject 24 -- dose reduced 100 -> 75 -> 55 mg/m2",
         "47" = "subject 47 -- each cycle delayed one week")

bind_rows(
  popFixed |> filter(id %in% ids) |> mutate(schedule = "fixed q3w"),
  popAdapt |> filter(id %in% ids) |> mutate(schedule = "label protocol")) |>
  mutate(who = lab[as.character(id)]) |>
  ggplot(aes(time/24, anc, colour = schedule)) +
  geom_hline(yintercept = c(0.5, 1.5), linetype = 2, colour = "grey50") +
  geom_line(linewidth = 0.7) +
  facet_wrap(~who, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("fixed q3w" = "grey45",
                                 "label protocol" = "#2166ac")) +
  labs(x = "day", y = "ANC (10^9/L)", colour = NULL)
```

The two panels show the two things a protocol can change.  Subject 47
is about *when*: the blue curve's cycles are 28 days apart instead of
21, and each dose is given from a count around 3--4 × 10⁹/L rather
than 1.1--1.5.  Subject 24 is about *how much*, and it is a more
sober picture -- reducing the dose does not lift the nadir off the
floor, it shortens the time spent there, from 9.2 days at grade 4 per
cycle to 5.8.

Either way, the grey and blue curves are the same model, the same
parameters and the same 126 days.  The only difference is that in the
blue one the event history was written by the solver.

## Notes from actually using it

A few things that will save you an afternoon:

- **Decision times must exist.**  The logic runs at output times.  If
  the assessment is not in `et()`, pin it with `mtime()`.

- **Guard against re-triggering.**  `anc >= 1.5` is true for weeks.
  Anchoring on `t %% 168 == 0` is what turns a standing condition into
  a single decision.

- **Bookkeeping is just a compartment with a zero derivative.**
  Counters, dose levels and "when is the next visit" all live in the
  state vector, and `replace()` is how you set them.

- **`amt` wants a symbol.**  Compute `doseMg <- mgm2*bsa` on its own
  line; `infuseDur(mgm2*bsa, ...)` will not parse.

- **Set `maxExtra`.**  It is a cheap way to find out that your
  condition fires 4000 times instead of six.

- **A pushed event lands in the row after the one that pushed it.**
  That is why the read-out above looks ahead with
  `c(diff(nextDue), 0)`.  Look behind instead and every value you
  report is one output step stale -- here that turned the lowest ANC
  at a dose into 1.90 instead of 1.72.

- **A decision at the last output time is dropped**, because there is
  no row left to put the event in.  Check that your horizon runs past
  the last decision you care about.

- **Reading decisions back out of the state vector only works if the
  decision times are on the grid.**  `filter(step == 504)` is exact
  equality on a float; it survives here because 168 and 504 land
  exactly on a 6 h grid.  Solve on `by = 5` and the summary silently
  reports zero doses rather than erroring.

- **`linCmt()` works too**, and `odeToLin()` will convert a *linear*
  ODE model that uses adaptive dosing into the analytic form -- worth
  it when you are running that kind of model over many trial designs.
  It cannot help the example above, whose PD is nonlinear.

## Where to read more

The [adaptive dosing
vignette](https://nlmixr2.github.io/rxode2/articles/adaptive-dosing.html)
is the reference: every helper, the `evid_()` low-level interface,
phantom dosing for transit models, adaptive sampling with `obs()`, and
the `linCmt()` and `odeToLin()` story.

The obvious next step is that these protocol rules are now part of the
model, which means they can be part of a trial simulation -- compare
two dose-modification tables, or ask what a different assessment
schedule would have bought you, without leaving `rxode2`.

## References

- Friberg LE, Henningsson A, Maas H, Nguyen L, Karlsson MO. [Model of
  chemotherapy-induced myelosuppression with parameter consistency
  across drugs](https://ascopubs.org/doi/10.1200/JCO.2002.02.140).
  *J Clin Oncol*. 2002;20(24):4713-4721.
- Ozawa K, Minami H, Sato H. [Population pharmacokinetic and
  pharmacodynamic analysis for time courses of docetaxel-induced
  neutropenia in Japanese cancer
  patients](https://pubmed.ncbi.nlm.nih.gov/17888031/).
  *Cancer Sci*. 2007;98(12):1985-1992.
- [TAXOTERE (docetaxel) prescribing
  information](https://products.sanofi.us/taxotere/taxotere.pdf),
  Sanofi-Aventis.
