---
title: "Adaptive dosing in rxode2"
subtitle: "When the protocol depends on the patient, the event table cannot be written in advance"
author: Matthew Fidler
date: '2026-09-09'
slug: []
categories: [rxode2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7.8, fig.height = 4.6,
                      dpi = 110, out.width = "100%")
library(rxode2)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
```

Every `rxode2` simulation you have ever run started the same way: you
coded down the doses first, and then you solved.  The event table is
fixed before the solver takes its first step.

That is fine for a fixed schedule.  It is not fine for most adaptive
dosing in the clinic.  Real protocols may read a lab value and then
decide: hold this cycle, cut the dose one level, start the rescue
infusion, draw an extra sample.  The dose that a patient receives on
day 42 is a *function of the trajectory* between day 0 and day 42 --
which means you cannot write it into an event table, because you do
not know it until you have solved.

`rxode2` can now modify an individual's event history *while the model
is being solved*.  The new [adaptive dosing
vignette](https://nlmixr2.github.io/rxode2/articles/adaptive-dosing.html)
covers the whole feature; this post
uses the famous Friburg myelosuppression model and a real dose-modification
table from a product label.  Everything below needs `rxode2` >= 5.1.7.

## The idea in three lines

Decision logic goes inside the `model({})` block, it runs at the times
the solver visits, and when a condition is met you push a new event:

```{r taste}
titrate <- function() {
  ini({
    ka <- 0.5
    cl <- 1
    v <- 10
  })
  model({
    d/dt(depot)   <- -ka*depot
    d/dt(central) <-  ka*depot - cl/v*central
    cp <- central/v
    # from day 1 on, top up if the daily check finds us below target

    # This is the code to say every 24 hours check to see if cp < 1,
    # then dose if needed
    if (t > 0 && t %% 24 == 0 && cp < 1) {
      bolus(50)
    }
  })
}

titrateSolve <- titrate |>
  rxSolve(et(amt = 100, time = 0) |> et(seq(0, 96, by = 1)))

ggplot(titrateSolve, aes(time, cp)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  geom_line(linewidth = 0.7, colour = "#2166ac") +
  theme_bw() +
  labs(x = "time (h)", y = "concentration",
       title = "Top up whenever the daily check finds us below target")
```

The helpers all push a standard NONMEM-style event, so there is
nothing exotic downstream:

| helper | `evid` | what it pushes |
|---|---|---|
| `bolus()` | 1 | a bolus dose, optionally with `ii`/`addl`/`ss` |
| `infuse()` | 1 | a fixed-**rate** infusion |
| `infuseDur()` | 1 | a fixed-**duration** infusion |
| `reset()` | 3 | a system reset |
| `replace()` | 5 | set a compartment to a value |
| `multiply()` | 6 | scale a compartment |
| `phantom()` | 7 | transit dosing bookkeeping (`tad()`, `podo()`) without mass |
| `obs()` | 0 | extra observation rows |
| `evid_()` | any | the low-level interface |

Note that the condition is only tested at times the solver actually stops
at; either the decision times have to be in the event table, or pinned
with an `mtime()` assignment:

```r
mtime(check48) <- 48        # note the assignment -- mtime(48) will not parse
if (t == check48 && cp < 1) {
  bolus(50)
}
```

A few things to keep in mind:

`cp < 1` stays true almost always (except the first dose) so:

 - alone it will fire almost every observation

 - For it to be useful, it needs to be anchored to a visit
   (`t == check48`) or to a visit schedule, like every 24 hours (`t %% 24 == 0`).

This can be easy to get wrong.  To prevent dosing at more times than
you expect, you can use `maxExtra` to only add a specific number of
doses per problem.

As a note, the `t > 0` above matters too: at time zero `cp` is still
`0`, so without it the model will add a dose at time zero, even though
the baseline dose is in the event table.


One last note, at the very end of the dosing if a new dose would be
added, it won't be shown in the observations since the model doesn't
check there or has no observations after the dose.

## A model that needs this

One of the standard places where you need adaptive dosing is cancer
treatment.  For example, chemotherapy-induced neutropenia.

[Friberg et al.
(2002)](https://ascopubs.org/doi/10.1200/JCO.2002.02.140) described
myelosuppression model under linear drug effect, three maturation
transit compartments, and a rebound feedback term
$(\mathrm{Circ}_0/\mathrm{Circ})^\gamma$.  We took the estimates of
docetaxel from [Ozawa, Minami and Sato
(2007)](https://pubmed.ncbi.nlm.nih.gov/17888031/), who fit a
three-compartment PK model together with the Friberg neutropenia model
in Japanese cancer patients: `CL` 35.7 L/h, `V1` 6.94 L, `Q2` 5.58
L/h, `V2` 7.39 L, `Q3` 12.5 L/h, `V3` 225 L, mean transit time 113 h,
$\gamma$ 0.196 and a slope of 17.9 mL/µg.  Their baseline neutrophil
count was each patient's own observed value, so for the typical
subject we simply take 5.05 × 10⁹/L.  (`gam` rather than `gamma`,
because `gamma()` is a function `rxode2` supports and you do not want
a parameter shadowing it.)

```{r model}
doce <- function() {
  ini({
    cl <- 35.7 # L/h
    v1 <- 6.94 # L
    q2 <- 5.58 # L/h
    v2 <- 7.39 # L
    q3 <- 12.5 # L/h
    v3 <- 225 # L
    circ0 <- 5.05 # baseline neurtophil count *10^9/L
    mtt <- 113 # mean transit time (hr)
    gam <- 0.196 # shape
    slope <- 17.9 # mL / ug
  })
  model({
    ktr <- 4/mtt
    cp  <- central/v1

    d/dt(central) <- -(cl/v1)*central - (q2/v1)*central + (q2/v2)*periph1 -
                      (q3/v1)*central + (q3/v3)*periph2
    d/dt(periph1) <-  (q2/v1)*central - (q2/v2)*periph1
    d/dt(periph2) <-  (q3/v1)*central - (q3/v3)*periph2

    circS <- max(circ, 1e-4)
    d/dt(prol) <- ktr*prol*(1 - slope*cp)*(circ0/circS)^gam - ktr*prol
    d/dt(tr1)  <- ktr*(prol - tr1)
    d/dt(tr2)  <- ktr*(tr1  - tr2)
    d/dt(tr3)  <- ktr*(tr2  - tr3)
    d/dt(circ) <- ktr*(tr3  - circ)

    prol(0) <- circ0
    tr1(0) <- circ0
    tr2(0) <- circ0
    tr3(0) <- circ0
    circ(0) <- circ0

    anc <- circ
  })
}
```

Six cycles of docetaxel 100 mg/m^2 as a one-hour infusion every three
weeks, in a 1.8 m^2 typical patient, is an ordinary event table:

```{r typical}
bsa <- 1.8

fixedEv <- et(amt = 100*bsa, dur = 1, cmt = "central", ii = 21*24, addl = 5) |>
  et(seq(0, 126*24, by = 6))

typical <- rxSolve(doce, fixedEv)

ggplot(typical, aes(time/24, anc)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = c(0.5, 1.5), linetype = 2, colour = "grey40") +
  theme_bw() +
  labs(x = "day", y = "ANC (10^9/L)",
       title = "Docetaxel 100 mg/m2 q3w, typical patient")
```

The nadir is `r round(min(typical$anc), 3)` * 10^9/L on day
`r round(typical$time[which.min(typical$anc)]/24, 1)`, sitting just
above the grade-4 line, and the count is back over 1.5 *10^9/L well
before the next cycle: deep grade 3, no dose held, no dose cut.
(The label puts the median nadir nearer day 7; the 113 h mean transit
time of this parameterisation is slower than the ~89 h Friberg
reported for docetaxel)  Nothing here needs adaptive
dosing.

## The protocol only matters for the patients in the tail

The reason the label carries a dose-modification table is the other
patients.  Add between-subject variability -- 30% on clearance, 35% on
the slope, 25% on baseline and 20% on mean transit time, illustrative
rather than taken from the paper -- and simulate 50 subjects on the
same fixed schedule:

```{r pop-fixed}
set.seed(20260909)
n <- 50
pop <- data.frame(
  id    = 1:n,
  cl    = 35.7*exp(rnorm(n, 0, 0.30)),
  slope = 17.9*exp(rnorm(n, 0, 0.35)),
  circ0 = 5.05*exp(rnorm(n, 0, 0.25)),
  mtt   =  113*exp(rnorm(n, 0, 0.20)))

popEv <- fixedEv |> et(id = 1:n)

popFixed <- rxSolve(doce, popEv, params = pop) |> as.data.frame()

doseDays <- (0:5)*21*24

popFixed |>
  filter(time %in% doseDays) |>
  summarise(`doses at ANC < 1.5` = sum(anc < 1.5), n = n())

popFixed |>
  mutate(cycle = pmin(6, floor(time/504) + 1)) |>
  group_by(id, cycle) |>
  summarise(days = sum(anc < 0.5)*6/24, .groups = "drop") |>
  summarise(`cycles with >7 d grade 4` = sum(days > 7),
            `subjects affected` = n_distinct(id[days > 7]))
```

Nine of the 300 planned doses land on a day when the neutrophil count
is below 1500 cells/mm^3, and 19 cycles across four subjects sit at
grade 4 for more than a week.  The [Taxotere
label](https://products.sanofi.us/taxotere/taxotere.pdf) tells you
what to do about both: do not dose below 1500 cells/mm^3, and after
febrile neutropenia or neutrophils under 500 cells/mm^3 for more than
one week, step 100 mg/m^2 down to 75 mg/m^2, and then to 55 mg/m^2 if it
happens again.

That is three coupled rules, and none of them can be written into
`et()` because the answer depends on the solution.

## Writing the label into the model

The protocol needs a memory: how long this cycle has been at grade 4,
which dose level we are on, when the next cycle is due.  The obvious
place to put those is the state vector -- two compartments with zero
derivatives for the level and the schedule, updated with `replace()`,
and a third integrating `(anc < 0.5)` for the counter.  That works,
but it is a poor trade.  Two of the three states store a number that
never changes between doses, and the third asks the solver to
integrate a step function, which is exactly the shape adaptive step
sizing hates.

`rxode2` has a better place for them.  Most left-hand-side variables
in a model block are
[**sticky**](https://nlmixr2.github.io/rxode2/articles/rxode2-sticky.html):
rather than being recomputed from nothing at each time point, they
keep whatever value you last gave them, and they start out `NA` for
each new individual.  That gives the standard idiom -- test for `NA`
to initialise, then update in place:

```r
if (is.na(level)) {
  level <- 1        # NA only at the first time point of each subject
}
```

Sticky variables make the protocol both cheaper and easier to read:
updating one is an ordinary assignment, where updating a compartment
needs a pushed `replace()` event.

Two details in the code below are load-bearing.  The block runs once
per *record*, not at internal solver steps -- which is what makes both
the counter and the weekly guard meaningful -- but a single time can
carry several records, an observation and a dose among them.  So the
counter accumulates `(time - lastT)` rather than a fixed step: a
repeat evaluation at the same time adds exactly zero, where the naive
`dayLow + (anc < 0.5)*6/24` would double-count on every dosing day.
And what makes a weekly assessment fire once is not `t %% 168 == 0` by
itself but `t >= nextDue` together with `nextDue` being pushed forward
inside the branch.

None of this touches the PK or the PD, so there is no reason to retype
them -- model piping appends the protocol to `doce` and leaves the
model underneath alone:

```{r protocol}
## Pipe the protocol onto the model we already have instead of writing
## the PK/PD out a second time.  Two copies of the same twenty lines
## are two copies to keep in step, and the one that drifts is always
## the one you are not looking at.
##
## `auto = FALSE` matters here: without it the pipe promotes every new
## symbol to a population parameter, `t` included, and the model then
## fails to parse.  `bsa` is the one genuinely new parameter, so it
## gets added on purpose afterwards.
doceProtocol <- doce |>
  model({
    ## --- protocol memory, in sticky variables ---------------------
    ## NA only at the first time point of each subject; after that
    ## these carry over from one time point to the next by themselves.
    if (is.na(level)) {
      level   <- 1        # 1 = 100, 2 = 75, 3 = 55 mg/m^2
      nextDue <- 21*24    # hour the next cycle is due
      dayLow  <- 0        # days at grade 4 since the last dose
      lastT   <- 0
    }
    dayLow <- dayLow + (anc < 0.5)*(time - lastT)/24
    lastT  <- time

    ## dose level for the cycle we are about to start
    newLevel <- level
    if (dayLow > 7) {
      newLevel <- level + 1
    }
    if (newLevel > 3) {
      newLevel <- 3
    }
    mgm2 <- 100
    if (newLevel > 1.5) {
      mgm2 <- 75
    }
    if (newLevel > 2.5) {
      mgm2 <- 55
    }
    doseMg  <- mgm2*bsa
    startMg <- 100*bsa

    ## --- cycle 1, then a weekly assessment ------------------------
    if (t == 0) {
      infuseDur(startMg, 1, central)
    }
    if (t > 0 && t < 126*24 && t %% 168 == 0 && t >= nextDue) {
      if (anc >= 1.5) {
        infuseDur(doseMg, 1, central)
        level   <- newLevel
        dayLow  <- 0
        nextDue <- t + 21*24
      } else {
        nextDue <- t + 7*24   # hold a week and reassess
      }
    }
  }, append = TRUE, auto = FALSE) |>
  ini(bsa <- 1.8)                     # m^2
```

Note that the amounts are assigned to plain variables first --
`infuseDur()` wants a single symbol for `amt`, not an expression --
and that the event table now carries *no doses at all*.  Every dose is
pushed by the model:

```{r pop-adaptive}
popAdapt <- rxSolve(doceProtocol,
                    et(seq(0, 126*24, by = 6)) |> et(id = 1:n),
                    params = pop, maxExtra = 500) |>
  as.data.frame()
```

The whole protocol -- three pieces of memory, a dose ladder and a
scheduler -- added no states at all:

```{r states}
rxState(doce)
rxState(doceProtocol)
```

Written with states instead, the same model carries 11 rather than 8,
and solving this trial takes about a third longer for bookkeeping the
solver should never have seen.  The
[sticky variable
vignette](https://nlmixr2.github.io/rxode2/articles/rxode2-sticky.html)
has the other standard uses -- running $C_{max}$ and $T_{max}$, time
after dose, anything you would otherwise reach for a state to
remember:

```r
if (is.na(cMax)) {
  cMax <- 0
  tMax <- 0
}
if (cp > cMax) {
  cMax <- cp
  tMax <- time
}
```

One consequence worth being explicit about: `dayLow` is now summed
over the output grid instead of integrated, so it counts what the
assessment schedule sees rather than the exact area under the
indicator.  That makes it mildly grid-dependent.  Everything that
follows -- 249 doses, the lowest count at a dose, four cycles held in
one subject, four subjects reduced -- is identical whether you solve
on a 1, 2, 4, 6 or 12 h grid.  What moves is the pair of numbers
sitting on the seven-day line: the 55/75 mg/m^2 split is 3/17 here and
4/16 on a 1 or 4 h grid, and the count of 21-day windows over a week
at grade 4 is five here and four on a 4 h grid.  In the same way,
subject 24 below reaches its second reduction one cycle later than the
ODE version did.

This is a deliberately literal reading of two lines of the label, not
a full protocol.  `dayLow` counts days below 0.5 × 10⁹/L whether or
not they are consecutive, where "for more than one week" implies a
single episode; febrile neutropenia, the label's other trigger, is not
modelled at all; and a patient still in trouble at 55 mg/m² should be
considered for discontinuation, whereas this model caps the level and
keeps dosing.  Each of those is a few more lines in the same block.

Because `nextDue` moves by 504 h when a dose is given and 168 h when a
cycle is held, the decisions the model made can be read straight back
out of the solution.  `nextDue` is a sticky variable, so it changes on
the row the decision was taken -- look behind, and the ANC you report
is the one the protocol actually acted on.  (Had it been a compartment
updated by `replace()`, the change would land one row *later*, and
this would have to look ahead instead.)

```{r decisions}
decisions <- popAdapt |>
  group_by(id) |>
  arrange(id, time) |>
  mutate(step = c(0, diff(nextDue))) |>   # sticky: changes on the decision row
  ungroup()

given <- decisions |> filter(step == 504)
held  <- decisions |> filter(step == 168)

data.frame(what = c("doses given (cycles 2+)", "lowest ANC at a dose",
                    "cycles held", "subjects ever held"),
           value = c(nrow(given), round(min(given$anc), 2),
                     nrow(held), n_distinct(held$id)))

table(`mg/m2 given` = given$mgm2)
```

No dose is now given below 1500 cells/mm³ -- the lowest count at any
administration is `r round(min(given$anc), 2)` × 10⁹/L -- and twenty
of the 249 later doses are given at a reduced level.  (249, not 250,
because subject 47's four one-week holds push its sixth cycle past the
end of the trial.)  Counting 21-day
windows the same way as before, the number spent at grade 4 for more
than a week falls from 19 to five:

```{r grade4}
decisions |>
  mutate(window = pmin(6, floor(time/504) + 1)) |>
  group_by(id, window) |>
  summarise(days = sum(anc < 0.5)*6/24, .groups = "drop") |>
  summarise(`windows with >7 d grade 4` = sum(days > 7))
```

Two caveats, because that number flatters the protocol.  Four of the
five survivors are *cycle 1*, which is identical in both arms by
construction -- the protocol cannot act until it has seen a cycle go
badly -- and the fifth is subject 24's cycle 3, before its second
reduction took effect.  And once a cycle is delayed a fixed 21-day
window stops lining up with a cycle, so this is a calendar comparison
rather than a cycle-by-cycle one.  The blunter summary is that total
time at grade 4 across the trial falls from
`r sum(popFixed$anc < 0.5)*6/24` to `r sum(popAdapt$anc < 0.5)*6/24`
days, about 9%: the protocol turns nine-day episodes into six-day
ones, which happens to straddle the seven-day threshold.

## The five patients it happened to

Four subjects were dose-reduced and one had cycles held.  Two of them
are worth looking at.

Subject 47 is slow to recover (mean transit time 203 h): under the
fixed schedule the count is still between 1.1 and 1.5 × 10⁹/L on every
planned dosing day from cycle 2 onwards, so every one of those doses
is one the label says not to give.  Subject 24 clears docetaxel slowly
(`CL` 20.3 L/h) and is sensitive to it (slope 31.7 mL/µg), spends over
nine days at grade 4 in cycle 1 and is still over the line two cycles
after the first reduction -- the only subject in the trial to walk the
whole 100 → 75 → 55 ladder.

```{r compare, fig.height=5.2}
ids <- c(24, 47)

lab <- c("24" = "subject 24 -- dose reduced 100 -> 75 -> 55 mg/m2",
         "47" = "subject 47 -- each cycle delayed one week")

bind_rows(
  popFixed |> filter(id %in% ids) |> mutate(schedule = "fixed q3w"),
  popAdapt |> filter(id %in% ids) |> mutate(schedule = "label protocol")) |>
  mutate(who = lab[as.character(id)]) |>
  ggplot(aes(time/24, anc, colour = schedule)) +
  geom_hline(yintercept = c(0.5, 1.5), linetype = 2, colour = "grey50") +
  geom_line(linewidth = 0.7) +
  facet_wrap(~who, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("fixed q3w" = "grey45",
                                 "label protocol" = "#2166ac")) +
  labs(x = "day", y = "ANC (10^9/L)", colour = NULL)
```

The two panels show the two things a protocol can change.  Subject 47
is about *when*: the blue curve's cycles are 28 days apart instead of
21, and each dose is given from a count around 3--4 × 10⁹/L rather
than 1.1--1.5.  Subject 24 is about *how much*, and it is a more
sober picture -- reducing the dose does not lift the nadir off the
floor, it shortens the time spent there.  Its own cycles go from 9.2
days at grade 4 before the first reduction to 5.8 after the second;
set against the fixed arm, which sits near 8.8 days a cycle
throughout, that is a couple of days a cycle rather than a rescue.

Either way, the grey and blue curves are the same model, the same
parameters and the same 126 days.  The only difference is that in the
blue one the event history was written by the solver.

## Notes from actually using it

A few things that will save you an afternoon:

- **Decision times must exist.**  The logic runs at output times.  If
  the assessment is not in `et()`, pin it with `mtime()`.

- **Guard against re-triggering.**  `anc >= 1.5` is true for weeks.
  Anchoring on `t %% 168 == 0` is what turns a standing condition into
  a single decision.

- **Bookkeeping belongs in sticky variables, not states.**  Counters,
  dose levels and "when is the next visit" carry over from one time
  point to the next on their own; `is.na()` is how you initialise
  them.  A zero-derivative compartment does the same job, but the
  solver then has to carry it.

- **`amt` wants a symbol.**  Compute `doseMg <- mgm2*bsa` on its own
  line; `infuseDur(mgm2*bsa, ...)` will not parse.

- **Set `maxExtra`.**  It is a cheap way to find out that your
  condition fires 4000 times instead of six.

- **A pushed event lands in the row after the one that pushed it; a
  sticky assignment does not.**  Reading state back out, that
  difference decides which way you look: `c(0, diff(x))` for a sticky
  variable, `c(diff(x), 0)` for a compartment set by `replace()`.  Get
  it backwards and every value you report is one output step stale --
  here that would turn the lowest ANC at a dose into 1.56 instead of
  1.72.

- **At the last output time the event is dropped but the sticky
  assignment is not.**  There is no row left to put a pushed event in,
  so the dose never happens -- but `nextDue` and `dayLow` are ordinary
  assignments and go through anyway, leaving the model's own record
  claiming a dose it never gave.  That is what the `t < 126*24` guard
  in the protocol is for: it changes no trajectory at all (drop it and
  the solution is bit-identical), it stops the read-out reporting 298
  doses instead of 249.  Note the asymmetry is new with sticky
  variables -- when the bookkeeping was a `replace()` event it was
  dropped along with the dose, and the counts came out right by
  accident.

- **Reading decisions back out of the state vector only works if the
  decision times are on the grid.**  `filter(step == 504)` is exact
  equality on a float; it survives here because 168 and 504 land
  exactly on a 6 h grid.  Solve on `by = 5` and the summary silently
  reports zero doses rather than erroring.

- **`linCmt()` works too**, and `odeToLin()` will convert a *linear*
  ODE model that uses adaptive dosing into the analytic form -- worth
  it when you are running that kind of model over many trial designs.
  It cannot help the example above, whose PD is nonlinear.

## Where to read more

The [adaptive dosing
vignette](https://nlmixr2.github.io/rxode2/articles/adaptive-dosing.html)
is the reference: every helper, the `evid_()` low-level interface,
phantom dosing for transit models, adaptive sampling with `obs()`, and
the `linCmt()` and `odeToLin()` story.

The obvious next step is that these protocol rules are now part of the
model, which means they can be part of a trial simulation -- compare
two dose-modification tables, or ask what a different assessment
schedule would have bought you, without leaving `rxode2`.

## References

- Friberg LE, Henningsson A, Maas H, Nguyen L, Karlsson MO. [Model of
  chemotherapy-induced myelosuppression with parameter consistency
  across drugs](https://ascopubs.org/doi/10.1200/JCO.2002.02.140).
  *J Clin Oncol*. 2002;20(24):4713-4721.
- Ozawa K, Minami H, Sato H. [Population pharmacokinetic and
  pharmacodynamic analysis for time courses of docetaxel-induced
  neutropenia in Japanese cancer
  patients](https://pubmed.ncbi.nlm.nih.gov/17888031/).
  *Cancer Sci*. 2007;98(12):1985-1992.
- [TAXOTERE (docetaxel) prescribing
  information](https://products.sanofi.us/taxotere/taxotere.pdf),
  Sanofi-Aventis.
