---
title: nlmixr2/rxode2 exploring data with rxode2 geoms
author: Matthew Fidler
date: '2025-05-31'
slug: []
categories: [rxode2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nlmixr2)
library(rxode2)
# suppress messages
rxode2:::rxSetSilentErr(1L)
```

## `rxode2` and `ggplot`

`rxode2` (and `nlmixr2`) uses ggplot internally.  This means most
things are compatible with `ggplot2`.

One thing that is not quite as widely known that `rxode2` has some
custom `geom` functions that are useful for exploring pharmacometrics
data.

### `geom_amt()` -- exploring **when** dosing occurs

`rxode2` will allow exploring time of dosing with the `geom_amt()`.
From the `geom_amt()` documentation we can see how this is applied:


```{r geom_amt}

library(rxode2)
library(units)
library(ggplot2)

mod1 <- function() {
  ini({
    KA <- 2.94E-01
    CL <- 1.86E+01
    V2 <- 4.02E+01
    Q  <- 1.05E+01
    V3 <- 2.97E+02
    Kin <- 1
    Kout <- 1
    EC50 <- 200
  })
  model({
    C2 <- centr/V2
    C3 <- peri/V3
    d/dt(depot) <- -KA*depot
    d/dt(centr) <-  KA*depot - CL*C2 - Q*C2 + Q*C3
    d/dt(peri)  <-                    Q*C2 - Q*C3
    d/dt(eff)   <- Kin - Kout*(1-C2/(EC50+C2))*eff
  })
}

## These are making the more complex regimens of the rxode2 tutorial

## bid for 5 days
bid <- et(timeUnits="hr") %>%
  et(amt=10000,ii=12,until=set_units(5, "days"))

## qd for 5 days
qd <- et(timeUnits="hr") %>%
  et(amt=20000,ii=24,until=set_units(5, "days"))

## bid for 5 days followed by qd for 5 days

et <- seq(bid,qd) %>% et(seq(0,11*24,length.out=100))

bidQd <- rxSolve(mod1, et, addDosing=TRUE)

# by default dotted and under-stated
plot(bidQd, C2) + geom_amt(aes(amt=amt))

# of course you can make it a bit more visible

plot(bidQd, C2) + geom_amt(aes(amt=amt), col="red", lty=1, linewidth=1.2)
```


While not terribly complicated it is a convenient way to add dosing to
your `ggplot2` object.


### `geom_cens()` -- exploring **what** the censoring means


Another useful way to see how your data is censored (according to
`nlmixr2` and Monolix) is to use the `geom_cens()` function:

```{r geom_cens}

library(rxode2)
library(ggplot2)
library(tidyverse)
library(xgxr)

# Get the data from xgxr
pkpd_data <-
  case1_pkpd %>%
  arrange(DOSE) %>%
  select(-IPRED) %>%
  mutate(TRTACT_low2high = factor(TRTACT, levels = unique(TRTACT)),
         TRTACT_high2low = factor(TRTACT, levels = rev(unique(TRTACT))),
         DAY_label = paste("Day", PROFDAY),
         DAY_label = ifelse(DAY_label == "Day 0","Baseline",DAY_label))

pk_data <- pkpd_data %>%
  filter(CMT == 2)

pk_data_cycle1 <- pk_data %>%
  filter(CYCLE == 1)


ggplot(data = pk_data_cycle1, aes(x = TIME, y = LIDV)) +
  geom_line(aes(group = ID), color = "grey50", linewidth = 1, alpha = 0.3) +
  geom_cens(aes(cens=CENS)) +
  xgx_geom_ci(aes(x = NOMTIME, color = NULL, group = NULL, shape = NULL), conf_level = 0.95) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units(units_dataset = "hours", units_plot = "days") +
  labs(y = "Concentration (ng/ml)", color = "Dose") +
  theme(legend.position = "none") +
  facet_grid(.~TRTACT_low2high)

```

Whenever a point is censored, it shows the censoring assumptions by
drawing a small box where the censoring occurs.

This is adapted from the nlmixr2 [xgxr and ggPMX vignette](https://nlmixr2.org/articles/xgxr-nlmixr-ggpmx.html).


## Other notes

This showed 2 different extra geoms you can use with your own
analysis!

You can also use the `rxTheme()` ggplot theme with other
ggplot objects (or apply your own theme to the ggplot objects
generated from `rxode2`).

I chose to highlight these extra `geom`s from `rxode2` because I don't
believe that these are currently very well known and can be
useful depending on your analysis.
