---
title: nlmixr2 inter occasion varibility
author: Matthew Fidler
date: '2026-01-16'
slug: []
categories: [nlmixr2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Inter-occasion variability

Happy new year!  With the holidays over we had the opportunity to
celebrate many special occasions.

I invite you to celebrate a new occasion for nlmixr2 -- inter-occasion
variability.

We now support inter-occasion variability estimation in `nlmixr2`!

### Implementing inter-occasion variability in models

To implement inter-occasion variability in `nlmixr2` models you simply
have to specify that the variable is an occasion variable, say `occ`:

```r
iov.cl ~ 0.1 | occ
```

The `~` syntax is similar to how between subject variability is
specified, the addition of `|` tells that this a between `occ`
variability.

A full model based on the theophylline example is given below:

```{r iov}

library(nlmixr2)

one.cmt <- function() {
  ini({
    tka <- 0.45 # Log Ka
    tcl <- log(c(0, 2.7, 100)) # Log Cl
    tv <- 3.45; label("log V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    iov.cl ~ 0.1 | occ
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl + iov.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}
```

Note that the inter-occasion variability is specified for clearance:

```r
cl <- exp(tcl + eta.cl + iov.cl)
```

Trying this out with the theophylline multiple dose dataset with an
additional occasion added:


```{r iovRun}

theo_iov <- nlmixr2data::theo_md
theo_iov$occ <- 1
theo_iov$occ[theo_iov$TIME >= 144] <- 2

fit <- suppressMessages(nlmixr(one.cmt, theo_iov, est="focei", control=list(print=0)))
print(fit)
```

Note that IOV is specified in the final output table in the same
column as the between subject variability:

```{r fitParFixed}
fit$parFixed
```

Note that some of the outputs are in a different format, for example
`$omega` gives both the between subject variability and between
occasion variability:


```{r iovVar}
fit$omega
```

Though, the `eta` table, and corresponding standard errors and related
values are missing IOV:

```{r eta}
fit$eta
```

The IOV is listed by the categories below:

```{r iovPrint}
fit$iov
```


### Implementation details

The implementation is based on pre-processing and post-processing the
models before running them.  This is done by "hooks" into the nlmixr2
estimation engine (you can see the code
[here](https://github.com/nlmixr2/nlmixr2est/blob/main/R/iov.R)) which
can be used to extend `nlmixr2` even more).

The pre-processing hooks are equivalent to the code [here](https://github.com/nlmixrdevelopment/nlmixr/issues/93#issuecomment-2093486042).

This process:

 - Creates a fixed between subject variability for each occasion/parameter in the model.

 - Estimates the standard deviation (in some way) of the between occasion varibility

 - Back-translates this to the variance

 As such, covariance/correlation in inter-occasion terms are not supported yet.

### Changing IOV items in the next release of nlmixr2est

The current release filters out the statistics based on each
inter-occasion varaibility between subject variaibility for many
terms.  However, the next release will also keep shrinkage
calculations for this and include an overall shrinkage calculation.

#### Note

The inter-occasion variability logo was created by Canva using AI.
