---
title: rxode2 2.0.9/2.0.10
author: Matthew Fidler
date: '2022-10-19'
slug: []
categories: [rxode2]
tags: [new-version]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rxode2)
```

`rxode2` 2.0.9 has been released, and `rxode2` 2.0.10 will be released
soon! I want to personally thank all those who have submitted issues,
and helped with the development.  Without the support `rxode2`
wouldn't be the tool it is today.

This is the first CRAN-visible release since `rxode2` 2.0.7 and I
would like to highlight a few new interesting features:

## 'rxode2' can now have more flexible model functions

The key features are:

- You do not need an `ini()` block any longer
- You do not need to specify an endpoint either with `~`

For example, this model is perfectly reasonable in `rxode2`:

```{r modelOnly}

set.seed(42)
rxSetSeed(42) # for parallel random number generator

one.compartment <- function() {
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    d / dt(depot) <- -ka * depot
    d / dt(center) <- ka * depot - cl / v * center
    F(depot) <- 3
    cp <- center / v
  })
}

m <- one.compartment()

m

# When solving you will have to manually supply the parameters

theta <- c(tka = 0.45, tcl = 1,tv = 3.45)

omega <- lotri({eta.ka ~ 0.6
  eta.cl ~ 0.3
  eta.v ~ 0.1
})

# Create an event table
et <- et(amt=300) %>%
  et(0,24, by=2) %>%
  et(id=1:12)

# simulate directly from the model
s <- rxSolve(m, et, theta, omega=omega)

library(ggplot2 )

plot(s, cp) + ylab("Concentration")
```

## You can now pipe omega matrices to set initial values

For example, using the above model you could the omega initial values:

```{r modelOmega}
m2 <- m %>%
  ini(omega)


print(m2)

# notice that the model now includes an `ini({})` block

# If you pipe the theta values, you can get a full model:

m2 <- m2 %>% ini(tka = 0.45, tcl = 1,tv = 3.45)

print(m2)

# Which of course still allows simple solving:
s <- rxSolve(m2, et)


plot(s, cp) + ylab("Concentration (model m2)")
```
## Piping classic `rxode2` models

With the above flexibility, this release also allows piping of classic
`rxode2` models:

For example:

```{r classicModel}
rx <-  rxode2({
  ka <- exp(tka + eta.ka)
  cl <- exp(tcl + eta.cl)
  v <- exp(tv + eta.v)
  d / dt(depot) <- -ka * depot
  d / dt(center) <- ka * depot - cl / v * center
  F(depot) <- 3
  cp <- center / v
})

print(rx)


m3 <- as.function(rx) %>%
  ini(omega) %>%
  ini(tka = 0.45, tcl = 1,tv = 3.45)

print(m3)
```

Note the use of `as.function()` here, though it may not always be required in the future.

## Why are there more dependencies for `rxode2`?

CRAN had requested that we reduce the compile time for 'rxode2' to
remain on CRAN.  This effectively was requesting the compile be
split out to separate packages.

Note, now the packages that `rxode2` depend on are:

- `rxode2parse` which is the parsing of the rxode2 low level language
  to C (and includes some solved linear compartment code at the
  moment)
- `rxode2random` which has the parallel safe random number generation
  routines, and some other random number generating functions like
  `cvPost()`
- `rxode2ll` which includes the new likelihood functions supported in
  this release (which will add generalized likelihood estimation in
  `nlmixr2`)
- `rxode2et` which the `rxode2` event table
  function `et()`, split off
  
# What about when CRAN is out of sync (or temporarily dropped the packages)

Because the dependencies of 'nlmixr2' and 'rxode2' are currently a
binary dependency of each other, if they were not compiled together
they will not work together.  You may get issues like "This was
compiled against a different version of PACKAGE" when trying to load
the package.

One approach is to use the r-universe to install the development
version everything, that is:

``` r
install.packages(c("dparser", "rxode2ll", "rxode2parse",
                   "rxode2random", "rxode2et", "symengine", "rxode2",
                   "nlmixr2est", "nlmixr2extra",  "nlmixr2plot",
                   "nlmixr2"),
                 repos=c(nlmixr2="https://nlmixr2.r-universe.dev",
                         symengine="https://symengine.r-universe.dev",
                         CRAN="https://cloud.r-project.org"))

```

I would only suggest this if CRAN doesn't seem to be working for your
setup.


## Full changes from 2.0.7

# rxode2 2.0.10

- Time invariant covariates can now contain 'NA' values.

- When a column has 'NA' for the entire id, now 'rxode2' warns about
  both the id and column instead of just the id.
  
- To fix some CRAN issues in 'nlmixr2est', make the version dependency
  explicit.

# `rxode2` 2.0.9

- Remove log likelihoods from `rxode2` to reduce compilation time and
  increase maintainability of `rxode2`. They were transferred to
  'rxode2ll' (requested by CRAN).
  
- Remove the parsing from `rxode2` and solved linear compartment code
  and move to `rxode2parse` to reduce the compilation time (as requested
  by CRAN).
  
- Remove the random number generation from `rxode2` and move to
  `rxode2random` to reduce the compilation time (as requested by
  CRAN).

- Remove the event table translation and generation from `rxode2` and
  move to `rxode2et` to reduce the compilation time (as requested by
  CRAN).
  
- Change the `rxode2` ui object so it is a compressed, serialized
  object by default.  This could reduce the `C stack size` problem
  that occurs with too many environments in R.
  
- Warn when ignoring items during simulations

- Export a method to change `rxode2` solve methods into internal integers

- Bug fix for time invariant covariates identified as time variant
  covariate when the individual's time starts after `0`.

# `rxode2` 2.0.8

## Breaking changes

- `rxgamma` now only allows a `rate` input.  This aligns with the
  internal `rxode2` version of `rxgamma` and clarifies how this will
  be used. It is also aligned with the `llikGamma` function used for
  generalized likelihood estimation.

- ui `cauchy` simulations now follow the ui for `normal` and `t`
  distributions, which means you can combine with transformations.
  This is because the `cauchy` is a `t` distribution with one degree
  of freedom.

- ui `dnorm()` and `norm()` are no longer equivalent to `add()`.  Now
  it allows you to use the loglik `llikNorm()` instead of the standard
  `nlmixr2` style focei likelihood.  This is done by adding `dnorm()`
  at the end of the line.  It also means `dnorm()` now doesn't take
  any arguments.
  
- Vandercorput normal removed (non-random number generator)

## New features

- Allow models in the `nlmixr2` form without an `ini({})` block

- Allow model piping of an omega matrix by `f %>% ini(omegaMatrix)`

- Standard models created with `rxode2()` can no be piped into a model function

- Families of log-likelihood were added to `rxode2` so that mixed
  likelihood nonlinear mixed effects models may be specified and run.

- The memory footprint of a `rxode2` solving has been reduced

- Piping now allow named strings (issue #249)

## Bug fixes

- `rxode2`'s symengine would convert `sqrt(2)` to `M_SQRT_2` when it
  should be `M_SQRT2`.  This has been fixed; it was most noticeable in
  nlmixr2 log-likelihood estimation methods
  
- `rxode2` treats `DV` as a non-covariate with `etTran` (last time it
  would duplicate if it is in the model).  This is most noticeable in
  the nlmixr2 log-likelihood estimation methods.
  
## New features

- A new flag (`rxFlag`) has been created to tell you where in the
  `rxode2` solving process you are.  This is useful for debugging. If
  outputting this variable it will always be `11` or calculating the
  left handed equations.  If you are using in conjunction with the
  `printf()` methods, it is a double variable and should be formatted
  with `"%f"`.
  
- An additional option of `fullPrint` has been added to `rxode2()`
  which allows `rprintf()` to be used in almost all of `rxode2()`
  steps (inductive linearization and matrix exponential are the
  exception here) instead of just the integration `ddt` step.  It
  defaults to `FALSE`.
