---
title: rxode2 plotting
author: Matthew Fidler
date: '2025-02-25'
slug: []
categories: [rxode2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nlmixr2)
library(rxode2)
```

This month I will focus on `rxode2`'s plotting functions. I think these
are very useful but not very well known. In general I will focus on
three things: plotting single subjects, plotting multiple subjects and
plotting confidence bands.

## Plot functions

### `rxode2` plotting

Of course discussion of plotting require a model and simulation, so we
will use the model from the original tutorial:

```{r rxModel}
library(rxode2)

## Model from rxode2 tutorial
m1 <- rxode2({
  KA <- 2.94E-01
  CL <- 1.86E+01
  V2 <- 4.02E+01
  Q <- 1.05E+01
  V3 <- 2.97E+02
  Kin <- 1
  Kout <- 1
  EC50 <- 200
  ## Added modeled bioavaiblity, duration and rate
  fdepot <- 1
  durDepot <- 8
  rateDepot <- 1250
  C2 <- centr / V2
  C3 <- peri / V3
  d/dt(depot) <- -KA * depot
  f(depot) <- fdepot
  dur(depot) <- durDepot
  rate(depot) <- rateDepot
  d/dt(centr) <- KA * depot - CL * C2 - Q * C2 + Q * C3
  d/dt(peri) <- Q * C2 - Q * C3
  d/dt(eff) <- Kin - Kout * (1 - C2 / (EC50 + C2)) * eff
  eff(0) <- 1
})
```

#### rxode2 plotting of single subject solved objects

The first part is plotting a simulation based on a single subject event
table:

```{r plot1}
ev <- et(timeUnits = "hr") %>%
  et(amt = 10000, ii = 12, until = 24) %>%
  et(seq(0, 24, length.out = 100))

s <- rxSolve(m1, ev)

plot(s)
```

One thing that is very easy to see is that the default plots all the
calculated states and parameters in the model.

Often you may not want to see all the plots.  For example, if you are
more interested in the `C2` concentration, you can subset to `C2` by
simply specifying the compartment after the solved object:

```{r plotC2}
plot(s, C2)
```

You can specify as many compartments or calculated values as you wish
with this method, for example if you are interested in both `C2` and
`eff` we would have:

```{r plotC2Eff}
plot(s, C2, eff)
```

This is one of the most basic types of things that plotting does for
you.  As a note, the type of units affects the x-label axis if you
have `xgxr` installed (for best plotting results this extra tool is
helpful).

```{r plotC2EffNoXgx}
withr::with_options(list(rxode2.xgxr = FALSE), {
  plot(s, C2, eff)
})
```

Now the unit ticks are in terms of 5 instead of 3 (which means 24
hours is not even displayed on the plot) by using
[`xgxr::xgx_scale_x_time_units()`](https://opensource.nibr.com/xgxr/reference/xgx_scale_x_time_units.html).

You can also change the axes to be semi-logarithmic with the
`log="x"`, `log="y"` or `log="xy"`, as documented in the `plot()`
function:

```{r plotLog}
plot(s, C2,  log = "x")

plot(s, C2,  log = "y")

plot(s, C2,  log = "xy")
```

This is done with [`xgxr::xgx_scale_x_log10()`](https://opensource.nibr.com/xgxr/reference/xgx_scale_x_log10.html) and [`xgxr::xgx_scale_y_log10()`](https://opensource.nibr.com/xgxr/reference/xgx_scale_y_log10.html)

Currently, the other documented features of `plot()` are unimplemented
(but may be implemented in the future).

One point is that the return of `plot` is a `ggplot2` object, that
means you can also use `ggplot2` types of operations on the object:

```{r plotGg}
library(ggplot2)
plot(s, C2) + xlim(6, 12) + theme_bw()
```

#### `rxode2` plotting of multiple subjects

`rxode2` plots multiple subject simulations a bit differently. I will
use model piping to add a log-normal between subject variability on
clearance, and then plot with a few subjects

```{r multiSubjects}
m2 <- m1 %>%
  model( CL <- 1.86E+01 * exp(eta.Cl)) %>%
  ini(eta.Cl ~ 0.4^2)

ev <- et(amount.units = "mg", time.units = "hours") %>%
  et(amt = 10000, cmt = "centr", until = 48, ii = 8) %>%
  et(0, 36, length.out = 100)

s <- rxSolve(m2, ev,  nSub = 3)

```

With only 3 subjects, the subject id is labeled with
[`ggrepel::geom_label_repel()`](https://ggrepel.slowkow.com/reference/geom_text_repel) if `ggreple` is installed in your system:

```{r plot3}
plot(s, C2, log="y")
```

If you get a large number of subjects they become gray spaghetti lines instead:

```{r plot10}
s <- rxSolve(m2, ev, nSub = 10)

plot(s, C2, log="y")
```

The break-point can be changed with the option `rxode2.spaghetti`
which defaults to 7 subjects.

```{r plot10.2}
options(rxode2.spaghetti=12)
plot(s, C2, log="y")
```

#### `rxode2` summarizing and plotting of multiple subjects confidence intervals

One of the things I do often is to create quantile bands from my
simulation.

Continuing from a simple simulation, we have:

```{r ci500}
s <- rxSolve(m2, ev, nSub=500)

ci <- confint(s, parm=c("eff", "C2"), 0.95)

# The numeric quantiles from `confint()`
print(ci)

plot(ci)

ci <- confint(s, parm=c("eff", "C2"), 0.95)
```

As mentioned in the notes, with enough simulations you can get a
confidence band around the interval.

```{r ci2}
s <- rxSolve(m2, ev, nSub=2500)

ci <- confint(s, parm=c("eff", "C2"), 0.95)

# The numeric confidence intervals:
print(ci)

plot(ci, log="y")
```

#### `rxode2` stratified quantile intervals

One of the most common tasks that I perform is stratify my simulation by certain sub-groups.

This can be done easily by a `rxode2` extension of `confint` to
stratify with the `by` argument.  An example of this is:

```{r ci2by}
m2 <- m1 %>%
  model( CL <- 1.86E+01 * exp(eta.Cl) + 3*(sex=="Female")) %>%
  ini(eta.Cl ~ 0.4^2)

ev2 <- ev %>% et(id=1:50) %>%
  as.data.frame(.) %>%
  dplyr::mutate(sex="Male")

ev3 <- ev %>% et(id=51:100) %>%
  as.data.frame(.) %>%
  dplyr::mutate(sex="Female")

evSex <- rbind(ev2, ev3)

s <- rxSolve(m2, evSex, nStud=500)

# NOTE: by is not a documented in `confint` but can be used
# by a rxode2 solved object
ci <- confint(s, parm=c("eff", "C2"), 0.95, by="sex")

# The numeric confidence intervals:
print(ci)

plot(ci)
```

This brings up a couple of points for discussion/understanding.

First, the defaults for `confint` are a bit different for `rxode2`
solved objects, this is clearly seen by using the `by` variable.

Other options other than `by` are:

- `ci`, in conjunction with `level`, controls
  levels of the uncertainty in the quantiles or confidence intervals calculated
- `mean` -- should the mean values be calculated instead of the
  quantiles. When this is true instead of the `base::quantile()`
  function, summaries are done with
  [`rxode2::meanProbs()`](https://nlmixr2.github.io/rxode2/reference/meanProbs.html),
  and all the options there are accepted (`useT`, `pred`, and `n`).
- `binom` -- should binomial values be calculated instead of the binom
  values.  When this is true, instead of using the `base::quantile()`
  function, summaries are done with
  [`rxode2::binomProbs()`](https://nlmixr2.github.io/rxode2/reference/binomProbs.html)
  and all the relevant options there are accepted (`n`, `m`,
  `ciMethod`, `pred`, etc)

For this blog post I am only focusing on quantiles (perhaps in a later
blog post I will cover mean/binom summaries).

### `ci` in `confint`

By default, the `confint` is actually the quantiles controlled by
`level` and `ci` options, I show extreme options of low and high
`ci`/`levels` as well as `ci=FALSE` to show what each option does.

First we setup the model and the data:

```{r ciLevelSetup}
m2 <- m1 %>%
  model( CL <- 1.86E+01 * exp(eta.Cl)) %>%
  ini(eta.Cl ~ 0.4^2)

s <- rxSolve(m2, ev, nSub=2500)

```

You can see wider confidence bands in the next plot because the overall levels
is `0.995`, but very little variability in the uncertainty of these
bands (since the `ci=0.25`):

```{r ci995}
c <- confint(s, parm=c("C2", "eff"), level=.995, ci=.25)

plot(c)

```

On the other hand, with a `level=0.25` the overall confidence bands
are smaller while the confidence around the lines are larger (`ci=0.995`)

```{r ciLevel}
c <- confint(s, parm=c("C2", "eff"), level=.25, ci=.995)
plot(c)
```

Note that if you specify only `level`, the `ci` matches the `level`
argument:

```{r oneLevel}
c <- confint(s, parm=c("C2", "eff"), level=.995)
plot(c)
```

You can also force the simualtion only one quantile calculation by
turning off `ci`:

```{r noCi}
c <- confint(s, parm=c("C2", "eff"), level=.995, ci=FALSE)

plot(c)
```

For these confidence bands you can see that the overall `level`
represents the overall percentiles, while `ci` represents the
smaller percentiles (which should be illustrated by the above plots).

The for each theoretical study, the quantiles are taken according to
the `level`.  Then these quantiles get confidence around them with
quantiles taken by the values in `ci`.

For single subject simulations, the number of subjects are split
approximately evenly.  For example with `2500` simulations, you would
split the number of studies (`nStud`) to be `round(sqrt(2500))`, or
`50` with `50` subjects apiece.  The first 50 subjects are in
`nStud=1`, second 50 subjects are in `nStud=2`, etc.

### Other plotting functions

Note there are some other plotting functions in `rxode2` like:


- [`rxode2::geom_amt()`](https://nlmixr2.github.io/rxode2/reference/stat_amt.html)
for showing dosing in `rxode2`/`NONMEM` types of datasets.

- [`rxode2::geom_cens()`](https://nlmixr2.github.io/rxode2/reference/stat_cens.html)
allows exploration of the censoring data style that is supported by
`nlmixr2`/`Monolix`.

These may be covered in the future blog and plots from `nlmixr2` would
likely be covered later.
