---
title: "nlmixr2 2.0.8 Objectively Surprising"
author: "Matt Fidler and the nlmixr2 Development Team"
date: "2022-10-25"
slug: []
bibliography: [refs.bib]
link-citations: true
csl: vancouver.csl
categories: [nlmixr2]
tags: [new-version]
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rxode2)
library(tidyverse)
library(nlmixr2)
```

Last time I blogged promised to talk about a few other things,
including:

-   Likelihood based on each observation (and how to get it)

-   Standard Errors / Hessians, etc for between subject variabilities or
    `eta`s (and how to get them)

Hessians for the individual between subject variability is also used for
the `focei` calculation. So, if you are impatient, I will give you brief
instructions on where to get each component of the likelihood:

-   The individual observation's likelihood contribution is contained in
    the datasets where the original (left) merged with the fit (right)
    in any of the following accessor methods: `fit$dataMergeLeft`,
    `fit$dataMergeRight`, `fit$dataMergeInner`, or `fit$dataMergeFull`.
    In these dataset an new column is added `$nlmixrLlikObs`

-   The individual -Hessian `etas` can be accessed by `fit$etaH` or
    `fit$phiH`

Other components may be accessed as well:

+------------------+---------------------------------------------------+
| Syntax           | Values returned                                   |
+==================+===================================================+
| `$phiH`, `$etaH` | -Hessian matrix                                   |
+------------------+---------------------------------------------------+
| `$phiC`, `$etaC` | Covariance matrix, standard deviations on         |
|                  | diagonals, correlations on off-diagonals          |
+------------------+---------------------------------------------------+
| `$phiR`, `$etaR` | Correlation matrix, standard deviations on        |
|                  | diagonals, correlations on off-diagonals          |
+------------------+---------------------------------------------------+
| `$phiSE`,        | Standard error of each individual's eta           |
| `$etaSE`         |                                                   |
+------------------+---------------------------------------------------+
| `$phiRSE`,       | |                                                 |
| `$etaRSE`        |  relative standard error of each individual's eta |
+------------------+---------------------------------------------------+

These all require that the `cwres` are in the fit because they come from
the `focei` calculations (and are also under the `focei` assumption).

# Objective function Motivating example

I was working with Bill Denney to prepare past course that features
`babelmixr2`. In this course, you can perform a `NCA` analysis (using
`PKNCA`), then use these values (and possibly calculate a unit
conversion) to create a initial `nlmixr2` PK model. This model has with
`NCA` derived initial estimates and ranges (and needed unit conversions
too)!

This is exciting to me, as someone who has been wanting this feature in
nonlinear mixed effects modeling packages like `nlmixr2` for quite
awhile.

## Two nearly identical models

Still, when testing this we came across the following (possibly)
surprising situation:

```{r theo1, echo=TRUE}
one.compartment <- function() {
  ini({
    tka <- 0.45 # Log Ka
    tcl <- 1 # Log Cl
    tv <- 3.45    # Log V
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    d/dt(depot) = -ka * depot
    d/dt(center) = ka * depot - cl / v * center
    cp = center / v
    cp ~ add(add.sd)
  })
}

fit1 <- nlmixr(one.compartment, nlmixr2data::theo_sd,  
               est="focei", control=list(print=0))
```

Now multiply the `cp` by 1000 and the observations by 1000 for a nearly
identical model (ie, change the scale for different units)

```{r theo2}
d2 <- nlmixr2data::theo_sd %>%
  mutate(DV=ifelse(AMT==0, DV*1000, DV))

# Use model piping to scale `cp`:
one.compartment %>%
  model(cp = 1000*center/v) %>%
  ini(add.sd=700)-> 
  m2

# Verify the new model
print(m2)

fit2 <- nlmixr(m2, d2, est="focei", control=list(print=0))
```

### Comparing Estimates

As expected the population estimates are similar:

```{r compare}
#fit1
print(fixef(fit1))

#fit2 
print(fixef(fit2))
```

Note that the additive error is (unsurprisingly) larger by a factor of
about 1000.

Still, the Omega matrices are similar too:

```{r compareOmega}
# fit 1
print(fit1$omega)
# fit 2 
print(fit2$omega)
```

And the etas:

```{r compareEtas}
# fit 1
print(fit1$etaObf)
# fit 2
print(fit2$etaObf)
```

The ETAs are similar too; You can also see the individual contribution
to the objective functions are quite different (`OBJI`). So it should be
no surprise that the objective functions are different:

```{r compareObjf}
# fit 1
print(fit1$objf)
# fit 2
print(fit2$objf)
```

## What about NONMEM?

You might say, well are these objective functions off? maybe `nlmixr2`
is broken? (If you see anything surprising of course submit a bug report
if you can).

Well, with the coming `babelmixr2` you can run the same models in NONMEM
(with certain caveats we will discuss later), and these objective
functions also are similar NONMEM between `nlmixr2` and `NONMEM` (which
is unsurprising since we use the NONMEM objective functions in Wang 2007
[@Wang2007] to validate our likelihood)

This means that `nlmixr2` is constitent with `NONMEM` and these
objective function differences are due to other factors.

## Exploring more with individual observation contribution

One of the new features is the ability to see individual observations
contribution to the likelihood in `focei` related methods.

This can help us explore the differences.

In `nlmixr2`, you can use the `fit$dataMergeInner` to merge the original
data and the fit data. During this merge process it will also add the
column `$nlmixrLlikObs`:

```{r, obsLlik}
dm1 <- fit1$dataMergeInner

dm1ll <- dm1 %>%
  select(ID, nlmixrLlikObs) %>%
  group_by(ID) %>%
  summarize(sllik=sum(nlmixrLlikObs))

dm2 <- fit2$dataMergeInner

dm2ll <- dm2 %>%
  group_by(ID) %>%
  summarize(sllik=sum(nlmixrLlikObs))


print(dm1ll)

print(dm2ll)

# It is clear that there are individual differences in log-likelihood
```

In the normal (non generalized likelihood) the observation likelihoods
are given by $l_{i, obs}$:

$$l_{i, obs} = -0.5\times\left(\frac{\textsf{IPRED}-\textsf{DV}}{v}\right)^2-0.5*\log(v)$$

Where $v=$ variance of the estimate at that point. In this case it is
$\textsf{add.sd}^2$

You can see part of the difference is the relative differences of this
term for subjects. Most of this is likely driven by the large (and
unsurprising) differences in the variance component.

If you want, you can see which observations give the biggest difference
by comparing point by point.

## Finishing up the likelihood calculation

A part of the individual Hessians are the other component that is used
in the likelihood calculation. With the new tools you can also see what
this contribution to each individual's likelihood is:

``` r
hess1 <- merge(fit1$etaObf, dm1ll) %>%
  mutate(hessLlik=OBJI-sllik)

hess2 <- merge(fit2$etaObf, dm2ll) %>%
  mutate(hessLlik=OBJI-sllik)

# Hess1
print(hess1)
# Hess2
print(hess2)
```

You can see the individual Hessian contribution is actually large in
this particular likelihood as well. (You can explore their difference
more using `$etaH` if you wish)

# Conclusion

Well that is everything for now. This illustrates a few things:

-   How to get individual likelihoods

-   How to split apart the likelihood contribution from the Normal
    assumption of the observations and the contribution from the
    hessian. (Note this works with generalized likelihood too)

-   Where to get standard errors of `etas`

I wish I had known where these came from earlier, but I seem to want to
know how things work. For a more in-depth reference you could use the
paper by Almquist [@Almquist2015] to dig into the full likelihood math.

### References
