---
title: nlmixr2 3.0
author: Matthew Fidler
date: '2024-09-18'
slug: []
categories: [rxode2, nlmixr2, babelmixr2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rxode2)
library(nlmixr2)
```
# nlmixr2 3.0 is here

`nlmixr2` 3.0 has been released and most of the related packages in
the `nlmixr2` ecosystem have been updated as well. Since there were a
few non-backward compatible changes and breaking changes, the version
has been bumped from 2 to 3. Most code will run the same, but because
of the breaking change, we changed the major version.

The big changes are:

- Non abi binary linkages between every package.  This means you will not
  get messages about needing to re-compile the packages during CRAN
  updates (and it means much more stable packages and the ability to
  make more frequent updates). To me, this is as important an
  achievement to maintainability as removing python's `sympy` from
  `nlmixr`.

- New syntax for lower triangular matrices.  I will likely discuss why
  in a later blog post, but for now I will show the differences:

```{r lotri}
# The old syntax (which is still supported)
lotri({
  a+b ~ c(1,
               0.1, 1)
  c ~ 1
})

# the new syntax, lower triangular matrices can be suplied row by row;
# if restarting it would restart the block matrix
lotri({
  a ~ 1
  b ~ c(0.1, 1)
  c ~ 1
})
```

- `rxode2`/`nlmixr2` will re-order matrices as needed to make them in
  a banded-matrix format preferred by the estimation tool.  This is
  done when parsing the matrices but can be seen here:

```{r rcm}
m <- lotri({
  a + b + c + d + e + f + g + h + i + j + k + l + m + n + o +
  p ~ c(0.4, 0, 0.3, 0, 0, 0, -0.1, 0, 0, 0.2, 0, 0, 0,
        0, 0.5, 0, 0, 0, 0, 0, 1.3, 0, 0, 0, 0, 0, -0.6, 0.8,
        0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0.2,
        0, 0, 0, 0, 0, 0, 0, 0, 0, 0.9, 0, 0, 0, 0, 0, 0, 0,
        0, 0, 0, 0.9, 0, 0, 0, 0, 0, 0, 0, 0, 0, -0.2, 0, 0.3,
        0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 2.1, 0.2, 0, 0, 0.2,
        0, 0, 0, 0, 0, 0, 0, 0, 0, 0.4, 0, 0, 0, 0, 0, -1.1,
        0.9, 0, 0, 0, 0, 0, 0, 0, 4.7, 0, 0, 0, 0, 0, 0, 0, 0,
        0, 0.5, 0, 0.2, 0, 0, 0, 1.9)
})

# note that m is in a non-banded matrix format
m

# Internally the lotri function rcm (standing for the reverse Cuthill
# McKee (RCM) algorithm algorithm) creates the banded matrix
m <- lotri::rcm(m)

# If you specified this as an ini block, you would then see the
# following code instead:
lotri::lotriAsExpression(m, useIni=TRUE)

# which is both much easier to read and also helps the nlmixr2
# estimation routines.

# If you prefer the old syntax on output you can use
# options(lotri.plusNames=TRUE)
withr::with_options(list(lotri.plusNames=TRUE),{
  lotri::lotriAsExpression(m, useIni=TRUE)
})
```

- In `rxode2` simulations, `iCov` has been refactored to run faster
  than it did before.

- In `rxode2` you can now assign a variable to a string (`a <- "string"`)

- Now you can specify per-variable interpolation (`linear(WT)`, or
  `nocb(WT)` etc) inside your model.

- There is a new type of handling of `mu`-referenced equations, called
  `mu`3. (We already have
  [`mu`2](https://blog.nlmixr2.org/blog/2024-07-08-mu2/)). `saem` will
  now message if it is using `mu`2 or `mu`3 referencing while
  modeling. Standard mu referencing is not called out.

 - There is a new log-likelihood profiling functions exported for use
   with you `nlmixr2` fits.

I have a [technical discussion on why/how things have changed](#api),
or [jump to the overall changes](#Changelog).  The next section
discusses why our dependencies have changed.

## Why has the number of dependencies changed?

In
[2022](https://blog.nlmixr2.org/blog/2022-10-19-rxode2-2.0.9-release/),
`CRAN` requested we split `rxode2`, giving us more packages to
maintain. This was mostly because of compilation time, and the
`stan`-based, large-compilation time `rxode2ll` is still its own
separate package.

In the last release, they requested that we reduce the number of
packages for `rxode2` because it took too much of their time to
maintain the split they requested.

This is because the `rxode2`, `rxode2random`, `rxode2et` and
`rxode2parse` have [binary
inter-dependencies](https://cran.r-project.org/doc/manuals/R-exts.html#Linking-to-native-routines-in-other-packages-1)
between all of the packages that were split off.

This means that `rxode2`/`nlmixr2est` had to depend on exact versions
of the packages `rxode2random`, `rxode2parse`, and `rxode2et`.  If
this did not happen, you could crash R.  A sentence in their manual
talks about how fragile this is:

> **NB:** this mechanism is fragile, as changes to the interface
> provided by **packA** have to be recognised by **packB**. The
> consequences of not doing so have included serious corruption to the
> memory pool of the R session. Either **packB** has to depend on the
> exact version of packA or there needs to be a mechanism for
> **packB** to test at runtime the version of **packA** it is linked
> to matches that it was compiled against.

I can see how this was very painful to maintain on CRAN's end (and it
was difficult to maintain from my end as well).

I must say, that CRAN is a small group of volunteers who manage a
great number of very useful R packages.  They make sure
that the packages more or less interact well with each other, which is
actually a non-trivial task. This makes `CRAN` just work, when compared
to other repositories where this is not enforced.

So, in short, "CRAN gives, CRAN takes away, blessed be the name of
CRAN" (to mis-quote Job).

Note that I have taken other steps to make this easier to maintain on
my side and CRAN's side too.  For those who are interested I will talk
about them in below in the [`ABI/ABI` section](#api).

<div id="Changelog">

## Changelog

### nlmixr2

- Export likelihood profiling functions

### nlmixr2plot

- Changes to work with new `rxode2`

### nlmixr2extra

- New `profile()` method for likelihood profiling (Issue #1)

- Changes to work with new `rxode2`

### nlmixr2est

- No binary linking to `rxode2`, `lbfgsb3c` and `n1q1`, which means
  that updating these will not make `nlmixr2est` crash without
  recompiling.

- New `mu`3 referencing will take context from the model to see if the
  algebraic expression can be completed from defined model variables;
  These variable would have to be unique.

### rxode2

#### Breaking Changes

- The model properties was moved from `$params` to `$props` so it does
  not conflict with the low level `rxode2` model `$params`

- Error when specifying `wd` without `modName`

- With Linear and midpoint of a time between two points, how `rxode2`
  handles missing values has changed.  When the missing value is lower
  than the requested time, it will look backward until it finds the
  first non-missing value (or if all are missing start looking
  forward).  When the missing value is higher than the requested time,
  the algorithm will look forward until it finds the first non-missing
  value (or if all are missing, start looking backward).

- The order of ODEs is now only determined by the order of `cmt()` and
  `d/dt()`. Compartment properties, `tad()` and other compartment
  related variables no no longer affect compartment sorting.  The
  option `rxode2.syntax.require.ode.first` no longer does anything.

- The handling of zeros "safely" has changed (see #775)

  - when `safeZero=TRUE` and the denominator of a division expression
    is zero, use the Machine's small number/`eps` (you can see this
    value with `.Machine$double.eps`)

  - when `saveLog=TRUE` and the x in the `log(x)` is less than or
    equal to zero, change this to `log(eps)`

  - when `safePow=TRUE` and the expression `x^y` has a zero for `x`
    and a negative number for `y` replace `x` with `eps`.

  Since the protection for divide by zero has changed, the results
  will also change. This is a more conservative protection mechanism
  than was applied previously.

- Random numbers from `rxode2` are different when using `dop853`,
  `lsoda` or `indLin` methods.  These now seed the random numbers in
  the same way as `liblsoda`, so the random number provided will be
  the same with different solving methods.

- The arguments saved in the `rxSolve` for items like `thetaMat` will
  be the reduced matrices used in solving, not the full matrices (this
  will likely not break very many items)

#### Possible breaking changes (though unlikely)

- `iCov` is no longer merged to the event dataset.  This makes solving
  with `iCov` slightly faster (#743)


#### New features

- You can remove covariances for every omega by piping with `%>%
  ini(diag())` you can be a bit more granular by removing all
  covariances that have either `eta.ka` or `eta.cl` by: `%>%
  ini(diag(eta.ka, eta.cl))` or anything with correlations with
  `eta.cl` with `%>% ini(diag(eta.cl))`

- You can also remove individual covariances by `%>% ini(-cov(a, b))`
  or `%>% ini(-cor(a,b))`.

- You can specify the type of interpolation applied for added dosing
  records (or other added records) for columns that are kept with the
  `keep=` option in `rxSolve()`. This new option is
  `keepInterpolation` and can be `locf` for last observation carried
  forward, `nocb` which is the next observation carried backward, as
  well as `NA` which puts a `NA` in all imputed data rows. See #756.

   - Note: when interpolation is linear/midpoint for
     factors/characters it changes to locf with a warning (#759)

   - Also note, that the default keep interpolation is `na`

- Now you can specify the interpolation method per covariate in the model:

  - `linear(var1, var2)` says both `var1` and `var2` would use linear
    interpolation when they are a time-varying covariate. You could
    also use `linear(var1)`

  - `locf()` declares variables using last observation carried forward

  - `nocb()` declares variables using next observation carried backward

  - `midpoint()` declares variables using midpoint interpolation

- `linear()`, `locf()`, `locb()`, `midpoint()`, `params()`, `cmt()`
  and `dvid()` declarations are now ignored when loading a `rxode2`
  model with `rxS()`

- Strings can be assigned to variables in `rxode2`.

- Strings can now be enclosed with a single quote as well as a double
  quote.  This limitation was only in the rxode2 using string since
  the R-parser changes single quotes to double quotes. (This has no
  impact with `rxode2({})` and ui/function form).

- More robust string encoding for symengine (adapted from
  `utils::URLencode()` and `utils::URLdecode()`)

- Empty arguments to `rxRename()` give a warning (#688)

- Promoting from covariates to parameters with model piping (via `ini()`) now
  allows setting bounds (#692)

 - Added `assertCompartmentName()`, `assertCompartmentExists()`,
  `assertCompartmentNew()`, `testCompartmentExists()`,
  `assertVariableExists()` `testVariableExists()`,
  `assertVariableNew()`, `assertVariableName()`, and
  `assertParameterValue()` to verify that a value is a valid nlmixr2
  compartment name, nlmixr2 compartment/variable exists in the model,
  variable name, or parameter value (#726; #733)

- Added `assertRxUnbounded()`, `testRxUnbounded()`, `warnRxBounded()`
  to allow `nlmixr2` warn about methods that ignore boundaries #760

- Added functions `tad0()`, `tafd0()`, `tlast0()` and `tfirst0()` that
  will give `0` instead of `NA` when the dose has not been
  administered yet.  This is useful for use in ODEs since `NA`s will
  break the solving (so can be used a bit more robustly with models
  like Weibull absorption).

- `rxode2` is has no more binary link to `lotri`, which means that
  changes in the `lotri` package will not require `rxode2` to be
  recompiled (in most cases) and will not crash the system.

- `rxode2` also has no more binary linkage to `PreciseSums`

- The binary linkage for `dparser` is reduced to C structures only,
  making changes in dparser less likely to cause segmentation faults
  in `rxode2` if it wasn't recompiled.

- A new model property has been added to `$props$cmtProp` and
  `$statePropDf`.  Both are data-frames showing which compartment has
  properties (currently `ini`, `f`, `alag`, `rate` and `dur`)
  in the `rxode2` ui model.  This comes from the lower
  level model variable `$stateProp` which has this information
  encoded in integers for each state.

- A new generic method `rxUiDeparse` can be used to deparse meta
  information into more readable expressions; This currently by
  default supports lower triangular matrices by lotri, but can be
  extended to support other types of objects like 'nlmixr2's
  `foceiControl()` for instance.

#### Bug fixes

- Fix `ui$props$endpoint` when the ui endpoint is defined in terms of
  the ode instead of lhs. See #754

- Fix `ui$props` when the ui is a linear compartment model without `ka` defined.

- Model extraction `modelExtract()` will now extract model properties.  Note that the model property of `alag(cmt)` and `lag(cmt)` will give the same value. See #745

- When assigning reserved variables, the parser will error. See #744

- Linear interpolation will now adjust the times as well as the values
  when `NA` values are observed.

- Fix when keeping data has `NA` values that it will not crash R; Also
  fixed some incorrect `NA` interpolations. See #756

- When using `cmt()` sometimes the next statement would be corrupted
  in the normalized syntax (like for instance `locf`); This bug was
  fixed (#763)

- `keep` will now error when trying to keep items that are in the
  rxode2 output data-frame and will be calculated (#764)

#### Big change

- At the request of CRAN, combine `rxode2parse`, `rxode2random`, and
 `rxode2et` into this package; The changes in each of the packages are
 now placed here:

#### Fix for merged packages

- Fix a bug when simulating nested variables for random simulation
  (was in `rxode2random` #25)

- As requested by CRAN remove the `C` code `SET_TYPEOF` which is no
  longer part of the `C` `R` `API` for `rxode2parse`

<div id="api">

## ABI/API

I am writing this technical aside because I couldn't find a solution
for this problem on the internet or in the writing R extension manual
(my internet searching skills could probably be improved).  Also note
that if you are using `Rcpp` and exposing the `C++` calls, this is
done for you (but not for `C` calls).

### What is an ABI?

An `ABI` stands for application binary interface and is determined by
the compiler and the platform at the time of compile. Each function
has an address in memory.  This is why when some down-stream
dependency like `rxode2parse` are updated the functions that are
specified change memory addresses.  When `rxode2` tries to call a
function from the parser, R memory's is corrupted and R crashes. Of
course you can fix this by recompiling the `rxode2` version but
usually waiting for the binary that links was a better
solution.

Addresses of other items other than C functions can also change (like
C/C++ structures), so direct access of these items is also something
that can cause memory corruption.

This ABI interface is what happens when you register a `C` callable in
any package.

### The solution -- creating an API

An `API` is the Application Programming Interface.  It has been
popularized by web `API`s and can be created for your R package with
packages like `plumber`, however it can be any application programming
interface.

#### Step One -- determine and export the C structures and C functions to be used

To create the `API` I needed to do two things:

- Figure out what functions I would be using (and therefore allow
  others to use) from `rxode2`.

- Figure out what structures that I would need to access (like the ODE
  states, the calculated values of the equations etc). This will also
  allow other packages to use these functions in their code as well.

The first thing that will need to be done is to create thin accessing
functions for the underlying `C` structures.  Since we are using `R`
which is really based on `C`, they will need to be C compatible. The
[code
below](https://github.com/nlmixr2/rxode2/blob/096e100cabe187833b884e9c183083f5fa71cbf0/src/rx2api.c#L77-L79)
shows how to access the calculated variables inside `rxode2`:

``` c
double *getIndLhs(rx_solving_options_ind* ind) {
  return ind->lhs;
}
```

The next step is to create a function that grabs the memory addresses
of these thin C structure access functions as well as any other functions
that you will expose in the `API`.

You wrap all of these functions and place them into a R list structure
using R's function `R_MakeExternalPtrFn()`. An example function is
below and is based on the [actual `rxode2` API pointer
function](https://github.com/nlmixr2/rxode2/blob/096e100cabe187833b884e9c183083f5fa71cbf0/src/init.c#L358-L533):

``` c
SEXP _rxode2_rxode2Ptr(void) {
  int pro = 0;  // Counter for the number of PROTECT calls
  // Create an external pointer for _lotriLstToMat
  SEXP rxode2rxRmvnSEXP = PROTECT(R_MakeExternalPtrFn((DL_FUNC)&_rxode2_rxRmvnSEXP, R_NilValue, R_NilValue)); pro++;
  // more code here
  // Unprotect all protected objects
  UNPROTECT(pro);

  // Return the list of external pointers
  return ret;
}
```

After creating the `C` function, you register it as you a standard
function R calls from `C`.  In this case we have:

``` c
void R_init_rxode2(DllInfo *info){
  R_CallMethodDef callMethods[]  = {
    {"_rxode2_rxode2Ptr", (DL_FUNC) &_rxode2_rxode2Ptr, 0},
    // more registered functions
    {NULL, NULL, 0}
  };
  static const R_CMethodDef cMethods[] = {
    {NULL, NULL, 0, NULL}
  };

  R_registerRoutines(info, cMethods, callMethods, NULL, NULL);
  R_useDynamicSymbols(info, FALSE);
}
```

and of course you need to register the functions to be seen in R,
which needs the following line in the NAMESPACE:

```
useDynLib(rxode2, .registration=TRUE)
```

Which for `rxode2` comes from the roxygen2 line:

``` r
#' @useDynLib rxode2, .registration=TRUE
```

Once all of that is done, you can create an exported R function that
[lists the function pointers](https://github.com/nlmixr2/rxode2/blob/096e100cabe187833b884e9c183083f5fa71cbf0/R/rxode-options.R#L19C1-L32C2):

``` r
#' Get the rxode2 function pointers
#'
#' This function is used to get the function pointers for rxode2.  This is
#' used to allow rxode2 to have binary linkage to nlmixr2est.
#'
#' @return a list of function pointers
#' @export
#' @author Matthew L. Fidler
#' @examples
#' .rxode2ptrs()
.rxode2ptrs <- function() {
  .Call(`_rxode2_rxode2Ptr`, PACKAGE = "rxode2")
}
```

Which (of course) works for `rxode2`:

```{r ptr}
rxode2::.rxode2ptrs()[1:3]
```

#### Step 2 create the user interface header

For me, I want the user interface functions to be the same as what I
declared in the `API`, used in the second package (like `nlmixr2est`).

To do this, I create a header called
[`rxode2ptr.h`](https://github.com/nlmixr2/rxode2/blob/37105c33944589fb6a07d0376d9b1c41ea7b61ab/inst/include/rxode2ptr.h)
in the `inst/include` directory.

In that header, I declare the type definition of each of the functions
that will be accessed in the other package as well as an external
function pointer to the exported function:

``` c
typedef double *(*getIndLhs_t)(rx_solving_options_ind* ind);
extern getIndLhs_t getIndLhs;
```

Then [create a `static inline` function](https://github.com/nlmixr2/rxode2/blob/37105c33944589fb6a07d0376d9b1c41ea7b61ab/inst/include/rxode2ptr.h#L167-L218) (that is it is only defined
when called) that takes the pointers from the `rxode2::.rxode2ptrs()`
and assigns them the function pointer `getIndLhs`:

``` c
static inline SEXP iniRxodePtrs0(SEXP p) {
  if (_rxode2_rxRmvnSEXP_ == NULL) { // only assign if not already assigned
    // more code
    getIndLhs = (getIndLhs_t) R_ExternalPtrAddrFn(VECTOR_ELT(p, 23));
    // more code
  }
  return R_NilValue;
}
```

Note that this function is keyed to the number of API elements exposed
and assumes a specific order.  Therefore, you should add elements
sequentially and make sure `API` needs less elements than exported by
`rxode2` (which is coded in as an exception in `nlmixr2est`)

Finally we create code that defines the function pointers in a single
file and creates an R expression that will register all the C pointer
functions.

``` c
#define iniRxode2ptr                                    \
  _rxode2_rxRmvnSEXP_t _rxode2_rxRmvnSEXP_ = NULL;      \
  // more code
getIndLhs_t getIndLhs = NULL;                         \
SEXP iniRxodePtrs(SEXP ptr) {                         \
  return iniRxodePtrs0(ptr);                          \
}                                                     \
```

This is all that needs to be done for `rxode2` to expose the `API`

#### Step 3 using the API in `nlmixr2est`

For most `C` (or `C++`) files if the header has an accommodation for
C++, simply using:

``` c
#include <rxode2ptr.h>
```

Will register and expose the functions as needed.

In one file, you will also need to declare the function pointers and
function pointer with the line:

``` c
#include <rxode2ptr.h>
// if in C++ wrap with extern "C" {}

// if you want to change the name of the function you can use #define
// iniRxodePtrs _your_fn_name

iniRxode2ptr // do not include a ; since it will be flagged as a significant warning by CRAN
```

When registering the Calls inside of `nlmixr2est` you will need to add the ``

``` c
SEXP iniRxodePtrs(SEXP ptr);
/// some code

void R_init_nlmixr2est(DllInfo *info){
  R_CallMethodDef callMethods[]  = {
    {"iniRxodePtrs", (DL_FUNC) &iniRxodePtrs, 1},
    // more registered functions
    {NULL, NULL, 0}
  };
  static const R_CMethodDef cMethods[] = {
    {NULL, NULL, 0, NULL}
  };

  R_registerRoutines(info, cMethods, callMethods, NULL, NULL);
  R_useDynamicSymbols(info, FALSE);
}
```

Last you need to register the API when loading `nlmixr2est`:

``` r
# This will be saved when compiled
# This way you will know if something has changed in the API
rxode2.api <- names(rxode2::.rxode2ptrs())

.iniRxode2Ptr <- function() {
  .ptr <- rxode2::.rxode2ptrs()
  .nptr <- names(.ptr)
  # Cheks for API changes
  if (length(rxode2.api) > length(.nptr)) {
    stop("nlmixr2est requires a newer version of rxode2 api, cannot run nlmixr2est\ntry `install.packages(\"rxode2\")` to get a newer version of rxode2", call.=FALSE)
  } else {
    .nptr <- .nptr[seq_along(rxode2.api)]
    if (!identical(rxode2.api, .nptr)) {
      .bad <- TRUE
      stop("nlmixr2est needs a different version of rxode2 api, cannot run nlmixr2est\ntry `install.packages(\"rxode2\")` to get a newer version of rxode2, or update both packages", call.=FALSE)
    }
  }
  .Call(`iniRxodePtrs`, .ptr,
        PACKAGE = "nlmixr2est")
}

.onLoad <- function(libname, pkgname) {
  # other code
  .iniRxode2Ptr()
}
```

This same procedure can be used to expose all the `API` functions to
`R`.
