---
title: "babelmixr2, nlmixr2 and NONMEM"
author: "Matt Fidler and the nlmixr2 Development Team"
date: "2022-11-11"
slug: []
bibliography: [refs.bib]
link-citations: true
csl: vancouver.csl
categories: [babelmixr2]
tags: [new-version, NONMEM]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (dir.exists("content")) {
  setwd("content/blog/2022-11-11-babelmixr2-nonmem")
}
if (!dir.exists("pk.turnover.emax3-nonmem")) {
  stop("not setup correctly")
}
if (file.exists("pk.turnover.emax3-nonmem/pk.turnover.emax3-rounding.qs")) {
  unlink("pk.turnover.emax3-nonmem/pk.turnover.emax3-rounding.qs")
}
if (file.exists("pk.turnover.emax4-nonmem/pk.turnover.emax4.qs")) {
  unlink("pk.turnover.emax4-nonmem/pk.turnover.emax4.qs")
}

library(rxode2)
library(tidyverse)
library(babelmixr2)
```

I remember attending a virtual ACoP where Tim Waterhouse said "This
person is so convincing that the could sell NONMEM to a nlmixr
developer".  I was in the wrong meeting so I laughed and connected to
the correct meeting.

While he is correct, I don't really want to purchase a NONMEM license,
and I would think that individual pharmacometricians are the same: they
don't want to buy a personal license for the software they use at work (although
CROs might be different here).

That being said, I have used `NONMEM` long before I helped develop
`nlmixr2`, and I've always appreciated all that `NONMEM` brings to the
pharmacometrics community. I remember when I ran my first `NONMEM` model
I was amazed and wondered how it could calculate both individual and
population effects of a complicated system.

I still think `NONMEM` has an important role in our pharmacometrics
ecosystem today.

Still, our vision stands:

> To develop an R-based open-source nonlinear mixed-effects modeling
> software that can compete with commercial tools and is suitable for
> regulatory submissions.

which means it would be really convenient to have an interface to
convert `nlmixr2` models to `NONMEM`, and other tools, to make everyone's lives easier.

With this in mind, I am proud to announce the first `nlmixr2` to
`NONMEM` translator in [`babelmixr2`](https://github.com/nlmixr2/babelmixr2).

While this has been done before, the *method* whereby we are converting
between the two is novel and has some surprising advantages.

# How to use `NONMEM` with `nlmixr2`

To use `NONMEM` in nlmixr, you do not need to change your data or your
`nlmixr2` dataset. `babelmixr2` will do the heavy lifting here.

Lets take the classic warfarin example to start the comparison with...

The model we use in the `nlmixr2` vignettes is:

```{r}
library(babelmixr2)
# The nonmem translation requires the package pmxTools as well.
# You do not need to load it, simply have it available for use.
pk.turnover.emax3 <- function() {
  ini({
    tktr <- log(1)
    tka <- log(1)
    tcl <- log(0.1)
    tv <- log(10)
    ##
    eta.ktr ~ 1
    eta.ka ~ 1
    eta.cl ~ 2
    eta.v ~ 1
    prop.err <- 0.1
    pkadd.err <- 0.1
    ##
    temax <- logit(0.8)
    tec50 <- log(0.5)
    tkout <- log(0.05)
    te0 <- log(100)
    ##
    eta.emax ~ .5
    eta.ec50  ~ .5
    eta.kout ~ .5
    eta.e0 ~ .5
    ##
    pdadd.err <- 10
  })
  model({
    ktr <- exp(tktr + eta.ktr)
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    emax = expit(temax+eta.emax)
    ec50 =  exp(tec50 + eta.ec50)
    kout = exp(tkout + eta.kout)
    e0 = exp(te0 + eta.e0)
    ##
    DCP = center/v
    PD=1-emax*DCP/(ec50+DCP)
    ##
    effect(0) = e0
    kin = e0*kout
    ##
    d/dt(depot) = -ktr * depot
    d/dt(gut) =  ktr * depot -ka * gut
    d/dt(center) =  ka * gut - cl / v * center
    d/dt(effect) = kin*PD -kout*effect
    ##
    cp = center / v
    cp ~ prop(prop.err) + add(pkadd.err)
    effect ~ add(pdadd.err) | pca
  })
}
```

Next you have to figure out the command to run `NONMEM` (it is often
useful to use the full command path).  You can set it in
`options("babelmixr2.nonmem"="nmfe743")` or use
`nonmemControl(runCommand="nmfe743")`.  I prefer the `options()`
method since you only need to set it once.  This could also be a
function if you prefer (but I will not cover using the function here).

Lets assume you have `NONMEM` setup appropriately.  To run the
`nlmixr2` model using `NONMEM` you simply can run it directly:

```{r}
testthat::expect_error(nlmixr(pk.turnover.emax3, nlmixr2data::warfarin, "nonmem",
                              nonmemControl(readRounding=FALSE, modelName="pk.turnover.emax3")))
```

Note that a few options you may note in the `nonmemControl()` here is
`modelName` which helps control the output directory of `NONMEM` (if
not specified `babelmixr2` tries to guess based on the model name based on the input).

Now if you wanted, you could do the standard approach of changing
`sigdig`, `sigl`, `tol` etc, to get a successful `NONMEM` model
convergence, of course that is supported.

One of the other approaches is to **ignore** the rounding errors that
have occurred and read into `nlmixr2` anyway:

```{r}
# Can still load the model to get information (possibly pipe) and create a new model
f <- nlmixr(pk.turnover.emax3, nlmixr2data::warfarin, "nonmem",
            nonmemControl(readRounding=TRUE, modelName="pk.turnover.emax3"))
```

You may see more work happening than you expected to need for an already
completed model.  When reading in a NONMEM model, `babelmixr2` grabs:

- `NONMEM`'s objective function value
- `NONMEM`'s covariance (if available)
- `NONMEM`'s optimization history
- `NONMEM`'s final parameter estimates (including the ETAs)
- `NONMEM`'s `PRED` and `IPRED` values (for validation purposes)

These are used to solve the ODEs *as if they came from an nlmixr2*
optimization procedure.

This means that you can compare the `IPRED` and `PRED` values of
`nlmixr2`/`rxode2` and *know immediately* if your model validates.
This is similar to the procedure Kyle Baron advocates for validating a
NONMEM model against a `mrgsolve` model (see
https://mrgsolve.org/blog/posts/2022-05-validate-translation/). 

The advantage of this method is that you need to simply write one model to
get a validated `roxde2`/`nlmixr2` model.

In this case you can see the validation when you print the fit object:
```{r}
print(f)
```

That is in this case:

```
    IPRED relative difference compared to Nonmem IPRED: 0%; 95% percentile: (0%,0%); rtol=7.3e-06
    PRED relative difference compared to Nonmem PRED: 0%; 95% percentile: (0%,0%); rtol=6.57e-06
    IPRED absolute difference compared to Nonmem IPRED: atol=7.97e-05; 95% percentile: (2.18e-06, 0.00064)
    PRED absolute difference compared to Nonmem PRED: atol=6.57e-06; 95% percentile: (2.75e-07,0.00337)

```

Which means there are very few differences between the predictions
of `rxode2` and `NONMEM`, or this model is "validated".

Since it *is* a `nlmixr2` fit, you can do interesting things with this fit that you couldn't do in `NONMEM` or even in another translator.  For example, if you wanted to add a covariance step you can with `getVarCov()`:

```{r}
getVarCov(f)
```

`nlmixr2` is more generous in what constitutes a covariance step.  The
`r,s` covariance matrix is the "most" successful covariance step for
`focei`, but the system will fall back to other methods if necessary.

While this covariance matrix is not `r,s`, and should be regarded with
caution, it can still give us some clues on why this things are not working in
`NONMEM`.

When examining the fit, you can see the shrinkage is high for `temax`, `tktr` and `tka`, so they could be dropped, makiing things more likely to converge in `NONMEM`.

If we use model piping to remove the parameters, the new run will
start at the last model's best estimates (saving a bunch of model development
time).

In this case, I specify the output directory `pk.turnover.emax4` with
the control and get the following:

```{r}
f2 <- f %>% model(ktr <- exp(tktr)) %>%
  model(ka <- exp(tka)) %>%
  model(emax = expit(temax)) %>%
  nlmixr(data=nlmixr2data::warfarin, est="nonmem",
         control=nonmemControl(readRounding=FALSE,
                               modelName="pk.turnover.emax4"))
```

You can see the `NONMEM` run is now successful and validates against
the `rxode2` model below:

```{r}
f2
```

One thing to emphasize: unlike other translators, you will know immediately
if the translation is off because the model will not validate.  Hence
you can start this process with confidence - you will know
immediately if something is wrong.

# Conclusion

The first release of `babelmixr2` includes a `NONMEM` translation function.
The advantages of this are:

- For `nlmixr2` development, we can easily compare to `NONMEM` to see how we're doing with respect to the current gold standard.
  
- For people who are using `rxode2` and `NONMEM`, writing a model with
  `nlmixr2` syntax and using it to run `NONMEM` will let you only
  write one model, and save you time debugging and coding it yourself.
  
- For pharmacometricians using `NONMEM`, you can take an unsuccessful
  `NONMEM` fit, get information (covariance shrinkage, etc) about the
  model and you will be able to make informed decisions on how to
  proceed.

Many of these advantages come from the fact that `babelmixr2` leans into supporting `nlmixr2` development for those fluent in `NONMEM` and
having `nlmixr2` available can help pharmacometricans in daily tasks, even when they need to use another tool.

The astute reader will also notice that the full model
runs in `nlmixr2`'s `focei` without adjustment.  I would like to caution
that this doesn't mean that `nlmixr2`'s focei is *better*: rather, it is
*different* (as mentioned in a previous blog post).  I have seen cases
in the past where something runs better in `NONMEM` than `nlmixr2` so
comparisons based on a single model should be regarded with caution (I
no longer have these examples available, though, soyou'll have to take my word for it).

Thanks for reading!

