---
title: "babelmixr2, nlmixr2 and NONMEM"
author: "Matt Fidler and the nlmixr2 Development Team"
date: "2022-10-25"
slug: []
bibliography: [refs.bib]
link-citations: true
csl: vancouver.csl
categories: [babelmixr2]
tags: [new-version, NONMEM]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (dir.exists("content")) {
  setwd("content/blog/2022-11-11-babelmixr2-nonmem")
}
if (!dir.exists("pk.turnover.emax3-nonmem")) {
  stop("not setup correctly")
}
if (file.exists("pk.turnover.emax3-nonmem/pk.turnover.emax3-rounding.qs")) {
  unlink("pk.turnover.emax3-nonmem/pk.turnover.emax3-rounding.qs")
}
if (file.exists("pk.turnover.emax3-nonmem/pk.turnover.emax4.qs")) {
  unlink("pk.turnover.emax3-nonmem/pk.turnover.emax4.qs")
}

library(rxode2)
library(tidyverse)
library(babelmixr2)
```

I remember attending a virtual ACoP where Tim Waterhouse said "This
person is so convincing that the could sell NONMEM to a nlmixr
developer".  I was in the wrong meeting so I laughed and connected to
the correct meeting.

While he is correct, I don't really want to purchase a NONMEM license,
I would believe that individual pharmacometricians are the same, they
do not want to buy a license of the software the use at work (although
CROs are different here).

That being said, I have used `NOMMEM` long before I helped develop
`nlmixr2`, and has always appreciated all that `NONMEM` brought to the
pharmacometrics community. I remember when I ran my first `NONMEM` model
I was amazed and wondered how it could calculate both individual and
population effects of a complicated system.

I still think `NONMEM` has an important part in our pharmacometrics
ecosystem today.

Still our vision stands:

> To develop an R-based open-source nonlinear mixed-effects modeling
> software that can compete with commercial tools and is suitable for
> regulatory submissions.

which means it would be really convenient to have an interface to
convert to `NONMEM` and other tools to make sure we can compete with
the commercial tools.

With this in mind, I am proud to announce the first `nlmixr2` to
`NONMEM` translator in `babelmixr2`.

While this has been done before, the *method* that we are converting
between the two is novel and has some surprising advantages.

# How to use `NONMEM` with `nlmixr2`

To use `NONMEM` in nlmixr, you do not need to change your data or your
nlmixr2 dataset.  `babelmixr2` will do the heavy lifting here.

Lets take the classic `warfarin` example to start the comparison with:

The model we use in the `nlmixr2` vignettes is:

```{r}
library(babelmixr2)
# The nonmem translation requires the package pmxTools as well.
# You do not need to load it, simply have it available for use.
pk.turnover.emax3 <- function() {
  ini({
    tktr <- log(1)
    tka <- log(1)
    tcl <- log(0.1)
    tv <- log(10)
    ##
    eta.ktr ~ 1
    eta.ka ~ 1
    eta.cl ~ 2
    eta.v ~ 1
    prop.err <- 0.1
    pkadd.err <- 0.1
    ##
    temax <- logit(0.8)
    tec50 <- log(0.5)
    tkout <- log(0.05)
    te0 <- log(100)
    ##
    eta.emax ~ .5
    eta.ec50  ~ .5
    eta.kout ~ .5
    eta.e0 ~ .5
    ##
    pdadd.err <- 10
  })
  model({
    ktr <- exp(tktr + eta.ktr)
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    emax = expit(temax+eta.emax)
    ec50 =  exp(tec50 + eta.ec50)
    kout = exp(tkout + eta.kout)
    e0 = exp(te0 + eta.e0)
    ##
    DCP = center/v
    PD=1-emax*DCP/(ec50+DCP)
    ##
    effect(0) = e0
    kin = e0*kout
    ##
    d/dt(depot) = -ktr * depot
    d/dt(gut) =  ktr * depot -ka * gut
    d/dt(center) =  ka * gut - cl / v * center
    d/dt(effect) = kin*PD -kout*effect
    ##
    cp = center / v
    cp ~ prop(prop.err) + add(pkadd.err)
    effect ~ add(pdadd.err) | pca
  })
}
```

Next you have to figure out the command to run `NONMEM` it is often
useful to use the full command path.  You can use it in
`options("babelmixr2.nonmem"="nmfe743")` or use
`nonmemControl(runCommand="nmfe743")`.  I prefer the `options()`
method since you only need to set it once.  This could also be a
function if you prefer (but I will not cover using the function here).

Lets assume you have `NONMEM` setup appropriately.  To run the
`nlmixr2` model using `NONMEM` you simply can run it directly:

```{r}
testthat::expect_error(nlmixr(pk.turnover.emax3, nlmixr2data::warfarin, "nonmem",
                              nonmemControl(readRounding=FALSE, modelName="pk.turnover.emax3")))
```

Note that a few options you may note in the `nonmemControl()` here is
`modelName` which helps control the output directory of `NONMEM` (if
not specified `babelmixr2` tries to guess based on the model name based on the input).

Now if you wanted, you could do the standard approach of changing
`sigdig`, `sigl`, `tol` etc., to get a successful `NONMEM` model
convergence, of course that is supported.

One of the other approaches is to **ignore** the rounding errors that
have occurred and read into `nlmixr2` anyway:

```{r}
# Can still load the model to get information (possibly pipe) and create a new model
f <- nlmixr(pk.turnover.emax3, nlmixr2data::warfarin, "nonmem",
            nonmemControl(readRounding=TRUE, modelName="pk.turnover.emax3"))
```

You may see more work than you expected to read in an already
completed model.  When reading in a NONMEM model, `babelmixr2` grabs:

- `NONMEM`'s objective function value
- `NONMEM`'s covariance (if available)
- `NONMEM`'s optimization history
- `NONMEM`'s final parameter estimates (including the ETAs)
- `NONMEM`'s `PRED` and `IPRED` values

These are used to solve the ODEs *as if they came from a nlmixr2*
optimization procedure.

This means that you can compare the `IPRED` and `PRED` values of
`nlmixr2`/`rxode2` and *know immediately* if your model validates.
This is similar to the procedure Kyle Baron advocates for validating a
NONMEM model against a `mrgsolve` model (see
https://mrgsolve.org/blog/posts/2022-05-validate-translation/). 

The advantage of this method is that you simply write one model to
have a validated `roxde2`/`nlmixr2` model.

In this case you can see the validation when you print the fit object:
```{r}
print(f)
```

That is in this case:

```
    IPRED relative difference compared to Nonmem IPRED: 0%; 95% percentile: (0%,0%); rtol=7.3e-06
    PRED relative difference compared to Nonmem PRED: 0%; 95% percentile: (0%,0%); rtol=6.57e-06
    IPRED absolute difference compared to Nonmem IPRED: atol=7.97e-05; 95% percentile: (2.18e-06, 0.00064)
    PRED absolute difference compared to Nonmem PRED: atol=6.57e-06; 95% percentile: (2.75e-07,0.00337)

```

Which means there is very little differences between the predictions
of `rxode2` and `NONMEM`, or this model is validated.

Since it *is* a `nlmixr2` fit, you can do interesting things with this fit that you couldn't do in `NONMEM` or even in another translator.  For example, if you wanted to add a covariance step you can with `getVarCov()`:

```{r}
getVarCov(f)
```

nlmixr2 is more generous in what constitutes a covariance step.  The
`r,s` covariance matrix is the "most" successful covariance step for
`focei`, but will fall back to other methods.

While this covariance is not `r,s`, and should be regarded with
caution, it scan still give clues on why this is not working in
`NONMEM`.

When examining the fit, you can see the shrinkage is high for temax tktr and tka, so they could be dropped with a model in nonmem that is more likely to converge in NONMEM.

If we use model piping to remove the parameters, then the new run will
start at the last model's best estimates (saving my model development
time too).

In this case, I specify the output directory `pk.turnover.emax4` with
the control and get the following:

```{r}
f2 <- f %>% model(ktr <- exp(tktr)) %>%
  model(ka <- exp(tka)) %>%
  model(emax = expit(temax)) %>%
  nlmixr(data=nlmixr2data::warfarin, est="nonmem",
         control=nonmemControl(readRounding=FALSE,
                               modelName="pk.turnover.emax4"))
```

You can see the `NONMEM` run is now successful and validates against
the `rxode2` model.


One thing to note, you will know immediately if the translation is off
because the model will not validate.  Hence you can start this process
with confidence that you will know immediately if something is wrong.

# Conclusion

The first release of `babelmixr2` includes a `NONMEM` translation.
The advantages of this are:

- For `nlmixr2` development, we can easily compare to NONMEM to try to
  make this competitive with commercial software.
  
- For people who are using `rxode2` and `NONMEM`, writing a model with
  nlmixr2 syntax and using it to run NONMEM will make sure you only
  write one model and save you time debugging and coding it yourself.
  
- For users, you can take an unsuccessful NONMEM step, get information
  (ie covariance shrikage etc) about the model and you will be able to
  make informed decisions on how to proceed.

Many of these advantages are because `babelmixr2` is not platform
agnostic, rather it leans into helping `nlmixr2` development and
having `nlmixr2` help pharmcometricans in daily tasks.

If you are an astute reader, you will also notice that the full model
runs in `nlmixr2`'s focei without adjustment.  I would like to caution
that this doesn't mean that `nlmixr2`'s focei is *better* rather it is
*different* (as mentioned in the previous blog post).  I have seen cases
in the past where something runs better in `NONMEM` than `nlmixr2` so
comparisons based on a single model should be regarded with caution (I
no longer have these examples though so I can display them in the blog
either).

