---
title: Stepwise covariate modeling with nlmixr2scm
author: Justin Wilkins
date: '2026-10-02'
slug: []
categories: [nlmixr2, nlmixr2scm]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# nlmixr2scm is on CRAN!

Happily, `nlmixr2scm`, which does stepwise covariate
modeling (SCM) for `nlmixr2` models, is now on CRAN. It took a while - 
seems they're very busy now that everyone is churning out packages with
AI - but it's finally there.

```r
install.packages("nlmixr2scm")
```

If you've built covariate models in NONMEM, you've probably used `scm`
in [Perl-speaks-NONMEM](https://github.com/UUPharmacometrics/PsN).
`nlmixr2scm` follows the same approach
([Jonsson & Karlsson, 1998](https://doi.org/10.1023/A:1011970125687)), and
it works directly on an nlmixr2 fit. You don't need to write control
streams or config files or any of that kind of thing, although the command
will probably be quite lengthy...

## How SCM works

SCM searches in two phases, and both use likelihood-ratio tests (you probably
know this, but a bit of underlining is always useful):

1. **Forward inclusion.** Starting from the base model, it tests every
   candidate parameter-covariate relationship. The one that gives the
   largest significant drop in OFV is added, and the search repeats until
   no remaining candidate passes the forward threshold (p < 0.05 by
   default).
2. **Backward elimination.** Starting from the final forward model, it
   removes each included relationship one at a time. Any relationship whose
   removal does not raise the OFV enough to pass a stricter threshold
   (p < 0.01 by default) is dropped.

`runSCM()` writes the covariate terms into the model body for you.
Continuous covariates are centered (at the median, or at a reference value
you choose), and categorical covariates are expanded into indicator columns.
It takes the original, untransformed data.

## A quick example

We'll start from a one-compartment warfarin PK model:

```{r base, message=FALSE, warning=FALSE}
library(nlmixr2)
library(nlmixr2scm)

pkdata <- nlmixr2data::warfarin[nlmixr2data::warfarin$dvid == "cp", ]

warf_pk <- function() {
  ini({
    tka <- log(1.15)
    tcl <- log(0.135)
    tv  <- log(7.0)
    eta.ka ~ 0.40
    eta.cl ~ 0.25
    eta.v  ~ 0.10
    prop.err <- 0.10
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv + eta.v)
    linCmt() ~ prop(prop.err)
  })
}

fit_base <- nlmixr2(warf_pk, pkdata, est = "focei",
                    control = foceiControl(print = 0))
```

Let's test body weight (as a power or linear relationship) and sex on
clearance and volume...

```{r scm, message=FALSE, warning=FALSE}
scm <- runSCM(
  fit        = fit_base,
  data       = pkdata,             # needed: the base model doesn't use wt or sex
  varsVec    = c("cl", "v"),       # parameters to test
  covarsVec  = "wt",               # continuous covariates
  catvarsVec = "sex",              # categorical covariates
  shapes     = c("power", "lin"),
  pVal       = list(fwd = 0.05, bck = 0.01),
  saveModels = FALSE,              # quick exploratory run: nothing written to disk
  print      = 0,
  confirm    = FALSE
)
```

Printing the result gives a short overview:

```{r print-scm}
scm
```

For the full picture, use `summary()`. It lists the options the search ran
with, compares the base model with the final forward and final backward
models (OFV, ΔOFV, AIC, BIC, parameter count and covariates), prints the
step-by-step and all-candidate tables, and tells you where the models and
report files were saved on disk, if they were:

```{r summary-scm}
summary(scm)
```

Underneath, the result is a list you can dig into:

- `summaryTable` has one row per candidate tested in both phases, with the
  OFV, ΔOFV, p-value, AIC/BIC, and whether the relationship was included.
- `resFwd[[1]]` and `resBck[[1]]` are the final forward and backward fits.
  These are ordinary nlmixr2 fit objects, so you can use your usual
  diagnostics and plots on them.

```{r results}
scm$resBck[[1]]$parFixedDf
```

If you want to control exactly which relationships are tested, pass
`pairsVec` instead of the full `varsVec` × `covarsVec` grid. You can also use
`includedRelations` to force known relationships (allometric weight on
clearance, for example) into the backward-elimination model.

## Real-world searches

Real-world SCM runs often fit hundreds of candidate models, and `nlmixr2scm` has
several features for that:

- **Caching and resume.** With `saveModels = TRUE` (the default), every
  candidate fit and step table is saved to `outputDir`. If a run is
  interrupted, call `runSCM()` again with the same `outputDir` and it picks 
  up where it stopped. `restart = TRUE` starts from scratch instead.
- **Parallel fitting.** `workers` fits the candidates within each step in
  parallel through `future`. `rxThreads` sets how many `rxode2` threads each
  worker uses. `runSCM()` checks `workers * rxThreads` against your physical
  core count before it starts, so that it doesn't quietly murder your computer
  by maxing out its capacity.

  ```r
  runSCM(fit_base, data = pkdata, varsVec = c("cl", "v"), covarsVec = "wt",
         workers = 4, rxThreads = 1)
  ```

- **Protection against bad fits.** Some candidate fits stall at the
  zero-effect initial estimate. This happens sometimes with ODE models. 
  `runSCM()` detects these stalled candidates
  and refits them from a starting value obtained with a quick 1-D profile.
  Candidates that land on an implausible OFV are retried from perturbed
  initial estimates, and the best attempt is kept. Both behaviours are
  switched on by default. By default, a p-value that underflows to zero
  also counts as implausible, which means very strong effects (an OFV drop
  of more than about 70 for one parameter) get retried too. If you trust
  those, `retryOnUnderflow = FALSE` saves the extra fits.
- **Fixed centering.** `centers = c(wt = 70)` centers weight at 70 kg
  instead of the per-dataset median. The coefficients then keep the same
  reference across datasets.

## Learn more

- The vignette gives the full argument reference and a longer worked
  example: `vignette("runSCM", package = "nlmixr2scm")`.
- Source, bug reports, and feature requests are on
  [GitHub](https://github.com/nlmixr2/nlmixr2scm).
- The [NEWS file](https://github.com/nlmixr2/nlmixr2scm/blob/main/NEWS.md)
  has release notes. 

## Acknowledgements

`nlmixr2scm` was written together with Matthew Fidler and Yaping Liu and contains
contributions from Bill Denney, Vipul Mann, Vishal Sarsani, and Christian
Bartels. The design owes a great deal to the original PsN `scm` implementation by
Lars Lindbom, Niclas Jonsson, Pontus Pihlgren, Mats Karlsson, Andrew Hooker,
Kajsa Harling, Rikard Nordgren and Svetlana Freiberga. Thank you all!
