---
title: "Tiny Doses for Tiny Babies -- How much is too much?"
subtitle: "A hands-on introduction to pharmacometrics in R"
author: Matthew Fidler & Janice Goh Jai Ni
date: '2026-08-11'
slug: []
categories: [nlmixr2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, fig.width = 7.8, fig.height = 4.6,
                      dpi = 110, out.width = "100%")
library(nlmixr2)
library(dplyr)
library(ggplot2)
rxode2::rxSetSilentErr(1L)
theme_set(theme_bw())
```

Preterm infants are highly susceptible to seizures primarily due to
brain immaturity and an increased vulnerability to acute brain
injuries and systemic imbalances, so if an infant is delivered
pre-term they may be likely to have seizures.

So, with a preterm infant with seizures -- what can we do in data
analytics to support the right drug and dose the first time?  As
stated by Paracelsus, the dose makes the poison: Too
little and the seizures continue, too much and the infant stops
breathing. What information do we have -- the birth weight (900 g) and
the [APGAR
score](https://medlineplus.gov/ency/article/003402.htm).

To complicate the analysis, this was not created from a well
controlled clinical study taking intensive (15+) blood samples from
these pre-term infants.  They only have less than a teacup of blood in
their entire body!  What exists is scraps of concentrations, one here
another 3+ days later across only a few infants.  All on different
doses, measured with different people and labs.

Even with these limitations, we now know that the correct dose of a
drug,
[phenobarbital](https://medlineplus.gov/druginfo/meds/a682007.html),
used to treat seizures in these infants.  It is in the drug label, a
dose given in mg/kg.

Getting a defensible dose out of that is what pharmacometrics does.
It helps understand what the drug does to the body and what the body
does to the drug.  It helps understand the concentrations of doses in
individuals and populations and helps pick the right drug for the
population and in some instances for the individual.

This post works through the analysis of this dataset end to end in R with
[nlmixr2](https://nlmixr2.org) -- from 155 scattered measurements to a defensible
dosing label while using an open-source tool for drug-development.

## If you have fit a mixed model, you have done half of this

Drug exposure is a dynamic process. As a drug is being absorbed, it is
also being distributed throughout the body, and cleared at the same
time. Pharmacometric (PMX) models help us to address these dynamic
processes acting together via ordinary differential equations (ODE). Think
like a filling a bucket of water with a hole at the bottom - how fast
the bucket fills depends on how fast the rate of water flows in vs
flows out.

In pharmacometrics we understand, like other fields, we have different
infant sizes, and different ways the drug flows, which means a method
to solve differential equations, and characterize differences between
individuals **at the same time**: the mean function inside the
nonlinear mixed model that is not a line or a spline, it is the solution to
a system of ODEs describing how drug moves through the body.

These movements and concentrations are characterized by two parameters
the *clearance* (volume of blood eliminated of drug per hour) and
*volume of distribution* (how widely the drug spreads through tissue),
you can use these are physiological quantities to inform the mechanism
of how the drug works:

- **They extrapolate.** Once you know clearance, you know the
  concentration for a dose nobody in your study ever received.

- **They can describe meaningful relationships.** Clearance scales with body
  size and kidney function; volume scales with body water. This has
  been studied by the field in models for quite awhile.  Therefore,
  when you are using population level ways of better describing the
  data (ie covariates), you put them on where biology says they
  belong, not where it simply better describe the data.

- **They separate the parts of a regimen.** The loading dose is
  governed by volume; the maintenance dose is governed by clearance.
  Estimating them separately is exactly what lets you design a regimen,
  and it is invisible if all you have is a fitted curve.

So the questions pharmacometrics answers are dosing questions: what
dose, how often, by what route; who needs a different one (children,
renal impairment, drug interactions, obesity etc); what happens when a dose
is missed; how much exposure produces the effect and how much starts
producing harm; and whether the Phase III trial is worth running at
all, at which dose.

Dosing sections of drug labels are built from these models. Pediatric
doses are routinely approved by adult model extrapolation rather than
and unjustified pediatric trial.


In the clinic, these models help therapeutic drug monitoring --
vancomycin (for MRSA infections), busulfan (for cancer), infliximab
(Autoimmune/inflamitory) -- where the need of the drug is great and
individual variability of effect / toxicity needs to be carefully
balanced in each individual.  Both FDA and EMA have written guidance on
population pharmacokinetics. It is a small field with an unusually
direct path from a model object to what happens to a patient.

## The data

In this example, the dataset is real and famous: 59 preterm infants
given phenobarbital, published by Grasela and Donn in 1985. It ships
with nlmixr2 as `pheno_sd`, and if you have `nlme` installed you
already have a copy of it as `nlme::Phenobarb`.

```{r data-head}
library(nlmixr2)
head(pheno_sd, 4)
```

The columns are the pharmacometric standard: `ID`, `TIME`, `AMT`
(amount or dose given), `DV` (dependent variable, in this case
concentration), `EVID` (1 = dose, 0 = observation) plus covariates
`WT` (kg) and `APGR` (Apgar score). The nlmixr2 format is very similar
to the NONMEM format since it is prevalent across the field.

One of the first things to do is do some exploration of the data:

```{r data-sparsity}
obs <- subset(pheno_sd, EVID == 0)

c(infants       = length(unique(pheno_sd$ID)),
  doses         = sum(pheno_sd$EVID == 1),
  concentrations = nrow(obs))

table(`samples per infant` = table(obs$ID))
```

Nine infants have exactly one concentration. Forty percent have two or
fewer. There are 589 doses and only 155 measurements — the infants were
dosed relentlessly and sampled infrequently during routine care.

You can see this clearly with the following exploratory plot:

```{r fig-data, fig.height = 5, echo = FALSE}
ord  <- pheno_sd %>% group_by(ID) %>%
  summarise(WT = first(WT), .groups = "drop") %>%
  arrange(WT) %>% mutate(row = row_number())
obsP  <- left_join(obs, ord, by = "ID", suffix = c("", ".o"))
doseP <- left_join(subset(pheno_sd, EVID == 1), ord, by = "ID",
                   suffix = c("", ".o"))

ggplot() +
  geom_point(data = doseP, aes(TIME, row), shape = 124,
             colour = "black", size = 1.6) +
  geom_point(data = obsP, aes(TIME, row, fill = DV), shape = 21,
             size = 2.4, colour = "black", stroke = .3) +
  scale_fill_viridis_c(name = "Conc.\n(mg/L)", option = "C", end = .92) +
  scale_y_continuous(breaks = c(1, 15, 30, 45, 59),
                     labels = function(i) sprintf("%.1f kg", ord$WT[i])) +
  labs(x = "Time (h)", y = "59 infants, ordered by body weight",
       subtitle = "Ticks are doses; circles are the 155 measured concentrations") +
  theme(panel.grid.minor = element_blank())
```


## A first model

To describe the model you can use an ordinary R function with two
blocks -- an `ini({})` and a `model({})` block:

```{r model-1}
pheno <- function() {
  ini({
    tcl <- log(0.008)   # log typical clearance (L/h)
    tv  <- log(0.6)     # log typical volume (L)
    eta.cl ~ 1          # between-infant variability on clearance
    eta.v  ~ 1          # between-infant variability on volume
    prop.sd <- 0.1      # proportional residual error
  })
  model({
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    d/dt(central) <- -(cl / v) * central
    cp <- central / v
    cp ~ prop(prop.sd)
  })
}
```

Three things are worth pointing out, because they are the whole
language:

- `ini({})` holds initial estimates. `<-` declares a fixed effect (and residual estimates)
  `~` declares a random effect with its initial variance.

- `model({})` is the model itself. `d/dt(central)` is a differential
  equation, declared in Lebneitz notation.  This one says drug leaves
  the central compartment at a rate proportional to how much is
  there. Dosing is driven by the data (`AMT`/`EVID`), so doesn't
  appears in this model code.

- The last line, `cp ~ prop(prop.sd)`, says the observed concentration
  is the predicted `cp` with proportional residual error. That is where
  the model meets dependent variable, `DV`.

Everything else is plain R. `WT` will be a column in the data;
`exp()` is `exp()`.

Fitting it:

```{r fit-1, results = "hide"}
# note here we use the := nlmixr2 save function to cache the fit
# since nonlinear mixed effects models can take some time to run.
fit1 := nlmixr2(pheno, pheno_sd, est = "saem",
                control = saemControl(print = 0, seed = 42),
                table = tableControl(cwres = TRUE, npde = TRUE))
```

You can print out the entire fit object, which is a modified
data-frame of predictions and estimates together.  Another way to look
at the data is simply to look at the estimate:

```{r fit-1-print}
fit1$parFixedDf[, c("Estimate", "Back-transformed", "BSV(CV%)")]
```

Typical clearance is about `r
signif(exp(fit1$parFixedDf["tcl","Estimate"]), 3)` L/h and typical
volume about `r signif(exp(fit1$parFixedDf["tv","Estimate"]), 3)` L.
But look at the last column: between-infant variability of **37% on
clearance and 47% on volume**. Two infants drawn at random from this
population can easily differ two-fold in exposure on the same dose.
That is the number that makes a single flat dose dangerous, and the
next question is whether any of it is explainable.

## One line, and half the variability disappears

These infants weigh between 0.6 and 3.6 kg -- a six-fold range in body
size. It would be strange if that did not matter. Larger bodies have
larger livers and kidneys and more water to distribute drug into, and
the standard way to encode that is allometric scaling: clearance scales
with weight to the 3/4 power, volume scales linearly.

In general this covariate, and others, are there to explain why
certain babies/individuals have more drug in their system than others.

The change to the model is one line each.  In nlmixr2 instead of
having to rewrite the whole model we can change pieces of the model
with model piping:

```{r model-2}

phenoWt <-  pheno |>
  # first we modify the cl line
  model(cl <- exp(tcl + eta.cl + cov.cl.wt*log(WT/1.3))) |>
  # then the v line
  model(v  <- exp(tv  + eta.v  + cov.v.wt*log(WT/1.3))) |>
  # then we add the covariate effects to the ini block
  ini(cov.cl.wt=0.75, cov.v.wt=1)

fit2 := nlmixr2(phenoWt, pheno_sd, est = "saem",
                control = saemControl(print = 0, seed = 42),
                table = tableControl(cwres = TRUE, npde = TRUE))
```

```{r compare1}
tibble(
  model = c("Base", "+ allometric weight"),
  OBJF  = c(fit1$objf, fit2$objf),
  AIC   = c(AIC(fit1), AIC(fit2)),
  `BSV CL (CV%)` = c(fit1$parFixedDf["tcl", "BSV(CV%)"],
                     fit2$parFixedDf["tcl", "BSV(CV%)"]),
  `BSV V (CV%)`  = c(fit1$parFixedDf["tv", "BSV(CV%)"],
                     fit2$parFixedDf["tv", "BSV(CV%)"])
) %>% mutate(across(where(is.numeric), ~ signif(.x, 4)))
```

The objective function drops by about `r round(fit1$objf - fit2$objf)`
points for two estimated parameters. Between-infant variability falls
from 37% to 17% on clearance and from 47% to 18% on volume.  This
means more than half of the variability in these infants is just body
weight which is why the infants are dosed by weight.

## Leaning into the AI estimation methods

Experimentally, as of nlmixr2 7.0, we also allow
variable-auto-encoding method (`vae`) to automatically select covariates
with the l0learn method from google labs.  These new nonlinear mixed
effects routines lean into the new advancements in AI and apply them
to this space.  Using this method we get:

```{r vae}
fit3 := nlmixr2(pheno, pheno_sd, est = "vae",
                control=vaeControl(print=0),
                table = tableControl(cwres = TRUE, npde = TRUE))
```

This shows a full covariate search would have picked weight throughout
as a covariate.  You can see the full model with the covariates filled in here:

```{r vaeCov}
print(fit3$ui)
```

And the same comparison for these weight covariates are here:

```{r compare2}
tibble(
  model = c("Base", "+ allometric weight"),
  OBJF  = c(fit1$objf, fit3$objf),
  AIC   = c(AIC(fit1), AIC(fit3)),
  `BSV CL (CV%)` = c(fit1$parFixedDf["tcl", "BSV(CV%)"],
                     fit3$parFixedDf["tcl", "BSV(CV%)"]),
  `BSV V (CV%)`  = c(fit1$parFixedDf["tv", "BSV(CV%)"],
                     fit3$parFixedDf["tv", "BSV(CV%)"])
) %>% mutate(across(where(is.numeric), ~ signif(.x, 4)))
```

Further reducing the variability in volume to 14%

## Does the model deserve to be believed?

First, individual profiles. This is where the borrowing of strength
becomes visible:

```{r fig-augpred, fig.height = 5, echo = FALSE}
nobs <- table(obs$ID)
wtI  <- pheno_sd %>% group_by(ID) %>% summarise(WT = first(WT), .groups = "drop")
sel  <- c(56, 28, 11, 2, 4, 8)
lab  <- setNames(sprintf("ID %s  (%.1f kg, %d sample%s)", sel,
                         wtI$WT[match(sel, wtI$ID)], nobs[as.character(sel)],
                         ifelse(nobs[as.character(sel)] == 1, "", "s")), sel)

ap <- as.data.frame(augPred(fit3)) %>% filter(id %in% sel, ind != "Observed")
ob <- as.data.frame(fit3) %>% filter(ID %in% sel) %>%
  transmute(id = as.character(ID), time = TIME, values = DV)
ap$id <- factor(ap$id, levels = sel); ob$id <- factor(ob$id, levels = sel)

ggplot(ap, aes(time, values, colour = ind)) +
  geom_line(linewidth = .7) +
  geom_point(data = ob, colour = "black", size = 2.2) +
  facet_wrap(~ id, scales = "free_x", labeller = labeller(id = lab)) +
  labs(x = "Time (h)", y = "Phenobarbital (mg/L)", colour = NULL) +
  theme(legend.position = "top")
```

Infants 56 and 28 contributed **one measurement each**, and both come
out with a complete concentration-time profile.
The individual (red) and population (blue)
curves nearly coincide -- with one data point there is nothing to move
the individual away from the population, which is exactly the right
behavior. For infants 2 and 8, with three samples each, the two curves
separate: there is now enough information to say this infant is not
typical.

Second, a prediction-corrected visual predictive check — simulate many
replicates of the study from the fitted model and ask whether the real
data looks like a plausible draw:

```{r fig-vpc, echo = TRUE, results = "hide"}
vpcPlot(fit3, n = 300, pred_corr = TRUE, show = list(obs_dv = TRUE)) +
  labs(x = "Time (h)", y = "Prediction-corrected phenobarbital (mg/L)") +
  theme_bw()
```

The observed median and 5th/95th percentiles (solid and dashed lines)
sit inside the simulated bands across the study. With 155 observations
the percentiles are noisy but it is not misspecified.

## Answering the actual question

This analysis showed we have
a model that says how clearance and volume change with body weight, so
we can simulate regimens for infants across the weight range and ask
which regimen actually keeps them in the therapeutic window of roughly
15-40 mg/L.

Two candidates. A **flat** regimen — one dose for every infant, chosen
to be correct for the average 1.5 kg baby in this study. And a
**weight-based** regimen in mg/kg. Both use a loading dose (which
volume of distribution determines) followed by twice-daily maintenance
(which clearance determines).

The simulation reuses the fitted object directly — `rxSolve()` takes
`fit3` and samples new infants from its estimated between-subject
distribution:

```{r sim, results = "hide"}
set.seed(99)
wts <- seq(0.7, 3.5, by = 0.2)

simOne <- function(w, reg) {
  flat <- grepl("^Flat", reg)
  ev <- et(amt = if (flat) 30  else 20   * w) %>%                # loading
        et(amt = if (flat) 1.9 else 1.25 * w,                    # maintenance
           ii = 12, until = 24 * 14, time = 12) %>%
        et(seq(24 * 7, 24 * 14, by = 4))                         # sample week 2
  rxSolve(fit3, ev, nSub = 400, iCov = data.frame(WT = w),
          addDosing = FALSE) %>%
    as.data.frame() %>% mutate(WT = w, regimen = reg)
}

grid <- expand.grid(
  WT = wts,
  regimen = c("Flat: 30 mg load + 1.9 mg q12h",
              "Weight-based: 20 mg/kg load + 1.25 mg/kg q12h"),
  stringsAsFactors = FALSE)

sim <- bind_rows(Map(simOne, grid$WT, grid$regimen))
```

```{r fig-regimen, echo = FALSE}
res <- sim %>% group_by(regimen, WT) %>%
  summarise(lo = quantile(cp, .05), med = median(cp),
            hi = quantile(cp, .95), .groups = "drop")

ggplot(res, aes(WT, med)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 15, ymax = 40,
           fill = "grey70", alpha = .35) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = regimen), alpha = .35) +
  geom_line(aes(colour = regimen), linewidth = .9) +
  scale_y_log10() +
  labs(x = "Body weight (kg)", y = "Phenobarbital, week 2 (mg/L)",
       colour = NULL, fill = NULL,
       subtitle = "Shaded band is the 15-40 mg/L therapeutic window; ribbons are 90% prediction intervals") +
  theme(legend.position = "top", legend.direction = "vertical")
```

```{r tab-regimen}
sim %>%
  filter(round(WT, 1) %in% c(0.7, 1.5, 2.3, 3.1)) %>%
  group_by(regimen, WT) %>%
  summarise(median = median(cp),
            `% in window` = 100 * mean(cp >= 15 & cp <= 40),
            .groups = "drop") %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%
  arrange(regimen, WT)
```

The flat regimen was chosen to be *right* for the average infant, and
it is: at 1.5 kg the median lands at 22 mg/L, comfortably mid-window,
with virtually every infant in range. But it fails at both ends. The
median 0.7 kg infant sits at 41 mg/L -- above the top of the window
entirely -- while at 3.1 kg the median has fallen to 12 mg/L and fewer
than 4% of infants reach a therapeutic concentration at all. That is a
three-and-a-half-fold exposure gradient across the population, driven
entirely by a covariate that was sitting in the dataset the whole time.

The weight-based regimen holds the median between 19 and 25 mg/L across
the entire weight range, with 96% or more of simulated infants inside
the window at every weight.

There is a subtlety worth noticing in that curve: the weight-based line
still drifts gently *upward* with weight then plateaus.

**A caveat, stated clearly:** this is a teaching example on a 1985
dataset, deliberately kept to one compartment and one covariate. It is
not dosing guidance. Real neonatal phenobarbital dosing accounts for
postnatal age and maturation, and would be under the care of a
qualified doctor and pharmacist. The *workflow*, though, is the real
one.  Also note, the original dataset was

## Where this sits in the R ecosystem

The part that is easy to miss in a walkthrough is that nothing above
required leaving R. Many tools in pharmacometrics require a shuttle
back and forth between 2 different tools and ecosystems.

The fit object is a modified tibble. `as.data.frame(fit2)` gives you
observations, predictions, residuals, and empirical Bayes estimates in
a tidy frame, which is why every plot in this post is plain `ggplot2`
and every summary is plain `dplyr`. There is no export step, no
intermediate file format, no separate post-processing tool with its own
data conventions. That sounds like a convenience and is really a
reproducibility property: the places where analyses usually become
irreproducible are the boundaries between tools, and this workflow does
not have any.

That matters concretely. The whole thing — data prep, fit, diagnostics,
simulation, report — is one Quarto or R Markdown document (this post is
one).

One approach to keep this reproducible is to in the dependencies with
`renv`, wire the fits into a `targets` pipeline so they only re-run when
the model changes, and someone else can reproduce the dosing
recommendation from the raw data with the R script or report.

The second thing worth knowing is that nlmixr2 is not a walled garden,
which is unusual for this field and is probably the strongest practical
argument for adopting it:

- [**babelmixr2**](https://nlmixr2.github.io/babelmixr2/) takes the
  *same model object* you just read and runs it through NONMEM or
  Monolix. You are not choosing a camp — you can fit in nlmixr2 and
  confirm in NONMEM, which is a genuinely useful thing to be able to do
  when the model matters. It also connects to `PKNCA` for
  non-compartmental initial estimates and `PopED` for optimal design.

- [**nonmem2rx**](https://nlmixr2.github.io/nonmem2rx/) and
  [**monolix2rx**](https://nlmixr2.github.io/monolix2rx/) go the other
  direction, importing finished NONMEM and Monolix runs into R so that
  a decade of legacy models can be simulated, plotted, and checked with
  the same tooling.

- [**rxode2**](https://nlmixr2.github.io/rxode2/) is the ODE solver
  underneath both fitting and simulation. The model used for the
  regimen comparison above is bit-for-bit the model that was fit — not
  a hand-re-implementation of it in a separate simulation tool, which
  is a classic source of quiet errors.

- The surrounding ecosystem — `xpose.nlmixr2` and `ggPMX` for
  standardized diagnostics, `tidyvpc` for VPCs, `shinyMixR` for run
  management, `nlmixr2rpt` for Word and PowerPoint reports — is
  described in more detail in an
  [earlier post](https://blog.nlmixr2.org/blog/2025-06-30-nlmixr2-ecosystem/).

And then the ordinary open-source argument, which lands harder in
pharma than people outside it expect: there is no license to negotiate.
An academic collaborator, a CRO, a reviewer, or a curious statistician
two floors away can run your analysis today.

## Getting started

If you are a pharmacometrician who has been meaning to try this, the
first model is a smaller step than it looks — everything above is about
forty lines of code.

```r
install.packages("nlmixr2")
```

That brings in the estimation engine, the ODE solver, and the plotting
and diagnostic packages. On Windows you will also want
[Rtools](https://cran.r-project.org/bin/windows/Rtools/), since models
are compiled to C for speed. Then:

```r
library(nlmixr2)
fit <- nlmixr2(pheno, pheno_sd, est = "saem",
               control = saemControl(print = 0))
```

Some places to go next:

- The [nlmixr2 website](https://nlmixr2.org) has the full model
  language reference and a gallery of worked examples.
- [nlmixr2lib](https://github.com/nlmixr2/nlmixr2lib) is a library of
  starting models, so a two-compartment model with transit absorption
  is a lookup rather than a derivation.
- If you are coming from NONMEM, `nonmem2rx` will translate a model you
  already trust, which is a considerably more comfortable way in than
  starting from a blank file.
- [nlmixr2 7.0](https://blog.nlmixr2.org/blog/2026-08-05-nlmixr2-7/),
  released this month, made
  `focei` solve in parallel across all OpenMP-enabled platforms, so the
  fits in this post are quicker than they would have been a year ago.

And if you are coming from outside the field entirely — from sports
analytics, from supply chain, from anywhere people fit hierarchical
models to messy observational data -- the thing worth taking away is the
shape of the problem. Fifty-nine subjects, most with one or two
observations, a mechanistic model with parameters that mean something,
and a decision at the end that someone has to make on Tuesday whether
or not the analysis is finished.

---

*The complete source for this post, including every model and figure,
is in the [blog repository](https://github.com/nlmixr2/blog). The
phenobarbital data is from Grasela TH Jr and Donn SM, "Neonatal
population pharmacokinetics of phenobarbital derived from routine
clinical data," Dev Pharmacol Ther. 1985;8(6):374-383.*
