---
title: nlmixr2 augmented plot
author: Matthew Fidler
date: '2025-03-31'
slug: []
categories: [nlmixr2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nlmixr2)
library(rxode2)
# suppress messages
rxode2:::rxSetSilentErr(1L)
```

This month I will on a single `nlmixr2`'s plot function that is shared
with `nlme`, `augPred()`.  I think this is useful but also harder to
find like the `rxode2` plots discussed last month.

The example of this feature is the phenobarbitol data:

```{r}
library(nlmixr2)
pheno <- function() {
  ini({
    tcl <- log(0.008) # typical value of clearance
    tv <-  log(0.6)   # typical value of volume
    ## var(eta.cl)
    eta.cl + eta.v ~ c(1,
                       0.01, 1) ## cov(eta.cl, eta.v), var(eta.v)
                      # interindividual variability on clearance and volume
    add.err <- 0.1    # residual variability
  })
  model({
    cl <- exp(tcl + eta.cl) # individual value of clearance
    v <- exp(tv + eta.v)    # individual value of volume
    ke <- cl / v            # elimination rate constant
    d/dt(A1) = - ke * A1    # model differential equation
    cp = A1 / v             # concentration in plasma
    cp ~ add(add.err)       # define error model
  })
}

# Note the suppress messages simply removes the output from the
# fit so it is easier to read in the blog.
fit <- suppressMessages(nlmixr(pheno, pheno_sd, "saem",
                               control=list(print=0),
                               table=list(cwres=TRUE, npde=TRUE)))

```

You can see the basic plots including the individual plots:

```{r plot}
p <- plot(fit)

# Here I am subsetting the plots to show only individual plots
p <- p[["All Data"]]

# In this case the list of plots is named starting with "individual"
w <- which(vapply(names(p), function(x) grepl("individual", x), logical(1)))

# This creates a new list of plots, and changes it to the same class
# as output by nlmixr2
p <- lapply(w,function(x) p[[x]])
class(p) <- "nlmixr2PlotList"


#This is a hack to suppress the warnings & messages from the plot function
suppressMessages(suppressWarnings(print(p)))
```

In particular the individual plots does not show the complete prediction.

One way to get the complete prediction is to add more points to the
`rxode2` prediction than what is in the original dataset.

In `NONMEM` (and in `nlmixr2`) you **can** add `EVID=2` predictions
into your input dataset, though the ODE solving mesh may change your
nlme solution.

Another method is to add the observations after the nlme fit is
complete by using `augPred()`

This is simple:

```{r augPred0}
ap <- augPred(fit)

head(ap)
```

You can see this is a dataset that you can plot yourself with any
package you would like.  Like `rxode2` there is a `ggplot2` method
attached to `plot` for `augPred()` datasets from `nlmixr2`:

```{r augPredP0}
# The suppressWarnings, supress messages and print is made
# to abbreviate the output, you can also simply use plot(ap)
suppressWarnings(suppressMessages(print(plot(ap))))
```

Here you see the affect of dosing on the outcome more clearly than the
traditional individual predictions.

This also works with multiple endpoint models:

```{r nlmeME}
pk.turnover.emax3 <- function() {
  ini({
    tktr <- log(1)
    tka <- log(1)
    tcl <- log(0.1)
    tv <- log(10)
    ##
    eta.ktr ~ 1
    eta.ka ~ 1
    eta.cl ~ 2
    eta.v ~ 1
    prop.err <- 0.1
    pkadd.err <- 0.1
    ##
    temax <- logit(0.8)
    tec50 <- log(0.5)
    tkout <- log(0.05)
    te0 <- log(100)
    ##
    eta.emax ~ .5
    eta.ec50  ~ .5
    eta.kout ~ .5
    eta.e0 ~ .5
    ##
    pdadd.err <- 10
  })
  model({
    ktr <- exp(tktr + eta.ktr)
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    emax = expit(temax+eta.emax)
    ec50 =  exp(tec50 + eta.ec50)
    kout = exp(tkout + eta.kout)
    e0 = exp(te0 + eta.e0)
    ##
    DCP = center/v
    PD=1-emax*DCP/(ec50+DCP)
    ##
    effect(0) = e0
    kin = e0*kout
    ##
    d/dt(depot) = -ktr * depot
    d/dt(gut) =  ktr * depot -ka * gut
    d/dt(center) =  ka * gut - cl / v * center
    d/dt(effect) = kin*PD -kout*effect
    ##
    cp = center / v
    cp ~ prop(prop.err) + add(pkadd.err)
    effect ~ add(pdadd.err) | pca
  })
}

# Like the prior model, we wish to suppress messages too:
fit.TOS <- suppressMessages(nlmixr(pk.turnover.emax3, warfarin, "saem", control=list(print=0),
                                   table=list(cwres=TRUE, npde=TRUE)))

```

Just like you can see the individuals from the standard plot:

```{r plotME}
p <- plot(fit.TOS)

# Here I am subsetting the plots to show only individual plots
cp <- p[["Endpoint:  cp"]]
pca <- p[["Endpoint:  pca"]]

# In this case the list of plots is named starting with "individual"
w <- which(vapply(names(cp), function(x) grepl("individual", x), logical(1)))

# This creates a new list of plots, and changes it to the same class
# as output by nlmixr2
cp <- lapply(w,function(x) cp[[x]])
class(cp) <- "nlmixr2PlotList"


# In this case the list of plots is named starting with "individual"
w <- which(vapply(names(pca), function(x) grepl("individual", x), logical(1)))

# This creates a new list of plots, and changes it to the same class
# as output by nlmixr2
pca <- lapply(w, function(x) pca[[x]])
class(pca) <- "nlmixr2PlotList"



#This is a hack to suppress the warnings & messages from the plot function
suppressMessages(suppressWarnings(print(cp)))
suppressMessages(suppressWarnings(print(pca)))
```

Now you see the augmented predictions separated by endpoint:

```{r augPredME}
suppressWarnings(suppressMessages(plot(augPred(fit.TOS))))
```

In this case, where there is more rich data, the differences are a bit
less drastic than the phenobarbitol example, but still more full
profiles are present for patients who discontinued when using `augPred()`

Also, as a note, the plots made inside `nlmixr2` are subset by
endpoints and can be subset to which-ever plot that you wish to have.
Perhaps someday easier method (with say `filter()`) can be used to
select the correct plots.

Overall, `augPred()` is an easy and useful way to add more complete predictions to any
model.
