---
title: nlmixr2/rxode2 mu referencing 3.0/4.0
author: Matthew Fidler
date: '2024-12-11'
slug: []
categories: [rxode2, nlmixr2]
tags: [mu]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nlmixr2)
library(rxode2)
```

This month, I will talk about about a new iteration of mu-referencing in nlmixr2, which I call mu3 and mu4.

## What is mu referencing in nlmixr2 -- Review from another post

From the [last blog post](https://blog.nlmixr2.org/blog/2024-07-08-mu2/) about mu-referencing, I will give a brief
overview of mu-referencing and what mu-2 referencing is and how it is
expanded a bit more.

mu-referencing is combining a fixed effect, random effect and possibly a covariate in the form:

$$
\theta_\mathsf{pop}+\eta_\mathsf{individual}+\theta_\mathsf{covariate}\times \mathsf{DataCovariate}
$$

Often they are placed in exponentials for these to be log-normally distributed like:

$$
\exp\left(\theta_\mathsf{pop}+\eta_\mathsf{individual}+\theta_\mathsf{covariate}\times \mathsf{DataCovariate}\right)
$$

In optimization routines like `saem`, these are switched out with a single parameter during optimization classically called $\phi$ in both NONMEM and Monolix.

Once the best $\phi$ is found, then the population, individual and covariates can be found by linear regression of the individual $\phi$ values versus the information in the optimization.

This linear model adds important stability and speed when determining these parameters in the mu-expression. (It also adds rules like they must not be time-varying for instance).

In old versions of rxode2 and nlmixr2 the $\mathsf{DataCovariate}$ had to be in the dataset itself. This classic weight covariate adjustment:

$$
\exp\left(\theta_{Cl}+\eta_{Cl}\right)\times\left(\frac{WT}{70}\right)^{3/4}
$$

would have to written:

$$
\exp\left(\theta_{Cl}+\eta_{Cl} + 3/4\times\log\left(\frac{WT}{70}\right)\right)
$$

If you wanted to estimate the population parameter $3/4$ to see if it approaches the correct value you could with:

$$
\exp\left(\theta_{Cl}+\eta_{Cl} + \theta_{Cl, \textsf{cov}}\times\mathsf{DataCovariate}\right)
$$

Where

$$
\mathsf{DataCovariate} =\log\left(\frac{WT}{70}\right)
$$

This is easy enough to do and adds stabilization.

However, with mu referencing 2.0 you can simply use an additive expression to setup mu-referencing:

$$
\exp\left(\theta_{Cl}+\eta_{Cl} + \frac{WT}{70}+\theta_{Cl, \textsf{cov}}\times\log\left(WT/70\right)\right)
$$

This is a bit more convenient than creating a column in the dataset that does this conversion and less user-based intervention to make `nlmixr2` use linear models when it can.

## Checking for mu2 referencing

In `rxode2` / `nlmixr2` you can check to see if your version of nlmixr2 supports the mu2 referencing by evaluating the functional form:

```{r muRef}
one.compartment <- function() {
  ini({
    tka <- log(1.57); label("Ka")
    tcl <- log(2.72); label("Cl")
    tv <- log(31.5); label("V")
    wt.cl <- 0.75; label("WT on CL")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  # and a model block with the error specification and model specification
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl + wt.cl*log(WT/70))
    v <- exp(tv + eta.v)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

one <- one.compartment()

print(one)
```

If the mu2 referencing is supported it will show the following mu reference table:

```
── μ-referencing ($muRefTable): ──
  theta    eta level                         covariates
1   tka eta.ka    id
2   tcl eta.cl    id log(0.0142857142857143 * WT)*wt.cl
3    tv  eta.v    id
```

Here it shows the transformation that is used when creating the transformed data used for mu-based covariate modeling. In this case, we have `log(0.0142857142857143 * WT)`. It is a bit different than what is written because it is prepossessed by `symengine` and looks at the derivative with respect to the covariate parameter `wt.cl`.

In general, I am excited by this new feature in nlmixr2 because it adds a new level of simplicity to user-based models and more often detects mu-referenced code when it may not have been detected in the past.

## What is new with mu referencing 3.0 and 4.0

The new mu3-referencing allows variables to be specified outside of the
mu-referenced expression (though not estimating values).

Another new feature in nlmixr2 is assigning string variables.
Internally, rxode2 converts this to a numeric and the converts to a
factor in the output.

For nlmixr2, this means string expressions can be converted to mu
referenced expressions too (which is mu4-referencing).


### mu 3.0 example

An example of the mu 3.0 referencing:

```{r mu3}
one.compartment <- function() {
  ini({
    tka <- log(1.57); label("Ka")
    tcl <- log(2.72); label("Cl")
    tv <- log(31.5); label("V")
    wt.cl <- 0.75; label("WT on CL")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  # and a model block with the error specification and model specification
  model({
    wt70 <- log(WT/70)
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl + wt.cl*wt70)
    v <- exp(tv + eta.v)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

one <- one.compartment()

f <- nlmixr2(one, theo_sd, "saem", control=saemControl(print=0))

print(f)
```

Note that the run info shows that `nlmixr2` shows that the mu-3 referencing was detected:

```{r mu3runInfo}
f$runInfo
```

This is only calculated when running `nlmixr2`


### mu 4.0 example

Here is an example:

```{r mu4}
one.compartment <- function() {
  ini({
    tka <- log(1.57); label("Ka")
    tcl <- log(2.72); label("Cl")
    tv <- log(31.5); label("V")
    wt.cl <- 0.75; label("WT on CL")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  # and a model block with the error specification and model specification
  model({
    if (WT < 70) {
      wt70 <- "Less than 70"
    } else {
      wt70 <- "Greater than 70"
    }
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl + wt.cl*(wt70 == "Less than 70"))
    v <- exp(tv + eta.v)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

one <- one.compartment()

f <- nlmixr2(one, theo_sd, "saem", control=saemControl(print=0))

print(f)
```

As above, you can see this mu-referenced expression by the `$runInfo`:

```{r mu4runInfo}
f$runInfo
```

## Conclusion

This new method of mu referencing allows more flexibility in model
specification and more cases where the models will be stable (without
the user having to make as many model changes)


### Why the Pokemon icon?

Originally with the mu2 referencing? Every time I hear mu2 I think of the ultimate genetically modified Pokemon [mewtwo](https://goofy-legends-gl.fandom.com/wiki/Mewtwo) (which is a link to where the image comes from).

Since this was expanded from the original article I used it as the
base, even though now we are at mu3 and mu4 referencing.

Sicne this article came out in December, the image was expanded with [holiday lights](https://freepik.com/free-vector/christmas-background-with-circular-border-glowing-lights_386293523.htm#fromView=search&page=1&position=9&uuid=f840e1ac-cd53-4aa1-8f69-c6d9e5d12d21).
