---
title: rxode2 calculating derived PK model
author: Matthew Fidler
date: '2024-08-05'
slug: []
categories: [rxode2, nlmixr2]
tags: [mu]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rxode2)
```

One of the things that can be useful from time to time is to calculate
different PK parameters based on whatever parameters you have
estimated.

There is a great function,
[`calc_derived()`](https://kestrel99.github.io/pmxTools/reference/calc_derived.html)
in `pmxTools` that allows this calculation of the derived parameters
(by my collaborators Justin Wilkins and Bill Denney).

I think this is an underrated function that can help many people with
typical calculations.

`rxode2` has the same type of function, which can be helpful to test
the `linCmt()` functions,
[`rxDerived()`](https://nlmixr2.github.io/rxode2parse/reference/rxDerived.html).

You may notice that this does not have as many arguments as the
`calc_derived()` this is because this uses the same method to detect
the parameters that `linCmt()` does (described
[here](https://nlmixr2.github.io/rxode2/articles/rxode2-model-types.html#solved-compartment-models)).

There is both advantages and disadvantages to have the algorithm
detect the variables.  One of advantages is calculating does not
require as much manual intervention (which I am for automation as much
as possible), but the disadvantage is incorrect detection of the parameters.

Here are the examples from the manual:

```{r}
## Note that rxode2 parses the names to figure out the best PK parameter
params <- rxDerived(cl = 29.4, v = 23.4, Vp = 114, vp2 = 4614, q = 270, q2 = 73)

## That is why this gives the same results as the value before

params <- rxDerived(CL = 29.4, V1 = 23.4, V2 = 114, V3 = 4614, Q2 = 270, Q3 = 73)

## You may also use micro-constants alpha/beta etc.

params <- rxDerived(k12 = 0.1, k21 = 0.2, k13 = 0.3, k31 = 0.4, kel = 10, v = 10)

## or you can mix vectors and scalars

params <- rxDerived(CL = 29.4, V = 1:3)

## If you want, you can round to a number of significant digits
## with the `digits` argument:

params <- rxDerived(CL = 29.4, V = 1:3, digits = 2)
```

Note that none of these show a data-frame, so an additional example of
this is:

```{r}
df <- data.frame(CL=1:10, V=20:29)

rxDerived(df)
```

Regardless of which you use, these functions are generally useful for doing PK translations


### Notes

The icon comes from [here](https://www.flaticon.com/free-icon/calculation_8123662)
