---
title: nlmixr2/rxode2 user functions to modify code
author: Matthew Fidler
date: '2025-01-24'
slug: []
categories: [rxode2, nlmixr2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nlmixr2)
library(rxode2)
```

This month I will talk about a new type of user function.  Previously,
I spoke of [user
functions](https://blog.nlmixr2.org/blog/2024-05-08-user-functions/)
that you can use in your code to extend the functionality of `rxode2`
and `nlmixr2`.

Recently we released the ability for certain functions to generate code.

I will go over examples that could be helpful to extend `rxode2` ui models:

- An example that allows arguments to be named inside of the `rxode2` ui
  models (though classic models still do not allow this).

- An example that allows linear models to be expanded and possibly
  even automatically initialized.

If interested, this is what was added to [allow neural network
ODE](https://github.com/braemd/pmxNODE/pull/1) to be integrated
directly in `rxode2`/`nlmixr2` using `pmxNODE`.

As a note, the user interface is also flexible in generating specific
code and is also used to allow `posologyr`
([here](https://github.com/levenc/posologyr/pull/47)) to use
`nlmixr2`/`rxode2` models directly for all of its clinical
optimization routines it is also what is used to generate code for
`NONMEM`, `Monolix` and `PopED` in `babelmixr2`.

In addition to this flexibility there is still one undocumented method to
register low level C functions from another package to extend
`rxode2`/`nlmixr2` even further.

The collection of these features and programming interface is unique
to `nlmixr2` and `rxode2` (compared to say `Monolix` or `NONMEM`).  As
in `posologyr` and `pmxNODE`, this allows the language to be expanded
in so many interesting and unique ways.

Also, this is a bit more focused on how to add code expansion to a
package, though it can work with your own code too; for many users
this is simply a way to add Neural Network ODEs (after loading
`pmxNODE`) to `rxode2`/`nlmixr2` or other similar sorts of functions.


## Functions to insert `rxode2` code into the current model

The first example is allowing named arguments in a function that
translates itself into a non-named form:

```r
f <- function() {
  model({
    a <- rxpois(lambda=lam)
  })
}

# Which will evaluate into a standard rxode2 function that does not
# support named arguments (since it is translated to C)
f()

# Which is still true in the standard rxode2:

try(rxode2({
  a <- rxpois(lambda=lam)
}))
```

This is accomplished by a combination of two functions, which are
highly commented:

``` r
rxUdfUi.rxpois <- function(fun) {
  # Fun is the language object (ie quoted R object) to be evaluated or
  # changed in the code
  .fun <- fun
  # Since the `rxpois` function is built into the rxode2 we need to
  # have a function with a different conflicts.  In this case, I take
  # the function name (fun[[1]]), and prepend a ".", which follows
  # `rxode2`'s naming convention of un-exported functions.
  #
  # This next evaluation changes the expression function to .rxpois()
  .fun[[1]] <- str2lang(paste0(".", deparse1(fun[[1]])))
  # Since this is still a R expression, you can then evaluate the
  # function .rxpois to produce the proper code:
  eval(.fun)
}

# The above s3 method can be registered in a package or you can use
# the following code to register it in your session:
rxode2::.s3register("rxode2::rxUdfUi", "rxpois")

# This is the function that changes the code as needed
.rxpois <- function(lambda) {
  # The first part of this code tries to change the value into a
  # character.  This handles cases like rxpois(lambda=lam),
  # rxpois(lam), rxpois("lam").  It also tries to evaluate the
  # argument supplied to lambda in case it comes from a different
  # location.
  .lam <- as.character(substitute(lambda))
  .tmp <- try(force(lambda), silent=TRUE)
  if (!inherits(.tmp, "try-error")) {
    if (is.character(.tmp)) {
      .lam <- lambda
    }
  }
  # This part creates a list with the replacement text, in this case
  # it woulb be rxpois(lam) where there is no equals included, as
  # required by `rxode2`:
  list(replace=paste0("rxpois(", .lam, ")"))
}

```

In general the list that the function needs to return can have:

- `$replace` -- The text that will be replaced

- `$before` -- lines that will be placed in the model before the
  current function is found

- `$after` -- lines that are added in the model after the current
  function is found

- `$iniDf` -- the initial estimates `data.frame` for this problem.
  While calling this function, you can retrieve the initial conditions
  currently used parsing you can get the prior value with
  `rxUdfUiIniDf()` and then you can modify it inside the function and
  return the new `data.frame` in this list element.  This allows you
  add/delete initial estimates from the model as well as modify the
  model lines themselves.

- `$uiUseData` -- when `TRUE`, this instructs `rxode2` and
  `nlmixr2est` to re-parse this function in the presence of data, this
  means a bit more function setup will need to be done.

- `$uiUseMv` -- when `TRUE` this instructs `rxode2` to re-parse the
  function after the initial model variables are calculated.

In addition to the `rxUdfUiIniDf()` you can get information about the
parser:

- `rxUdfUiParsing()` returns if the rxode2 ui function is being parsed
  currently (this allows a function to be overloaded as a `udf` for
  calling from `rxode2` as well as a function for modifying the
  model).

- `rxUdfUiNum()` during parsing the function you are calling (in the
  example above `rxpois()` can be called multiple times.  This gives
  the number of the function in the model in order (the first would
  give `1` the second, `2`, etc).  This can be used to create unique
  variables with functions like `rxIntToLetter()` or `rxIntToBase()`.

- `rxUdfUiIniLhs()` which gives the left-handed side of the equation
  where the function is found.  This is also a `R` language object.

- `rxUdfUiIniMv()` gives the model variables for parsing (can be used
  in functions like `linCmt()`)

- `rxUdfUiData()` which specifies the data that are being used to
  simulate, estimate, etc.

- `rxUdfUiEst()` which gives the estimation/simulation method that is
  being used with the model.  For example, with simulation it would be
  `rxSolve`.

### Using model variables in `rxode2` ui models

You can also take and change the model and take into consideration the
`rxode2` model variables before the full `ui` has completed its
parsing.  These `rxode2` model variables has information that might
change what variables you make or names of variables. For example it
has what is on the left hand side of the equations (`$lhs`), what are
the input parameters (`$params`) and what is the ODE states
(`$state`)).

If you are using this approach, you will likely need to do the
following steps:

- When data are not being processed, you need to put the function in
  an `rxode2` acceptable form, no named arguments, no strings, and
  only numbers or variables in the output.

- The number of arguments of this output needs to be declared in the
  `S3` method by adding the attribute `"nargs"` to method.  For
  example, the built in `testMod1()` ui modification function uses
  only one argument when parsed

Below is a commented example of the model variables example:

```{r rxModelVarsUiModel}

testMod1 <- function(val=1) {
  # This converts the val to a character if it is somthing like testMod1(b)
  .val <- as.character(substitute(val))
  .tmp <- suppressWarnings(try(force(val), silent = TRUE))
  if (!inherits(.tmp, "try-error")) {
    if (is.character(.tmp)) {
      .val <- val
    }
  }
  # This does the UI parsing
  if (rxUdfUiParsing()) {
    # See if the model variables are available
    .mv <- rxUdfUiMv()
    if (is.null(.mv)) {
      # Put this in a rxode2 low level acceptible form, no complex
      # expressions, no named arguments, something that is suitable
      # for C.
      #
      # The `uiUsMv` tells the parser this needs to be reparsed when
      # the model variables become avaialble during parsing.
      return(list(replace=paste0("testMod1(", .val, ")"),
                  uiUseMv=TRUE))
    } else {
      # Now that we have the model variables, we can then do something
      # about this
      .vars <- .mv$params
      if (length(.vars) > 0) {
        # If there is parameters available, this dummy function times
        # the first input function by the value specified
        return(list(replace=paste0(.vars[1], "*", .val)))
      } else {
        # If the value isn't availble, simply replace the function
        # with the value.
        return(list(replace=.val))
      }
    }
  }
  stop("This function is only for use in rxode2 ui models",
       call.=FALSE)

}

rxUdfUi.testMod1 <- function(fun) {
  eval(fun)
}

# To allow this to go to the next step, you need to declare how many
# arguments this argument has, in this case 1.  Bu adding the
# attribute "nargs", rxode2 lower level parser knows how to handle
# this new function.  This allows rxode2 to generate the model
# variables and send it to the next step.

attr(rxUdfUi.testMod1, "nargs") <- 1L

# If you are in a package, you can use the rxoygen tag @export to
# register this as a rxode2 model definition.
#
# If you are using this in your own script, you need to register the s3 function
# One way to do this is:
rxode2::.s3register("rxode2::rxUdfUi", "testMod1")

## These are some examples of this function in use:

f <- function() {
  model({
    a <- b + testMod1(3)
  })
}

f <- f()

print(f)

f <- function() {
  model({
    a <- testMod1(c)
  })
}

f <- f()

print(f)

f <- function() {
  model({
    a <- testMod1(1)
  })
}

f <- f()

print(f)

```

### Using data for `rxode2` ui modification models

The same steps are needed to use the data in the model replacement;
You can then use the data and the model to replace the values inside
the model.  A worked example `linMod()` is included
that has the ability to use:

- model variables,
- put lines before or after the model,
- add initial conditions
- And use data in the initial estimates

```{r linMod}

# You can print the code:
linMod

# You can also print the s3 method that is used for this method

rxode2:::rxUdfUi.linMod
```
