---
title: nlmixr2/rxode2 mu referencing 2.0
author: Matthew Fidler
date: '2024-07-08'
slug: []
categories: [rxode2, nlmixr2]
tags: [mu]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nlmixr2)
library(rxode2)
```

This month, I will talk about about a new iteration of mu-referencing in nlmixr2, which I call mu2.

## What is mu referencing in nlmixr2

mu-referencing is combining a fixed effect, random effect and possibly a covariate in the form:

$$
\theta_\mathsf{pop}+\eta_\mathsf{individual}+\theta_\mathsf{covariate}\times \mathsf{DataCovariate}
$$

Often they are placed in exponentials for these to be log-normally distributed like:

$$
\exp\left(\theta_\mathsf{pop}+\eta_\mathsf{individual}+\theta_\mathsf{covariate}\times \mathsf{DataCovariate}\right)
$$

In optimization routines like `saem`, these are switched out with a single parameter during optimization classically called $\phi$ in both NONMEM and Monolix.

Once the best $\phi$ is found, then the population, individual and covariates can be found by linear regression of the individual $\phi$ values versus the information in the optimization.

This linear model is important.

In old versions of rxode2 and nlmixr2 the $\mathsf{DataCovariate}$ had to be in the dataset itself. This classic weight covariate adjustment:

$$
\exp\left(\theta_{Cl}+\eta_{Cl}\right)\times\left(\frac{WT}{70}\right)^{3/4}
$$

would have to written:

$$
\exp\left(\theta_{Cl}+\eta_{Cl} + \frac{WT}{70}+3/4\times\log\left(WT/70\right)\right)
$$

If you wanted to estimate the population parameter $3/4$ to see if it approaches the correct value you could with:

$$
\exp\left(\theta_{Cl}+\eta_{Cl} + \frac{WT}{70}+\theta_{Cl, \textsf{cov}}\times\mathsf{DataCovariate}\right)
$$

Where $$\mathsf{DataCovariate}$$ is defined in the data as $\log\left(WT/70\right)$.

This is easy enough to do and adds stabilization.

However, with mu referencing 2.0 you can simply use an additive expression to setup mu-referencing:

$$
\exp\left(\theta_{Cl}+\eta_{Cl} + \frac{WT}{70}+\theta_{Cl, \textsf{cov}}\times\log\left(WT/70\right)\right)
$$

This is a bit more convenient than creating a column in the dataset that does this conversion and less user-based intervention to make `nlmixr2` use linear models when it can.

## Checking for mu2 referencing

In `rxode2` / `nlmixr2` you can check to see if your version of nlmixr2 supports the mu2 referencing by evaluating the functional form:

```{r muRef}
one.compartment <- function() {
  ini({
    tka <- log(1.57); label("Ka")
    tcl <- log(2.72); label("Cl")
    tv <- log(31.5); label("V")
    wt.cl <- 0.75; label("WT on CL")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  # and a model block with the error specification and model specification
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl + wt.cl*log(WT/70))
    v <- exp(tv + eta.v)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

one <- one.compartment()

print(one)
```

If the mu2 referencing is supported it will show the following mu reference table:

```         
── μ-referencing ($muRefTable): ──  
  theta    eta level                         covariates
1   tka eta.ka    id                                   
2   tcl eta.cl    id log(0.0142857142857143 * WT)*wt.cl
3    tv  eta.v    id                                   
```

Here it shows the transformation that is used when creating the transformed data used for mu-based covariate modeling. In this case, we have `log(0.0142857142857143 * WT)`. It is a bit different than what is written because it is prepossessed by `symengine` and looks at the derivative with respect to the covariate parameter `wt.cl`.

In general, I am excited by this new feature in nlmixr2.

### Why the Pokemon icon?

Why the icon for mu2 referencing? Every time I hear mu2 I think of the ultimate genetically modified Pokemon mewtwo.
