---
title: "nlmixr2save: fits you can read, runs you don't repeat"
subtitle: "Portable text-based fits and the := caching operator, on a poster at ACoP 2026"
author: Matthew Fidler
date: '2026-10-11'
slug: []
categories: [nlmixr2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      collapse = TRUE, comment = "#>")
library(nlmixr2)
library(nlmixr2save)
# Keep the saved fits and cache files out of the blog source tree
cacheDir <- file.path(tempdir(), "nlmixr2save-post")
dir.create(cacheDir, showWarnings = FALSE)
knitr::opts_knit$set(root.dir = cacheDir)
```

<div style="position:relative;padding-bottom:56.25%;height:0;overflow:hidden;max-width:100%;">
<iframe src="https://www.youtube.com/embed/TD7yhaaQToE"
        title="nlmixr2save: fits you can read, runs you don't repeat"
        style="position:absolute;top:0;left:0;width:100%;height:100%;border:0;"
        allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"
        allowfullscreen></iframe>
</div>

[Watch on YouTube](https://youtu.be/TD7yhaaQToE)

If you are at [ACoP 2026](https://acop2026.eventscribe.net/) (October
11-14, National Harbor, Maryland), come find the `nlmixr2save` poster.
This post covers the same material, with code you can run.

`nlmixr2save` is on CRAN and version 0.2.0 is out now:

```r
install.packages("nlmixr2save")
```

## Two problems

**Problem 1: Fits you can't open later.** Nonlinear mixed-effects
models can take minutes to weeks to fit. Once one is done, you want
to keep it. The usual way is `saveRDS()`, but that ties the result to R's
binary format and to how `nlmixr2` stored the fit internally.

That second tie actually broke. `nlmixr2` used the `qs` package to
compress fits. When `qs` was removed from CRAN, `nlmixr2est` moved to
`qs2`, which broke backward compatibility for saved fits. It later
moved to base R serialization, and old `qs2` fits can still be read
only if `qs2` is installed. A fit that depends on how a particular
package version stored its internals is not really an archive.

**Problem 2: Running the same model again for no reason.** During
development you source a script, or render a Quarto/R Markdown report,
over and over. Each time, every `nlmixr2()` call runs again, even when
nothing about the model, data, or settings changed.

`nlmixr2save` fixes both.

## A model to work with

```{r model}
one.cmt <- function() {
  ini({
    tka <- 0.45; label("Log Ka")
    tcl <- log(c(0, 2.7, 100)); label("Log Cl")
    tv <- 3.45; label("log V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}
```

## Saving a fit as text

`saveFit()` breaks a fit into its parts and writes each one as text.
Models, including the final estimated model, become `nlmixr2`/`rxode2`
R source. Tables (estimates, objective function values, the fit
table, the data) become CSV. Everything is bundled into one `.zip`:

```{r saveFit}
fit <- nlmixr2(one.cmt, theo_sd, est = "focei",
               control = list(print = 0))

saveFit(fit)  # writes fit.zip

zip::zip_list("fit.zip")[, c("filename", "uncompressed_size")]
```

Unzip it and you can read nearly all of the fit in a text editor, a
spreadsheet, or another language (a couple of small components, such as
the session info, are still `.rds`). No R is needed to see what was
estimated. In R, `loadFit()` rebuilds the fit:

```{r loadFit}
fit2 <- loadFit("fit")
fit2$parFixedDf
```

A loaded fit is a working fit, not a snapshot. Diagnostic plots, VPCs,
and model piping all work, so you can use it as the starting point
for the next model:

```{r pipe, eval = FALSE}
plot(fit2)
vpcPlot(fit2)
fit3 <- fit2 |> model(ka <- exp(tka)) |> nlmixr2(est = "focei")
```

The saved fit also records the `nlmixr2est` and `rxode2` versions that
produced it. If you load it later with different versions installed,
you get a warning explaining why a rerun might not match exactly.

## Caching with `:=`

The second tool is the one you will use daily. Change `<-` to `:=`:

```r
fit <- nlmixr2(one.cmt, theo_sd, est = "saem")   # always runs
fit := nlmixr2(one.cmt, theo_sd, est = "saem")   # runs once, then loads
```

That is the only change. Here is how `:=` works:

1. It computes a hash of the model, the data, and every estimation
   option. For random calls, such as simulations, it also records the
   random seed at the start.
2. If a cache file with that hash (and, for random calls, the same
   starting seed) exists, `:=` loads it. For random calls, it also
   restores the ending seed state, so the rest of your script sees the
   same random numbers it would have seen if the call had run.
3. Otherwise the call runs and the result is saved.

The first call fits the model:

```{r cache1}
system.time(
  fitS := nlmixr2(one.cmt, theo_sd, est = "saem",
                  control = list(print = 0))
)
```

The same call again loads from disk:

```{r cache2}
system.time(
  fitS := nlmixr2(one.cmt, theo_sd, est = "saem",
                  control = list(print = 0))
)
```

Change anything that matters and it runs again. Here, the number of
burn-in iterations changes:

```{r cache3}
system.time(
  fitS := nlmixr2(one.cmt, theo_sd, est = "saem",
                  control = list(print = 0, nBurn = 250))
)
```

Hashing the data is smarter than hashing the whole data frame. The
hash only uses the columns that affect estimation: standard columns,
model covariates, and `table$keep` columns. If you add a notes column
to the data, `fitS` still loads from the cache. If you change `DV`, it
refits.

### Beyond `nlmixr2()`

`:=` works with any function. For fits, the cache is the text `.zip`
above. For everything else, it is an `.rds` file that stores the
result and the hash:

```r
set.seed(42)
sim := rxSolve(fit, nStud = 10)
```

The first time the script runs, this simulates and saves `sim.rds`
with the starting seed. The next time the script is sourced, it starts
from the same `set.seed(42)`, so `:=` loads `sim.rds` and moves the
seed to where the simulation left it. If the starting seed is
different, for example a second call later in the same session, `:=`
simulates again.

To make your own stochastic function seed-aware, register it with
`saveFitRandom("mySimulate")`.

Two notes:

- `:=` has to see the long-running call directly.
  `suppressMessages(nlmixr2(...))` always runs, because the inner call
  runs before `:=` can check the cache.
- The cache file is named after the variable on the left, so `fitS`
  becomes `fitS.zip`.

## Options for reports and projects

Version 0.2.0 adds options for cached work that lives in a project or
a rendered document:

| option | default | what it does |
|:---|:---|:---|
| `nlmixr2save.dir` | `"."` | directory for cache files |
| `nlmixr2save.prefix` | `""` | prefix for cache file names, e.g. one per report |
| `nlmixr2save.check` | `TRUE` | set to `FALSE` to trust any existing cache and only run when the file is missing |
| `nlmixr2save.data` | `TRUE` | set to `FALSE` to leave the original data out of saved fits |
| `nlmixr2save.checkVersion` | `TRUE` | warn (or ask, when interactive) if a cache came from different package versions |

`nlmixr2save.check = FALSE` is useful when the cache is committed to
version control. The document renders the same way on every machine,
even with different `nlmixr2` versions installed. When you do want to
rerun everything, `nlmixr2saveInvalidate()` deletes every file in
`nlmixr2save.dir` whose name starts with `nlmixr2save.prefix`. Give the
cache its own directory or prefix first; with the defaults, that means
every file in the working directory.

## Sharing a fit without the data

Often you can share a model but not the subject-level data behind it.
`nlmixr2saveShare()` takes an existing saved fit and writes a smaller
copy next to it. The original is not changed:

```r
nlmixr2saveShare("fit")                # fit-noData.zip
nlmixr2saveShare("fit", noFit = TRUE)  # fit-noData-noFit.zip
```

The first copy removes the data and keeps the predictions and
residuals. The second copy also removes the per-observation table,
leaving the model, estimates, objective function, eta table, and
parameter history. Anything that needs the data won't work on the
copy, including VPCs and refitting. The [sharing
vignette](https://nlmixr2.github.io/nlmixr2save/articles/sharing-fits.html)
lists exactly what is lost.

## Summary

- `saveFit()` / `loadFit()`: fits stored as R source and CSV that stay
  readable and don't depend on a binary format.
- `fit := nlmixr2(...)`: rerun only when something changed, with
  reproducible random seeds.

If you source scripts over and over or render reports with fits in
them, try switching `<-` to `:=`.

More information:

- [nlmixr2save documentation](https://nlmixr2.github.io/nlmixr2save/)
- [Using nlmixr2save](https://nlmixr2.github.io/nlmixr2save/articles/uses.html)
- [Version tracking](https://nlmixr2.github.io/nlmixr2save/articles/version-tracking.html)
- [GitHub](https://github.com/nlmixr2/nlmixr2save) for issues and ideas

## References

1. Fidler M, Wilkins JJ, Hooijmaijers R, et al. Nonlinear mixed-effects
   model development and simulation using nlmixr and related R
   open-source packages. *CPT Pharmacometrics Syst Pharmacol*.
   2019;8(9):621-633.
2. Wang W, Hallow KM, James DA. A tutorial on RxODE: simulating
   differential equation pharmacometric models in R. *CPT
   Pharmacometrics Syst Pharmacol*. 2016;5(1):3-10.
