---
title: nlmixr2 7.0's covariance step, all grown up
author: Matthew Fidler
date: '2026-08-17'
slug: []
categories: [nlmixr2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

<div style="position:relative;padding-bottom:56.25%;height:0;overflow:hidden;max-width:100%;">
<iframe src="https://youtu.be/vN9NKgBxKFs"
        title="nlmixr2 7.0: the covariance step, all grown up"
        style="position:absolute;top:0;left:0;width:100%;height:100%;border:0;"
        allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"
        allowfullscreen></iframe>
</div>

[Watch on YouTube](https://youtu.be/vN9NKgBxKFs)


In the [7.0 post](/blog/2026-08-05-nlmixr2-7/), I talked mostly about
parallel solving in `focei`. This time I want to talk about a feature
that has been request frequently: a full covariance step, and other
refinements to the covariance steps.

Two things changed there, and they are related:

- Nearly any covariance method can now be requested from nearly any
  estimation method, and you can switch a finished fit to a different
  one without refitting.

- The default covariance step now actually covers every estimated
  parameter, not just the structural ones.

Neither of these will show up in a benchmark. Both of them will show
up the next time you stare at a `$parFixed` table wondering why a
residual-error parameter has a standard error, or wondering whether
your `saem` fit's confidence intervals would look different under a
different covariance method.

The easiest way to look at this is the most modeled drug --
theophylline:

```{r oneCmt}
library(nlmixr2)

pheno <- function() {
  ini({
    tcl <- log(0.008) # typical value of clearance
    tv <-  log(0.6)   # typical value of volume
    ## var(eta.cl)
    eta.cl + eta.v ~ c(1,
                       0.01, 1) ## cov(eta.cl, eta.v), var(eta.v)
    # interindividual variability on clearance and volume
    add.sd <- 0.1    # residual variability
  })
  model({
    cl <- exp(tcl + eta.cl) # individual value of clearance
    v <- exp(tv + eta.v)    # individual value of volume
    ke <- cl / v            # elimination rate constant
    d/dt(A1) = - ke * A1    # model differential equation
    cp = A1 / v             # concentration in plasma
    cp ~ add(add.sd)       # define error model
  })
}

```

## The covariance step now includes every parameter

Look at `add.sd` in that model. It is the additive residual error
standard deviation -- a parameter you fit just like `tka`, `tcl` or
`tv`. Before 7.0, if you ran

```{r theoFocei}
fit := nlmixr2(pheno, pheno_sd, est="focei")
```

`add.sd` would come back in `fit$parFixed` with an estimate and
nothing else -- no SE, no %RSE, no confidence interval.

Now all parameter describing how much noise is in your assay did not.

```{r theoFoceiFit}
print(fit)
```

The old behavior was an intentional choice for a few reasons:

1. We as a `nlmixr2` team think when simulating with uncertainty,
   `omega` and `sigma` terms should be simulated using the inverse
   Wishart distribution instead of standard errors.

2. To get the `omega` and residuals standard errors required quite a
   bit more effort than the approach we originally took.

In 7.0 a full covariance step gives a real standard error.

Fit the model above and you'll see something like `add.sd` at an SE of
about `0.008` (roughly 0.9% RSE) -- a number that, before this release,
simply was not there.

There are only a few things that are still not calculated:

- `fix()`ed  parameters
- inter-occasion variability (IOV) parameter, or
- it is a mixture-probability parameter on the mlogit scale (yes, we
  have mixtures now too, more about this later).

If you look at the full covariance (`$cov`) you can also see the omega
values:

```{r theoFoceiCov}
fit$cov
```
As a note about naming for the omega matrix

Every other estimated theta -- structural, covariate, and now every
residual-error parameter -- is included, and `covFull=TRUE` (the
default) reports the full theta + residual + `Omega` covariance, with
the `Omega` rows named by the random effect (`om.eta.cl`,
`cov.eta.cl.eta.v`) so you can find what you're looking for without
guessing at row order.

## Any covariance, on any method

The second change is that `covMethod` stopped being tied to a specific
estimation method. Historically, the SAEM Louis stochastic-approximation
FIM only came out of `saem`, and the importance-sampling Monte Carlo
observed information only came out of `imp`/`impmap`. If you wanted
that covariance on a `focei` fit, you were out of luck.

Now `sa` and `imp` are covariance methods first and estimation methods
second. Either one can be requested as the `covMethod` of any
mixed-effects fit, computed post-fit at the converged estimates:

```{r switchCov}
# the SAEM Louis stochastic-approximation FIM, on a focei fit,
# without refitting anything
setCov(fit, "sa")

# the importance-sampling Monte Carlo observed information, same fit
setCov(fit, "imp")
```

The `setCov()` will calculate the covariance and integrate it into the
fit. Once calculated, it will be saved and you can switch between
covariances by `setCov()` without the overhead of calculating it again.

Here's the full list of `covMethod` tokens you can reach for, and what
each one actually is:

- `"r,s"` -- the sandwich estimator `solve(R) %*% S %*% solve(R)`,
  where `R` is the Hessian of the objective and `S` is the sum of each
  individual's gradient cross-product at their empirical Bayes
  estimate. This is the finite-difference default for the `focei`
  family.
- `"r"` / `"s"` -- just the Hessian-based or cross-product-based half
  of the sandwich, if you want to compare them.
- `"analytic"` -- the exact analytic observed-information covariance,
  matching NONMEM's `$COV MATRIX=R` but it is exact rather than
  finite-differenced, and it covers additive/proportional/combined
  error, censored M2/M3/M4 observations, an estimated `boxCox()` or
  `yeoJohnson()` lambda, `foce+`, `matExp()`/`indLin()` models, and
  mu-referenced covariate parameters. When a model falls outside that
  scope it says so and falls back to the finite-difference sandwich,
  rather than failing outright.
- `"sa"` -- the SAEM Louis stochastic-approximation Fisher information
  (Kuhn & Lavielle 2005): a dedicated covariance phase resimulates the
  individual parameters at the converged estimate and Monte-Carlo
  averages the observed-information integrand into a FIM decoupled
  from the cooling schedule. No longer `saem`-only.

- `"imp"` -- importance-sampling Monte Carlo observed information, in
  the style of NONMEM `METHOD=IMP`. No longer `imp`/`impmap`-only.

- `"vi"` -- the variational covariance from `emvi`/`fbvi` (can't be
  calculated outside `emvi`/`fbvi`).

- `"nlme"` -- `nlme`'s own standard errors, for when you want them
  as-is rather than recomputed (can't be calculate outside by `nlme`).

- `"linFim"` / `"fim"` -- SAEM's linearized Fisher information, and
  the (noisier) Fisher information matrix accumulated during
  estimation itself, both usable as fallbacks or direct
  requests. (Can't be calculated outside of `saem`).

- `""` -- skip the covariance step entirely.

Because every family now recognizes this shared vocabulary, the
default `covMethod` per family also changed where it made sense to:

- the FOCEI family (`focei`/`foce`/`laplace`/`agq`) defaults to
  `"r,s"`,

- `est="saem"` keeps `"sa"` as its default, now with `"analytic"` and
  `"linFim"` as fallbacks when a model is out of `sa`'s scope,

- `est="imp"`/`"impmap"`/`"qrpem"` default to `"imp"` (the old
  `impCov=TRUE` argument is gone -- `covMethod="imp"` is just the
  default now),

- `est="nlme"` keeps `nlme`'s own covariance (`"nlme"`) as the
  default, but can now also recompute at the converged estimate with
  `covMethod="analytic"` or the finite-difference methods,

- `est="npag"`/`"npb"`, which previously reported no covariance at
  all, now compute one by default (`"imp"`),

- `est="emvi"`/`"fbvi"` keep their variational covariance (`"vi"`) as
  the default, and now honor an explicit `covMethod` override instead
  of silently overwriting it.

## Why this is worth having

None of this changes what a fit estimates. What it changes is how much
you can trust -- and cross-check -- what comes out of the covariance
step.

The nonparametric family going from "no covariance at all" to a real
one is the most obvious win, but the quieter one is being able to spot
check an `"r,s"` finite-difference covariance against the exact
`"analytic"` one, or against `"sa"`, without paying for a second fit.
If two covariance methods agree on a parameter's SE, that's real
signal. If they don't, that's worth knowing before you write the
number into a report.

## References

Kuhn E, Lavielle M. Maximum likelihood estimation in nonlinear mixed
effects models. *Comput Stat Data Anal*. 2005;49(4):1020-1038.
