---
title: nlmixr2 and tidyvpc
author: Matthew Fidler
date: '2026-02-27'
slug: []
categories: [nlmixr2]
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### VPCs -- a new method for VPCs in nlmixr2

I am excited to announce a new VPC method available in `nlmixr2` --
`tidyvpc`.

This method can be built on-top of previous code and provides slightly
different visualizations than the `vpc` package.

Below is an example that shows the two different methods:

```{r pheno}
library(nlmixr2)
library(patchwork)


pheno <- function() {
  ini({
    tcl <- log(0.008) # typical value of clearance
    tv <-  log(0.6)   # typical value of volume
    ## var(eta.cl)
    eta.cl + eta.v ~ c(1,
                       0.01, 1) ## cov(eta.cl, eta.v), var(eta.v)
                      # interindividual variability on clearance and volume
    add.err <- 0.1    # residual variability
  })
  model({
    cl <- exp(tcl + eta.cl) # individual value of clearance
    v <- exp(tv + eta.v)    # individual value of volume
    ke <- cl / v            # elimination rate constant
    d/dt(A1) = - ke * A1    # model differential equation
    cp = A1 / v             # concentration in plasma
    cp ~ add(add.err)       # define error model
  })
}

fit <- nlmixr(pheno, pheno_sd, "saem",
              control=list(print=0),
              table=list(cwres=TRUE, npde=TRUE))

# Using the VPC package, a regular VPC
p1 <- vpcPlot(fit, show=list(obs_dv=TRUE))

## A prediction-corrected VPC
p2 <- vpcPlot(fit, pred_corr = TRUE, show=list(obs_dv=TRUE))


# using the tidyvpc package with method="tidyvpc"
# Using the VPC package, a regular VPC
p1 <- vpcPlot(fit, show=list(obs_dv=TRUE))

## A prediction-corrected VPC
p2 <- vpcPlot(fit, pred_corr = TRUE, show=list(obs_dv=TRUE))

p1 / p2


p1 <- vpcPlot(fit, method="tidyvpc")

## A prediction-corrected VPC
p2 <- vpcPlot(fit, pred_corr = TRUE, method="tidyvpc")

p1 / p2

```

This different ways to handle and visualize the VPCs.

I have been long wanting to add `tidyvpc` as an option to the
`vpcPlot()` function in `nlmixr2` since `tidyvpc` is also an excellent
VPC package.


## What inspired adding `tidyvpc` package now?

This was driven by a CRAN request to fix `nlmixr2plot` errors
that started when the excellent `vpc` package was updated.

I have submitted a bug-fix to the `vpc` package which hopefully will
be integrated in the next CRAN release.

In the meantime, there are a few VPCs that `nlmixr2` could plot but
cannot plot any longer using the `vpc` package.

With this new release, you now have a few more options to produce a VPC.

First, you can try the new method `vpcPlot(..., method="tidyvpc")`.

Second, you can use:

``` r
remotes::install_github("mattfidler/vpc")
```

To use the bug fixed VPC package.

Or, you can roll your own VPC code if you wish.
